% Thesis Acknowledgements ------------------------------------------------


%\begin{acknowledgementslong} %uncommenting this line, gives a different acknowledgements heading
%\begin{acknowledgements}      %this creates the heading for the acknowlegments

\begin{quoting}
  \singlespace
    \begin{center}
  {\LARGE \bfseries  Acknowledgements}\\
  \vspace*{0.5cm}
  \end{center}
\noindent

My supervisor, Professor Tom Allen, has been a great support and inspiration ever since we first corresponded about the possibility of me doing a PhD in Durham, in the spring of 2012. His style as a supervisor has been superb: calm and unperturbed, yet always sharp and focused, readily available to offer insightful comments and valuable guidance. Thank you, Tom.

Second, I would like to thank Durham Law School for offering me a place at their department and for treating me well while I have been there. Thanks also to Professor Leigh and Professor Masterman for taking an interest and giving me valuable comments following my first year review.

Third, I would like to thank Professor Jacques Sluysmans, Professor Hanri Mostert, and Professor Leon Verstappen, for welcoming me to their regular colloquia on expropriation law. Attending and speaking at these meetings has been a very valuable experience for me, allowing me to meet other researches, from many different jurisdictions, who are also working on expropriation law. A special thanks to Professor Sluysmans, Dr Emma Waring and Dr Stijn Verbijst for inviting me to contribute a chapter on Norway in their book on expropriation in Europe. Another special thanks to Dr Waring for sending me a copy of her doctoral thesis on private takings; it has been a great help and inspiration for my own work. Also a special thanks to Dr Ernst Marais and Bj\"{o}rn Hoops for organising an excellent conference in Rome and being very helpful and welcoming to new members of the expropriation research community. Hopefully, this community will stay together and continue to prosper.

Fourth, I would like to thank Ustinov College for welcoming me as a student in Durham and providing a relaxed and friendly atmosphere during my first year as a PhD student. Thanks also to the friends I met there, including Julia, Alan, Noel, Meghan, Lloyd, Peter, and Alma. A special thanks to Isabel Richardson, for being both a highly valued friend and an excellent proofreader.

Fifth, I would like to thank family, friends and colleagues in Norway, especially Ragnhild, Truls and Piotr (who is just on holiday in Germany, I am sure). A special thanks to my father and my brother, for motivation, guidance, and support. A special thanks also to my sisters and our mothers, for inspiration. Thanks to Einar Sofienlund for sharing his insight and providing invaluable information about small-scale hydropower. Thanks also to Johan Fr Remmen, for making clear why this thesis should be written. Hopefully, I have made a good start towards doing justice to this subject.

Lastly, I would like to thank Marijn Visscher. No doubt, coming to Durham was the best decision I ever made.

\end{quoting}


%\end{acknowledgements}
%\end{acknowledgmentslong}

% ------------------------------------------------------------------------

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "../thesis"
%%% End: 


% A note on the format. I could not get the acknowledgements macro to work properly, despite some effort, and so I redesigned it rather messily, as you will see above. If you enter text, it will work -- but the more technologically competent among you will probably be able to fix it. If and when that is done, if you could send the resultant document back to the Law Faculty to update it, that would be great. 