\chapter{Taking Waterfalls}\label{chap:4}

\section{Introduction}\label{sec:intro4}

In this Chapter, I address expropriation of waterfalls in more depth, particularly the administrative practices that have evolved in relation to such expropriation. My main aim is to shed light on how these practices impact on the position of owners and local communities, and to assess them against general principles of expropriation law. 

Most legal scholars in Norway emphasise that the law of expropriation contains many procedural safeguards and that these are meant to ensure a careful assessment of owners' interests against the needs of the public. By contrast, the water authorities tend to consider expropriation of waterfalls as a minor issue, a natural consequence of a development license. In relation to the licensing question, moreover, the established practice is to focus on the environmental consequences of development, not how interference in property affects owners and local communities.

As I show in this Chapter, this has had a significant {\it practical} effect, also in relation to how the authorities assess the material question of whether or not an expropriation order should be granted. In particular, a {\it presumption} has developed, whereby the administrative decision-makers considers a license to undertake large-scale development as an indication that an expropriation order should also be granted.\footnote{The leader of hydropower licensing division of the NVE made an explicit statement to this effect in \cite{flatby08}.} Importantly, this presumption has remained in place, even though the context and consequences of waterfall expropriation have changed since the liberalisation of the hydropower sector in the early 1990s. In the present Chapter, I give a detailed presentation of the history of current practices, before I illustrate their effects by considering the recent case of {\it Jørpeland}.\footnote{See \cite{jorpeland11}.}

The structure of this Chapter is as follows. I begin in Section \ref{sec:explaw} by giving a brief overview of expropriation law generally, as well as special statutory rules relating to hydropower. In Section \ref{sec:twp}, I present the historical context. In short, I argue that because hydropower generation was seen and organised as a public service, expropriation for this purpose enjoyed a high degree of political legitimacy. In addition, the lack of an open market meant that owners could not benefit commercially from developing hydropower themselves. Hence, their financial loss following expropriation was limited. In fact, expropriation (or voluntary sale) to a large-scale developer was in most cases the best thing an owner could hope for, in so far as he wished to benefit financially from his waterfall rights.

%The story begins in the late 19th century, when the first statutory authorities for such expropriation began to emerge. Initially, these authorities were very narrow, however, and they did not cover expropriation of waterfalls, only additional land and rights that waterfall owners might need to develop their resource. 
%
%Later, in the early 20th century, the public sector began to expropriate waterfalls for hydropower, but this was only authorised on narrowly defined conditions, to enable the state to provide a new public service: electricity supply. Private companies could not expropriate waterfalls unless they were already majority owners in the local area, a situation that did not change until the passage of the \cite{wra00}. 
%
%Even before this, when hydropower development was still seen as a public service, it was sometimes met with resistance by local people and environmental groups, particularly as the state begun to pursue large-scale projects after World War Two. I discuss the case law that developed in this regard, particularly in relation to procedural rules. I conclude that the public-sector characteristics of hydropower development led the courts to defer very broadly to the discretion of the executive and the legislature, also in relation to the content and scope of provisions in administrative law. 

In Section \ref{sec:twpp}, I pinpoint the end of the early era to the introduction of a general purpose expropriation authority introduced in the \cite{wra00}. For the first time in Norwegian history, waterfalls could now be expropriated for purely commercial gain, also by private companies. I go on to briefly present the resulting legal framework.
%I note that this change in the law was not given much attention by the executive committee that prepared the act. It was described merely as a ``simplification'' of existing rules. Moreover, the legislature did not address the change at all when the Act passed through parliament. In fact, the wording of the \cite{wra00} does not explicitly make clear that private expropriation is now permitted. However, the Act empowers the executive to decide, using directives, what class of legal persons can be given a license to expropriate waterfalls. Such a directive has been issued, with little or no debate, granting the possibility to benefit from an expropriation for hydropower development to ``anyone''.
%I present the law relating to expropriation of waterfalls in some depth, before I go on to consider more concretely how the procedure plays out in the context of for-profit takings. Here I anchor the presentation in the recent Supreme Court case of {\it Jørpeland}, where the issue of procedural legitimacy arose after a commercial company was granted permission to deprive local owners of waterfall rights that the owners wished to make use of in their own hydropower project.\footcite{jorpeland11}

In Section \ref{sec:jorpeland}, I use {\it Jørpeland} to show how owners' standing under administrative law is extraordinarily weak, particularly compared to the magnitude of their present-day interests in hydropower. Moreover, I argue that the Supreme Court adheres to a narrow perspective on the meaning of property protection, taking it to be an issue that begins and ends with the question of compensation. In my opinion, this fails to do justice to the most important issue that arises when waterfalls are taken for profit, namely the question of democratic legitimacy. 

%As discussed in Chapter \ref{chap:4}, the hydropower sector has now become depoliticised, expert-dominated and market-oriented. In light of this, my overarching argument is that the continued annihilation of local property rights threatens to render the state's functions in this sector subservient to the interests of the most powerful market actors, not the interests of the Norwegian people.

\section{Norwegian Expropriation Law: A Brief Overview}\label{sec:explaw}

As mentioned in Chapter \ref{chap:2}, the right to property is entrenched in s 105 of the Norwegian Constitution. There, it is made clear that when property is taken for public use, full compensation is to be paid to the owner. The public use requirement is understood very widely, or not regarded as a requirement at all. However, it is a rule of unwritten constitutional law that administrative decisions which affect the rights of individuals can only be carried out when they are positively authorised by law.\footnote{See generally \cite{hogberg11}.} Moreover, the Constitution is not understood as providing an authority to expropriate, it merely presupposes that expropriation is possible.\footnote{See, e.g., \cite[6]{fleischer86}.} Hence, the administrative branch must rely on other provisions that authorises compulsory acquisition of property on specific terms.

Historically, there was no general act relating to expropriation, and a range of different acts provided the necessary authority to expropriate for specific purposes, such as roads, public buildings, and schools.\footnote{See \cite[11-12]{nut54}.} Today, many of these authorities have been collected, broadened, and included in the \cite{ea59}.\footnote{Act no 3 of 23 October 1959 Relating to Expropriation of Real Property.} Still, many specific authorities remain, such as s 16 of the \cite{wra17}, which gives an automatic right to expropriate to the holder of a watercourse regulation license.

Following the \cite{wra00}, the general authority used to expropriate waterfalls has been included in the general act on expropriation.\footcite[2 no 51]{ea59}. Here it is stated that expropriation may take place in order to facilitate ``hydropower production''. In addition, it is made clear that expropriation can only be authorised if the benefits undoubtedly outweigh the harms. This sets expropriation orders apart from the various hydropower licenses discussed in Sections \ref{sec:wra00}-\ref{sec:ea} of Chapter \ref{chap:3}. In relation to the latter, in particular, the benefit is required to outweigh the harm, but it need not be ascertained that this is {\it undoubtedly} the case. However, the practical significance of this difference is limited. In particular, the additional requirement means that it should be clear that the benefit is greater, but does not imply that the benefit has to be significantly greater.\footnote{See \cite{lovenskiold09}.}

The authorising authority is the King in Council. However, the Act also makes clear that this authority can be delegated further, to ministries or other state bodies that the King in Council may instruct.\footnote{See \cite[5]{ea59}.} The compensation to the owner is determined following a judicial procedure administered by the special appraisement courts.\footnote{\cite[2]{ea59}.} 
The \cite{ea59} states that unless the Kind in Council decides otherwise, expropriation orders may only be granted to state or municipality bodies. This is formulated as a limiting principle, but in effect it serves as a general authorisation for the executive to decide, without parliamentary involvement, what class of legal persons may be granted expropriation licenses. 

For many purposes, directives have been issued that extend the class of possible beneficiaries to any legal person, including companies operating for profit. In 2001, such a directive was issued for the authority to expropriate in favour of hydropower production.\footnote{See Directive no 391 of 06 April 2001.} In addition to providing a general authority for expropriation, the \cite{ea59} also contains several procedural rules. These are collected in Chapter 3 of the Act. The Act sets out minimal requirements for what an application for an expropriation license must include, stating that it should make clear who will be affected, how the property is to be used, and what the purpose of acquisition is.\footnote{See \cite[11]{ea59}.} In addition, the Act requires the applicant to specify exactly what property they require, and to include information about the type of property in question and the current use that is made of it.

The owners are to be notified, and the starting point is that every owner is to be given individual notice, although this obligation falls away when it is ``unreasonable difficult'' to fulfil\footnote{See \cite[12]{ea59}, para 2.} If individual notification is too difficult, it is sufficient that the documents of the case are made available at a suitable place in the local area. A public announcement must also be made in the official notification publication of the government, as well as in two widely read local newspapers.\footnote{See \cite[12]{ea59}.}

The licensing authority is required to ensure that the facts of the case are clarified to the ``greatest extent possible''.\footnote{The Norwegian expression is ``best råd er'', which literally means ``best possible way''. See \cite[12]{ea59}, para 2.} This formulation seems very strict. At the same time, the exact meaning of the phrase  ``greatest possible extent'' is clearly not so easy to pin down. Moreover, administrative practice from several fields, including the hydropower sector, suggests that when expropriation takes place to implement a public plan, little attention is devoted to expropriation as a special issue, separate from planning considerations.\footnote{In relation to zoning plans, this has been made clear in a series of Supreme Court decisions, see \cite{namsos98,bo99}. In relation to hydropower, see Section \ref{sec:jorpeland}.}

A decision to grant an expropriation license must be justified, and the parties should be informed of the reasons for the decision.\footnote{See \cite[12]{ea59}, para 3.} This rule is largely superfluous, as the obligation to give reasons would in most cases also follow independently from administrative law, c.f., Section \ref{sec:paa67}.

The costs incurred by owners in relation to a pending application for expropriation against them is to be covered by the applicant.\footnote{See \cite[15]{ea59}.} The exact formulation is that the applicant 
is obliged to cover the costs that ``the rules in this chapter carry with them''. That is, the applicant is obliged to cover the costs that are related to the owners' rights pursuant to Chapter 3 of the \cite{ea59}. In practice, an applicant will be denied costs if the competent authority takes the view that they are unreasonable or disproportionate to his interests in the case.\footnote{If the case progresses to an appraisement dispute, the competent authority to decide on costs is the appraisement court. Otherwise, the decision is left with the executive. See \cite[15]{ea59}.}

%Particularly problematic are cases for which there is no clear division between those aspects of the case that relate to expropriation and those that relate to other licenses or land use planning more generally. This is the situation, for instance, in relation to hydropower development. In such cases, it is unusual for local owners to get any significant coverage of costs relating to the application processing. Legal expenses, for instance, are rarely covered unless they are incurred in relation to a subsequent appraisement dispute. This can be a problem for owners that wish to resist expropriation. Obviously, it is crucial for them to voice convincing objections already at the application processing stage.

In addition to the procedural rules in the \cite{ea59}, many rules of administrative law apply in expropriation cases. In the next section, I give a brief overview of this area of law, including the most important rules of the \cite{paa67}.

\subsection{The Public Administration Act}\label{sec:paa67}

Starting in the late 19th century, the importance of public administration gradually increased in Norway.\footnote{See \cite[8-12]{nut58}.} This development gained momentum after the Second World War, when administrative bodies also came to be placed more directly under centralised political control.\footnote{See generally \cite{gronlie00}.} At the same time, the established system based on legal expertise and strict adherence to the letter of the law was replaced by a form of management that actively sought to pursue political goals. As a result, the ambit of administrative decision-making power widened significantly. Many new administrative bodies were set up, while many of those already established were empowered greatly by statutory authorities that only specified their purpose and competence in broad strokes. Importantly, this style of legislation, which is still the norm, leaves great room for the exercise of administrative discretion, coupled with greater centralised control.

As administrative bodies became increasingly powerful, concerns arose regarding the relative lack of procedural safeguards to protect the individuals affected by administrative decisions. This concern was also fuelled by the fact that as the importance of state regulation increased, so did the power of the administrative branch to make decisions that would directly affect the rights and obligations of specific individuals.\footnote{See \cite[12-16]{nut58}.}

In response to this, a general statute was proposed that would set out some minimum standards of due process for administrative decision-making. This proposal eventually became the \cite{paa67}.\footnote{Act no 86 of 10 February 1967 Relating to Procedure in Cases Concerning the Public Administration.} The Act sets out the fundamental procedural principles that the executive is meant to follow when preparing to make administrative decisions. Some rules apply to any such decision, but a particularly important class of rules apply specifically to so-called {\it individual decisions}, which affect the rights and responsibilities of one or more specific persons.\footcite[2]{paa67} Clearly, owners of property covered by an expropriation license fall into this category, so that owners are considered protected parties to the individual decision to grant such a license.

Many of the rules in the \cite{paa67} mirror those of the \cite{ea59}, but tend to include broader and more detailed formulations. For instance, the duty to give advance notice is accompanied by more information about what kind of information such a notice must contain.\footnote{See \cite[16]{paa67}. Similarly to the advance notice rule in the \cite{ea59}, it may be permissible to not give individual notices if the parties are difficult to reach.} In particular, it is said that ``the advance notification shall explain the nature of the case, and otherwise contain such information as is considered necessary to enable the party to protect his interests in a proper manner''.\footcite[16]{paa67} 

Hence, it is not enough simply to inform the party that a case is under way, the Act also explicitly stipulates that the notice has to meet a minimum quality standard. In relation to expropriation of waterfalls this becomes potentially significant, especially in light of the practice I discussed in Chapter \ref{chap:3}, whereby expropriating parties sends out these notices themselves. One may ask, in particular, what owners are to think when they receive a letter from a commercial company stating that unless a friendly settlement can be reached, their waterfalls will be expropriated.\footnote{A formulation along these lines was used by the expropriating party in \cite{sauda09}. In general, according to my experience, a generic letter is sent by the developer to those private individuals who may be affected, with no individuation based on their interests in the case (e.g., based on whether they are waterfall owners or simply affected by the building works). Clearly, such an approach does not encourage waterfall owners to engage in the planning process in a manner commensurate with the fact that they own the natural resource that the planning pertains to.}

The duty to assess cases also follows from the \cite{paa67}, mirroring the rules of the \cite{ea59}. The formulation is similarly imprecise, as it declare that cases are to be ``clarified as thoroughly as possible'' before a decision is made.\footcite[17]{paa67} Importantly, however, the \cite{paa67} includes specific rules that oblige the authorities to inform parties about information they retrieve during their assessment of the case, and to submit such information for comments in so far as the party must be assumed to have an interest in it.\footnote{See paras 2 and 3 of \cite[17]{paa67}.}

The duty to justify and give reasons for administrative decision is also expressed more clearly in the \cite{paa67}. The duty applies to most individual decision, with some narrowly defined exceptions concerning cases when no party can be assumed to be dissatisfied, or when giving grounds would involve disclosing information to which the party is not entitled.\footnote{See \cite[24]{paa67}. Moreover, the King is authorised to limit the duty to give grounds when ``special circumstances so require''. All these exceptions are unusual, and hardly ever apply to hydropower cases.} As to the content of the reasons given, the authorities should mention the relevant rules authorising the decision, outline the factual assessment the underlies it, as describe the main considerations that have been decisive for the use of discretionary power.\footcite[25]{paa67} In case law, the duty to give reasons has become quite important, as the Supreme Court has declared that insufficient reasons can indicate shortcomings of the assessment, which might in turn lead to the conclusion that the decision is invalid.\footnote{See \cite{isene81,hauge00}.} However, the duty to give reasons in hydropower cases is understood to pertain to the licensing question as a whole, so that the authorities are not obligated to give individuated reasons to waterfall owners, pertaining specifically to the expropriation question.\footnote{See \cite{sauda09,jorpeland11} (discussed in more depth in Section \ref{sec:jorpeland}).}

Sometimes, the parties to an administrative decision are ill-equipped to look after their interests, even if the safeguards mentioned above are respected. This situation often occurs in hydropower cases, as many waterfall owners can not be expected to possess the technical, commercial, and legal knowledge necessary to understand what their ownership of this natural resource could entail, if an expropriation order is not granted. The \cite{paa67} establishes a general duty to provide guidance, to ensure that the parties are able to look after their interests in the ``best possible way''.\footcite[11]{paa67} However, it is explicitly stated that the level of guidance must be adapted to the circumstances and the capacity that the agency has for offering such assistance. At the same time, it is made clear that  the decision-making agency must assess, on their own motion, the parties' need for guidance.

To summarise, both the law of expropriation and general administrative law impose a range of procedural rules that ordinarily apply to expropriation cases. In principle, these apply also when waterfalls are expropriated. But in practice, they are completely overshadowed by the special rules that regulate the licensing procedure in such cases. I return to this issue in more depth in Section \ref{sec:jorpeland}. First, I elaborate a little on statutory rules that specifically target expropriation for hydropower, within the context of the relevant licensing procedures.

\section{Special Rules for Waterfalls}\label{sec:special}

As I mentioned in Chapter \ref{chap:3}, Section \ref{sec:wra17}, the \cite{wra17} establishes an automatic right to expropriate rights needed to undertake a watercourse regulation. This is not understood to include a right to expropriate waterfalls needed for the hydropower development. However, it includes a right to transfer water away from a river course for development somewhere else. This is a {\it de facto} license to expropriate a waterfall, as the water disappears form the river in which the owners have rights. This form of interference is also treated as an expropriation of waterfalls in relation to the compensation issue, as compensation is paid for the value of the waterfalls as such.\footnote{See \cite{jorpeland11}.} Formally, however, the interference is not considered a waterfall expropriation, but rather as an expropriation of a right to deprive waterfalls of water (and, hence, their value).

The question arises whether or not the rules in the \cite{ea59} and the \cite{paa67} apply in such cases. In theory, the answer is clear: they do apply. First, the rules in the \cite{paa67} express general principles of administrative law, pertaining to individual decision in general, including expropriation decisions, but also licensing decisions, in so far as they affect the rights of individuals. This condition is clearly fulfilled in cases pursuant to the \cite{wra17}. Second, the \cite{ea59} explicitly states that it applies to expropriation pursuant to the \cite{wra17}.\footnote{See \cite[30]{ea59}.} However, here it is also said that the rules in the \cite{ea59} only apply in so far as they are ``suitable'' and do not ``contradict'' sector-specific rules.\footcite[30]{ea59} This points to the potential caveat that while a range of rather strict procedural rules apply in theory, they may be ignored in practice, in so far as they are deemed redundant or unsuitable.

This is practically significant. In particular, the established practice among the water authorities is to regard the procedural rules in the \cite{wra17} as exhaustive.\footnote{This is made clear in \cite{jorpeland11}, where this practice also got a stamp of approval from the Supreme Court.} In addition, the material assessment requirement in the \cite{ea59} is not considered to have any independent significance alongside the assessment criterion in the \cite{wra17}.\footnote{Again, see \cite{jorpeland11}.} This is so even though case law on the former assessment criterion emphasises the interests of owners in a way that case law on the latter does not.\footnote{In addition, the formulation in \cite[2]{ea59} contains the additional qualification that the benefit of interference must ``undoubtedly'' outweigh the harm. No corresponding formulation is included in the \cite[8]{wra17}. Instead, there the formulation is that a license should ``normally'' not be given, unless the benefits outweigh the harms.}

%However, there is no doubt that the rules of the \cite{paa67} apply to takings of water rights pursuant to the \cite{wra17}. Moreover, there is no doubt that when a separate expropriation license is sought for waterfalls, these rules, as well as the rules in \cite{ea59} both apply. In practice, they nevertheless play a minimal role when the water authorities assess cases, as the assessment is unified, and the focus remains on balancing environmental and energy interest.

In light of this, it is very hard for owners to challenge the legality of a decision to allow expropriation of their waterfalls, particularly when expropriation takes place pursuant to the \cite{wra17}.\footnote{It follows from the discussion in Chapter \ref{chap:4} that large-scale development projects almost always involve a license pursuant to the \cite{wra17} (or such that the rules from this Act, including s 16 on expropriation apply pursuant to the \cite{wra00}).} Moreover, even if the \cite{wra17} does not authorise expropriation of waterfall rights, the water authorities tend to approach the expropriation in the same way, as long as the development is classified as a large-scale project. For such projects, the authorities tend to rely on a presumption that the conditions to permit expropriation are fulfilled whenever a development license may be granted.\footnote{See \cite{flatby08}.}

Hence, in order to defend themselves, owners must proceed in a roundabout manner by addressing the licensing question as such. Moreover, in order to argue that the expropriation is unlawful on procedural grounds, the owners must effectively demonstrate that the water authorities dealt with the case in contravention of sector-specific rules and practices pertaining primarily to the licensing question. This is a daunting task, particularly in light of case law developed during the period of monopoly regulation, when the courts tended to defer greatly to the judgement of the administrative branch.\footnote{The deferential stance was expressed most clearly in the {\it Alta} case discussed in Section \ref{sec:twp} below.}

In relation to procedural rules, the expropriation question is typically not recognised as a separate question at all. If environmental aspects have been dealt with thoroughly, procedural objections will tend to fail, even when these objections do not pertain to environmental issues, but rather to the interests of owners.\footnote{I believe this striking conclusion is justified by case law from recent years, which shows that courts continue to rely on the {\it Alta} case as the primary precedent when assessing procedural complaints in hydropower cases. See, e.g., \cite{sauda09,jorpeland11}. As I will show in Section \ref{sec:twp}, the {\it Alta} case had nothing to do with expropriation and property rights. Rather, it arose from environmental concerns (and concerns about property-less indigenous people). In particular, the procedural objections that were raised in {\it Alta} were specifically related to the assessment of environmental consequences, an assessment that is usually carried out very thoroughly by the authorities in hydropower cases (even more than usual in the {\it Alta} case, which was already very controversial).} Nevertheless, for a normative assessment of the law, it is interesting to consider how the practices and rules pertaining to expropriation of waterfalls, singled out for separate study, compare to the requirements of general administrative and expropriation law.

The rest of this Chapter is devoted to addressing this issue. I start by tracking the history of current practices. I believe this is quite important, because the context of expropriation has changed dramatically compared to the time when many of the current practices developed. Moreover, going back even further helps demonstrate that expropriation was once largely superfluous, and quite unheard of, in the Norwegian hydropower sector. For these reasons, I believe the historical perspective is very fruitful. Hence, I now embark on a chronological presentation, where I first address the period prior to the reform implemented by the \cite{ea90}.

\section{Taking Waterfalls for Progress}\label{sec:twp}

Historically, Norwegian law did not contain a general authority for the government to expropriate waterfalls, neither on its own behalf nor on behalf of private parties.\footnote{See \cite[29]{amundsen28}.} In the \cite{wra88}, a range of provisions authorised appropriation of water and land for specific purposes, but the criteria were narrow.\footnote{See \cite[69-85]{dahl88}. In addition, the purpose of expropriation was largely understood to be binding also on future use, so that the taker would not gain unrestricted control over the rights he acquired, but were obliged to use them in accordance with the public purpose for which expropriation was authorised. See, e.g., \cite[133-140]{rygh12}.} Waterfalls as such could never be made subject to expropriation, and expropriation of other rights could only be permitted in so far as the affected owners were not thereby deprived of any water power that they could reasonably make use of themselves.\footnote{See \cite[58,60]{dahl88}.}

More specifically, expropriation for hydropower development was only permitted when it would serve as an aid to owners who needed to acquire surrounding land in order to exploit waterfalls that they already owned.\footnote{See the \cite[15-16]{wra88}. See also the commentary in \cite[60-65]{dahl88}.} At the same time, owners of waterfalls could apply for licenses to engage in various industrial exploits, in some cases also when this would prove damaging to other landowners, for instance through deprivation of water or flooding.\footnote{See \cite[14]{wra88}. See also the commentary in \cite[54-60]{dahl88}.} These rules are similar to many of the rules found in contemporaneous mill acts from the US, c.f., the discussion in Chapter \ref{chap:2}, Section \ref{sec:hop}. As in the US, the kinds of takings in question here could be classified as economic development takings. However, the economic development potential itself was never taken in these cases. Rather, the takings only targeted additional rights that were needed in order for the owner to realise a potential for development of property that they already owned.

In fact, an important principle of expropriation law at this time was that no property could be taken if the taker's interest in that property was the same as that of its current owner.\footnote{See \cite[168-170]{dahl88}.} This applied regardless of whether or not the owners, subjectively speaking, were likely to pursue those interests optimally. Hence, expropriation of water power was ruled out already as a matter of principle. In particular, as the regulatory system of the day made private hydropower development possible, a private owner of a waterfall was held to possess a hydropower interest. As a result, they could not be deprived of waterfalls or water power by a taker whose interest was also related to hydropower development.

Following industrial advances, the interest in hydropower exploded in the late 19th century.\footnote{See \cite[58-59]{falkanger87}.} As a result, the state increasingly came to see it as a political priority to secure that waterfalls were used in the public interest. The most important expressions of this came in the form of the licensing acts presented in Chapter \ref{chap:3}, Sections \ref{sec:wra17} and \ref{sec:ica17}. Recall that the \cite{ica17} set up a licensing framework that would make it hard for speculators to purchase waterfalls. The \cite{wra17}, on the other hand, established the principle that the right to regulate the flow of water in a river system did not belong to the owner at all, but the state. Still, this did not imply any difference in the right to waterfalls for hydropower generation. This right still belonged to the local landowners.

However, around the same time, parliament passed legislation that authorised expropriation of waterfalls for the benefit of public bodies, also when the purpose was hydropower development.\footnote{Legislation that made it possible to expropriate waterfalls to the benefit of the municipalities was introduced in 1911, and a similar authority that authorised expropriation in favour of the state appeared in 1917, see \cite[29]{amundsen28}.} In 1940, these authorities were consolidated and integrated in the general water resources legislation, through the \cite{wra40} (which was later replaced by the \cite{wra00}). Still, the authority to expropriate waterfalls could be granted only to the state and the municipalities. Moreover, an expropriation order could in principle only be granted when it was necessary in order to implement state projects, or to satisfy the general demand for electricity. For the municipalities, moreover, it was explicitly required that the purpose of expropriation was electricity supply in the local district.\footnote{See the \cite[148]{wra40}. See also the commentary in \cite[201-210]{sorensen41}.}

Hence, the public purpose of expropriation remained in focus, and expropriation licenses could not in principle be granted to private or purely commercial entities. Moreover, it was felt that benefit sharing was required, so special rules were introduced to ensure that the taker would have to pay {\it more} than full market value compensation (typically a 25 \% premium, but in some cases the owner was also given a right to opt for compensation in the form of a proportion of the electricity output of the plant).\footnote{See \cite[70-91,184,210]{sorensen41}. I also note that in case private parties already owned more than 50 \% of the waterfall they wished to exploit, they could, on certain conditions, be granted permission to expropriate the remaining rights. See the \cite[55]{wra40}. See also the commentary in \cite[70-74]{sorensen41}. This was a novel rule in the 1940 Act, which contradicted earlier theories about the legitimacy of allowing expropriation for private benefit.}

Expropriation was intended to have a clear public character; in so far as the letter of the law was respected, it would appear that expropriation was indeed taking place in the public interest. Moreover, the public had to benefit directly, and locally (in case of municipal expropriation). Economic development in itself did not seem to be a sufficiently public purpose to justify expropriation. 

On the other hand, as I showed in Chapter \ref{chap:4}, the electricity supply in Norway just after the passage of the \cite{wra40} was quite well developed. Hence, in light of the seemingly strict public interest requirement of the expropriation authority, one would perhaps have expected expropriation for hydropower development to remain a rare occurrence. Instead, the use of expropriation for this purpose exploded after the war, as the state itself engaged much more actively engaged in hydropower development, also for more commercially oriented industrial exploits.\footnote{See generally \cite{skjold06}.}

Hence, despite the spirit and wording of the \cite{wra40}, this was the time when expropriation of waterfalls became a measure to facilitate economic development, although, for the time being, on non-commercial, politically governed, terms. In particular, the increased prevalence of expropriation seen during this time had little to do with any particularly pressing need to supply electricity to the people. Rather, it was a consequence of an increased political demand for industrial hydropower, combined with the fact that the hydropower sector was reorganised, so that it became a more tightly controlled monopoly, under increasingly centralised political control.\footnote{See \cite{skjold06,thue06b}.}

As I mentioned in Chapter \ref{chap:4}, many of the local, privately owned, hydropower plants were shut down during this period, as a result of an explicit policy meant to establish monopoly power in the sector. As the scale of development grew massively following the Second World War, local communities were marginalised in the decision-making processes concerning resource management. Moreover, the benefits from development would more often tend to accrue to urban areas. At the same time, the negative effects were of course mainly felt by the typically rural communities where the resource was found. At the same time, hydropower became increasingly politically sensitive, mainly as a result of growing environmental concerns about the consequences of large-scale development on nature.

The interpretation of the supply requirement in the \cite{wra40} was relaxed significantly over the years, especially following the development of the national electricity grid. It was no longer obvious, from a technical point of view, when exactly a hydropower development could be said to qualify as making a contribution to the local electricity supply. The electricity was not necessarily used locally, but, indirectly, also the local supply situation might be said to improve.

However, the rule that private parties could not expropriate waterfalls was still respected. It remained in place until 2000. Only then, some 90 years after the introduction of a general expropriation right for the state and the municipalities, did it become possible for any commercial entity to acquire waterfalls compulsorily for hydropower.\footnote{Moreover, this outcome was achieved by an executive directive, not an explicit act of the legislature, as discussed in Section \ref{sec:twpp}.}

In light of this, the vast majority of cases dealing with waterfall expropriation under Norwegian law can not be looked at as takings for profit, even though they increasingly became economic development takings. Certainly, the desire for economic development played a crucial part in motivating state and municipality development projects in post-war Norway. But their activities in this regard were not themselves commercial in nature. Rather, supplying electricity was regarded as a public service, one that would in turn stimulate commercial activity in other areas of the economy.\footnote{See generally \cite{thue06b,skjold06}.}

In the following subsection, I present the case law that developed during this time. In light of how the courts have chosen to approach recent controversies, this body of case law is still highly relevant, even though the context of interference has changed.

\subsection{Case Law from before 1990}\label{sec:prelib}

The period before liberalisation was not free from conflict regarding the legitimacy of measures undertaken to facilitate hydropower development in Norway. Indeed, the original move by which the state asserted control over the sector in the early 1990s was highly controversial. At this time, there was a general feeling of unease about how far the state could go in regulating and monopolising the hydropower sector without thereby depriving the owners of waterfalls of constitutionally protected rights.

This debate culminated in the conflict surrounding the rule of reversion that was introduced by the licensing acts passed between 1906 and 1917. As mentioned, the rule of reversion meant that in order to sell a waterfall to a private development, the owners and the purchaser had to apply for a license that was only ever granted on the condition that after some number of years, at most 60, the state would acquire the waterfalls without paying compensation. The question that arose was whether this was merely a regulation of the permitted use of waterfalls, or whether it should be regarded as expropriation, so that compensation would have to be paid pursuant to s 105 of the Constitution.

The conflict over this issue became fierce, with some influential legal scholars, attacking the rule of reversion as a ploy to confiscate Norwegian waterfalls without paying compensation to the owners.\footnote{See \cite{morgenstierne14}.} However, in a 4-3 decision, the Supreme Court held that s 105 did not apply, since reversion was merely a licensing condition, not an act of expropriation.\footcite{johansen18} No owner was compelled to hand over his property to the state, or to sell it to a private party so that the state would eventually acquire it.

One of the judges voting with the majority summed up their view by commenting that he would not regard it as expropriation if the state were to forbid sale of waterfalls to private parties altogether. Why then, he asked, should it be regarded as expropriation if such a sale was allowed to take place only on specific conditions? Against this, the minority argued that the licensing requirement as such was so severe that it had to be regarded as a {\it de facto} expropriation that entitled the owners to compensation. Moreover, as the purpose was clearly to ensure that waterfalls were eventually brought under state ownership, the minority did not think is was appropriate to consider reversion merely as a regulation of use.

After the Supreme Court upheld the rule of reversion, the legal foundation for the hydropower monopoly began to solidify. This monopoly developed gradually, however, and as I have already noted the use expropriation to facilitate hydropower did not become commonplace until after the Second World War. However, following the increased state-led activity in the sector after the war, conflicts surrounding hydropower development began to intensify once again. This time, however, the main objection was based on environmental concerns, not the position of local owners and communities.

After the Second World War, the state pursued very large-scale projects, involving significant levels of regulation of the water flow in large and important river systems. Moreover, it became common to divert water over great distances, to collect water from several different rivers in a common reservoir for joint exploitation. Such projects became known as ``gutter'' projects. In general, the opposition to hydropower grew proportionally to the scale of typical development projects.

While the focus was on environmental effects, the interest of local people also featured strongly in these debates. Moreover, the local interest were often aligned with the environmental interests. Large-scale hydropower projects, in particular, would tend to cause nuisance, or even significant loss, to traditional forms of agriculture. Therefore, in a situation when local owners could not themselves benefit commercially  from hydropower, their response was often to oppose it.

\subsubsection{{\it Alta}}\label{sec:alta}

The patterns of conflict that emerged during this time converged towards the highly controversial case of {\it Alta}, which came before the Supreme Court in 1982 after a long period of high-intensity conflict going back to the mid-seventies.\footnote{See \cite{alta82}. For a commentary by a leading legal scholar in Norway, see \cite{eckhoff82}.} This case features an added complication, since the affected locals mostly lacked formal title to the property they sought to defend. This was because the development in question would take place in the northernmost part of Norway, in the native land of the Sami people.

Norway has a history of discrimination against the Sami, and as their culture is largely nomadic, their land rights were never formalised in the law. As a result, the natural resources in Finnmark are largely owned by the state. However, the Sami have successfully struggled for their rights to use the land, particularly for their nomadic form of reindeer farming, with an extensive additional reliance on fishing and hunting.

The plans to develop large-scale hydropower in a Sami area therefore raised particularly strong criticism, also from environmental groups and groups fighting for aboriginal rights. A broad political mass movement was mobilised in opposition to the plans, eventually resulting in several serious cases of civil disobedience, including what might today be classified as ``terrorism''.\footnote{In particular, there were several instances when local people blew up equipment that was meant to be used to construct the hydropower plant. In one famous episode, the ``terrorist'' miscalculated, resulting in the loss of his own arm. Apart from a few such episodes, however, the protests were relatively peaceful.} The case was also dealt with by the court, as the Sami interests claimed, primarily on the basis of administrative law, that the development licenses that had been granted for the development were invalid.

At first sight, the {\it Alta} case is not particularly relevant to the question of expropriation. However, as the Norwegian regulatory system focuses on the development issue, with little or no separate attention paid to the issue of expropriation, the case has in fact been highly significant to the owners of waterfalls. It effectively serves as the primary measuring stick with which the executive and the courts assess the level of substantive and procedural protection that local people are entitled to, regardless of whether or not they have waterfall interests.\footnote{See \cite{sauda09,jorpeland11}.}

Due to the controversy surrounding the case, it was admitted directly from the district court to the Supreme Court in plenum. The presiding judge commented that as far as he knew it was the longest and most extensive civil case that the Court had ever heard.\footcite[254]{alta82} In an opinion totalling 138 pages, the Court argues that the decision to grant the license was valid. The opinion deals mostly with procedural rules. The substantive arguments, and arguments relating to international law, were not subjected to much scrutiny, as the Court was clearly quite confident that no objection could be raised against the licenses on such grounds.

However, the opponents of the development had pointed out a very wide range of purported shortcomings of the decision-making process. First, it was already clear that the original licensing application did not meet the requirements stipulated in the \cite[5]{wra17}. Essentially, the original application contained little more than technical details about the planned development, with hardly any identification or assessment of deleterious effects, neither to private nor public interests. This shortcoming had been openly acknowledge by the water authorities themselves, who had nevertheless initiated a public hearing, citing an electricity deficit in the northern part of Norway. 

The Supreme Court concluded that this was ``clearly unfortunate''.\footcite[265]{alta82} However, several reports and assessments had subsequently been provided by the water authorities, to fill the gaps left open by the initial application. For this reason, the Supreme Court argued, the initial mistakes might be irrelevant to the validity question, since it was licensing process as a whole that should be assessed. Shortcomings at specific stages in the assessment would not be given weight unless they could be seen to imbue the process with a dubious character overall.\footcite[265]{alta82}

The Court then moved on to assess whether the process as a whole fulfilled procedural requirements. This turned largely on the question of whether or not the authorities' assessment of deleterious effects fulfilled the criterion in \cite[8]{wra17}, c.f., \cite[16]{paa67} In this regard, those who objected to the license pointed to a range of negative effects that they believed had not been considered, or had not been considered in enough depth. 

In relation to nomadic reindeer interests, for instance, it was argued that the water authorities had failed to adequately consider the indirect consequences of development. These effects were described as ``catastrophic'' by an expert testifying before the Court. By contrast, the water authorities had not given much weight to the possibility of indirect consequences, citing the difficulty involved in attempting to quantify such effects.

After considering the reports and assessments in some depth, the Supreme Court did not find fault with the procedure in this regard. Importantly, the Court stressed that the water authorities were well aware of the possibility of indirect negative consequences. They simply chose, as a matter of expert discretion, not to place much weight on such consequences. This, according to the Supreme Court, could be regarded as an expression of disagreement with those claiming (later) that the effects would be catastrophic. As a result, the grounds for claiming procedural error disappeared, as the lack of attention directed at indirect consequences was held to reflect a factual assessment, made pursuant to a discretionary power that could not be made subject to judicial review.

More generally, the Court's opinion on this point reflects how hard it is to apply procedural rules in a context where it is assumed that the administrative branch has a wide margin of appreciation. Importantly, the Court makes statements of principle in this regard, that serve to limit the scope of judicial review under procedural rules in hydropower cases. In particular, the Court concludes that many of the relevant procedural rules in such cases by their very nature tend to be largely ``discretionary' '. As the licensing decision itself is a discretionary one, the argument goes, it is appropriate to admit to the executive a wide discretionary authority to decide for themselves also how to interpret many of the vague requirements of administrative law. By contrast, the view taken by the appellants, based on the idea that the content and scope of such rules is a purely judicial question, is described by the Court as ``overly formalistic''.

The Court makes a second statement of principle, namely that the scope of assessment required for the purposes of reaching a licensing decision is not in any event as extensive as the level of assessment that is required in a subsequent appraisement dispute. Hence, the meaning of the obligation to clarify cases to the best possible extent is put into perspective: assessments of deleterious effects may be omitted at the decision-maker's discretion even in circumstances when such assessments are practically relevant to the level of compensation payable and {\it will} in fact have to be provided at a later stage.

More concretely, in relation to the negative effects on fishing, the {\it Alta} Court conceded that the assessments could have been better, but pointed out that the purpose of assessment was only to answer yes or no to development, not to give a detailed presentation of its effects.\footcite[330]{alta82}. Crucially, the Court goes on to note that in so far as mistakes are uncovered as a result of insufficient assessment, this will influence the compensation payments and can also motivate subsequent regulation.\footcite[330]{alta82} 

In effect, the risk of error is downplayed by making reference to the compensation right and the regulatory authority of the state. This echoes the dichotomy mentioned in Chapter \ref{chap:3}, whereby there is a tendency in Norwegian law to perceive the interests of local people in purely monetary terms, while the state is assumed to be the sole protector of social and environmental values attached to property.

In relation to some negative effects of the {\it Alta} development, it was made apparent that they had not been considered at all. In addition, it was clear that erroneous information had been used in relation to some issues, particularly regarding alternative ways to meet the need for electricity in Finnmark, as well as the extent of this need. The Supreme Court agreed that this was a flaw, but held that it did not imply invalidity of the license. In this regard, a third statement of principle was made. The Court held, in particular, that the duty to consider alternatives -- different ways in which the public purpose could be satisfied -- was very limited.

It was sufficient to mention some alternatives, but it was not required to give any detailed assessment. This position of principle, in turn, was used by the Court to argue that the errors in the information provided about alternative were unlikely to have affected the outcome of the case.\footcite[346]{alta82} In fact, alternatives {\it had} been assessed in quite some depth. But since this was regarded as more than what was necessary, the factual errors that had been made were disregarded as insignificant. 

Interestingly, this was the conclusion despite the fact that clearly erroneous information had been  handed over to parliament, who had approved the plans on three separate occasions, always under reference to the precarious electricity situation in Finnmark. Here the Court established a principle whereby the nature of possible alternatives is considered a marginal issue in relation to assessment of licensing applications, regardless of what political decision-makers emphasise when {\it they} consider the case. 

In {\it Alta}, for instance, the Court's perspective appears to be at odds with how the parliament approached the case. There was little doubt that the favourable political assessment of the plans depended heavily on the perceived electricity crisis in Finnmark and the supply situation in Norway generally, as well as the perceived inadequacies of alternative solutions.

In relation to this question, the legal council acting for the state in {\it Alta} suggested explicitly that as these aspects came into focus only at the political stage of the decision-making, they were largely irrelevant to question of validity.\footcite[341]{alta82} This line of argument is rather striking, since the decision to grant the license was very much a political one. The information gathered by the water authorities, in particular, was put to use in a largely political decision-making context. In light of this, it seems that the procedural rules were, if anything, {\it more} important to observe in so far as they pertained to the quality of the factual basis that would be used in subsequent political assessments.

The Supreme Court did not approach the matter from this angle, but how exactly it reasoned in this regard is not clear from its opinion. In fact, it is worth noting how briefly the Court comments on this particular issue compared to other aspects of the case.\footnote{See also the surprise expressed in \cite[349-351]{eckhoff82}.} The Court instead goes into great detail regarding purported weaknesses of the licensing procedure that seem minor, comparatively speaking.

In relation to the duty to assess alternatives, the Court says nothing expect that the duty is very limited. For the details, which demonstrate factual inadequacies in the material given to the political decision-makers, the Court only refers briefly to the state's arguments. These arguments, based on the contention that the inadequacies were not significant, is accepted with no further discussion.\footcite[346]{alta82}

The dismissive attitude towards the duty to correctly assess alternatives is a controversial aspect of the {\it Alta}-decision which has been criticised by legal scholars.\footnote{See, e.g., \cite[580-584]{backer86}.} Today, maintaining such a dismissive attitude becomes particularly problematic also with respect to owners' rights. Indeed, alternatives are no longer limited to other public projects that can potentially provide the same public service at a smaller environmental and social cost. In addition, it is very natural to consider the possibility of owner-led projects proposed in commercial competition to the applicant's project. In so far as the duty to assess these alternatives is construed as loosely as the duty to assess alternatives was construed in {\it Alta}, it will hardly be reassuring for those owners of waterfalls that oppose commercial development projects based on their own hydropower plans.

In the next two sections, I will argue that no significant adjustments have been made to the way the water authorities approach their assessment of alternatives. Moreover, the dismissive attitude to this question in {\it Alta} was upheld in the case of {\it Jørpeland}, in a situation when the NVE had provided manifestly erroneous information to the Ministry.

\section{Taking Waterfalls for Profit}\label{sec:twpp}

As I mentioned in the previous section, private companies could not expropriate waterfalls in Norway prior to the passage of the \cite{wra00}. Moreover, the public purpose requirement was formulated quite strictly, particularly in cases when the development was undertaken by municipality companies. I also mentioned how the hydropower sector developed after the Second World War from a sector dominated by private and municipality companies, to a sector dominated by the state. This, in turn, was accompanied by increased conflicts and doubts regarding the legitimacy of the established licensing procedures, particularly the highly centralised nature of the decision-making process. 

Even so, the debate at this time was still very much anchored in a system that presupposed political management of the hydropower sector as a public service provider. Importantly, the conflicts rarely, if ever, involved significant commercial interests on the part of the local waterfall owners. Many critics argued that the fiscal interest of the state could not be used to justify destruction of nature and local patterns of land use. But in financial terms, the value of what was destroyed was typically negligible compared to the value of the hydropower development.

As a result, controversies relating to the legitimacy of interference involved only the waterfall rights at their periphery. More focused conflicts involving waterfalls specifically arose in relation to the question of compensation, but the issues typically discussed in this regard were also of relatively minor structural importance, although they could of course be important enough for the individuals directly affected.

In Chapter \ref{chap:3}, I presented the reform of the energy sector of the early 1990s, after which hydropower development has been regarded as a commercial pursuit. Following the regulatory reform, a new general statute dealing with water resources was also proposed, eventually leading to the passage of the \cite{wra00}. This Act provided the first every authority for the state to allow developers to take waterfalls compulsorily for profit. Moreover, it made possible the later executive directive by which waterfalls could be expropriated and handed over to {\it any} legal person, including private companies.

After the legal and regulatory reforms of the 1990s, takings of waterfalls for hydropower have become takings for profit. But this change in the function of expropriation received little attention when these reforms were introduced. Moreover, when the \cite{wra00} was proposed, the increased scope of expropriation was not singled out for political consideration when the Ministry presented their proposal to parliament. The new expropriation authority for waterfalls is described merely as a ``simplification'' of the law.\footcite[223-225]{otprp39}

This is hardly accurate. However, the commission appointed by the Ministry to prepare the Act also adopted a very low-key approach to expropriation. The commission mentioned that its proposals would imply increased scope for expropriation, but it did not discuss the desirability of this in any depth.\footcite[235-237]{nou94} 

The report from the commission, which totalled almost 500 pages, devotes only three pages to the new expropriation authority. Here the committee notes that a range of different authorities for expropriation has long co-existed in the law, with many of them positing strict and concrete public interest requirements as a precondition for granting a license. This, the commission argues, is not a very ``pedagogical'' way of providing expropriation authorities.\footcite[235]{nou94} Moreover, the commission notes that it runs the risk of omitting important purposes for which expropriation should be possible. Hence, the commission proposes to replace all older authorities by a sweeping authority that makes expropriation possible for any project that involves ``measures in watercourses''.\footcite[235-236]{nou94}

The commission comments that their formulation might seem wide, but remark that this is not a problem since the executive can simply refuse to issue an expropriation order in so far as they regard expropriation as undesirable.\footcite[235]{nou94} The commission does not reflect on the  constitutional consequences of such a perspective, neither in relation to property rights nor in relation to the balance of power between the legislature, the executive and the courts. Instead, the commission offers a very brief presentation of the rationale behind dropping the local supply restriction for municipal expropriation. Here they remark simply that these rules complicate the law and might make desirable expropriations impossible.\footcite[235]{nou94} No clarification is offered as to what kind of desirable expropriations the committee think might be left out. 

Importantly, the committee do not relate their proposals to the recent market-based reform of the energy sector. Hence, the obvious practical consequences of their proposal, namely that expropriation will be made available as a profit-making mechanism for commercial companies, is not discussed or critically assessed.

The issue of {\it who} should be permitted to benefit from an expropriation license is also dealt with very superficially. In this regard, the commission structures their presentation around the so-called {\it redemption} rule of the \cite{wra40}. As mentioned briefly in Section \ref{sec:twp}, this rule made it possible for the majority owners of a waterfall to compulsorily acquire minority rights, if this was necessary to facilitate hydropower development. Hence, it was a rule that provided only a limited opportunity for private takings, restricted to owners themselves or external developers that had been able to reach a deal with a locally based majority.

The main justification given by the commission for introducing a general private takings authority is that the special redemption rule had not been much used in practice.\footcite[236]{nou94} Why this is an argument in favour of opening up for private expropriation in general is not made clear. Indeed, it seems just as natural to regard it as an argument {\it against} doing so. Why extend the possibility for private expropriation if the demand for such expropriation has been limited? 

Presumably, the commission thought there would be a demand for such expropriation in the future, but this is not stated explicitly, nor is the appropriateness of meeting such a demand discussed. As to the requirement that private takers must already control a majority of the waterfall rights in the local area, the commission only remarks that it regards such a restriction as old-fashioned.\footcite[236]{nou94} No discussion is offered regarding the consequences for local waterfall owners.

Since the passage of the \cite{wra00}, it has become clear that the new authority for expropriation is a highly significant, and controversial, aspect of the Act. During the last 14 years, an unprecedented number of cases has raised the issue of legitimacy of expropriation of waterfalls. Today, practically all cases of expropriation imply that local owners are deprived of a development potential in favour of a commercial actor seeking development of broadly the same kind. According to the law before the \cite{wra00}, expropriation in such circumstances would be hard to justify in general, and impossible for private companies.

In {\it Sauda}, a case before the court of appeal, this came into focus, as the waterfall owners protested a license that granted a private company the right to expropriate their waterfalls.\footnote{See \cite{sauda09}.} The owner's principal argument was that the executive directive granting such rights to private parties was in fact invalid, since it had not been sanctioned by parliament. Formally, this argument was rather weak, since the \cite{ea59} had been amended in such a way that the executive was explicitly authorised to determine the class of legal persons that could be granted a license to expropriate for hydropower development. However, the owners argued that the executive had not appropriately informed parliament that this would be the consequence of the amendment. In particular, the amendment itself had been passed as a mere formality following the adoption of the \cite{wra00}. 

The owners backed up their case by pointing to interviews with two of the members of the parliamentary committee that had prepared the case. Neither of them could recollect that they had been aware that a right to expropriate for private developers would result from the Act they had passed. This was not conveyed to them by the executive. Moreover, it was not explicitly stated anywhere in the Act itself. Rather, it followed implicitly from three different sections in two separate Acts. In the entire collection of preparatory documents, the change was discussed only once, in the report from the committee to the Ministry. This, the owners argued, meant that the purported authority was not in fact constitutionally valid, since parliament had not intended that it should be introduced in Norwegian law.

Unsurprisingly, this argument was rejected. It had to be assumed that parliament understood the consequences of their own legislative actions. However, while the owners lost the validity dispute, the level of compensation was dramatically increased compared to earlier practice. Because of this, the development company appealed the decision to the Supreme Court, with the owners lodging a counter-appeal regarding the question of legitimacy. The Supreme Court decided not to hear the case. 

Indeed, it had recently addressed the compensation question from a similar angle in the paradigmatic case of {\it Uleberg}.\footnote{See \cite{uleberg08}.} Here a new principle of market-value compensation was introduced, for those cases when small-scale development by owners was deemed to have been ``foreseeable'' in the absence of expropriation. In these cases, the owners can now expect at least 10-20 times more in compensation than they would get under the traditional approach, which was based on a purely theoretical estimation of the value of the waterfall to the taker.\footnote{For a more in-depth presentation of this, I refer to \cite[71-76]{dyrkolbotn15}.}

In addition to raising the compensation issue and the issue of constitutional legitimacy, {\it Sauda} also raised several procedural questions regarding how the water authorities prepare cases involving expropriation of waterfalls. The owners were unsuccessful also in this regard. However, their procedural arguments foreshadow the later case of {\it Jørpeland}, where the owners were successful in the district court, which held that the procedures developed in the takings for progress era were no longer appropriate.\footnote{See the decision by Stavanger Tingrett in \cite{jorpeland09}.} 

This decision was overturned on appeal, however, a decision that was in turn upheld by the Supreme Court, in a decision relying on the precedent set by {\it Alta}.\footnote{See the decision by Gulating Lagmannsrett in \cite{jorpeland11a}. See also the Supreme Court decision in \cite{jorpeland11} ({\it Jørpeland}).}  I will now consider this case in some depth, since this will help bring out how administrative practices relating to expropriation for hydropower actually function in the context of commercial development. In particular, it will help shed light on the situation that arises when the owners oppose the expropriation because they have competing development plans.

\section{{\it Ola Måland v Jørpeland Kraft AS}}\label{sec:jorpeland}

The expropriating party in this case was a public-private commercial partnership, Jørpeland Kraft AS. This is a limited liability company which is jointly owned by Scana Steel Stavanger AS, with 1/3 of the shares, and Lyse Kraft AS, who is the majority shareholder, with the remaining shares. The former is a private steelworks company located in the small town of Jørpeland in Rogaland county, south-western Norway. Historically, this company was a major employer in Jørpeland, which is located by the sea, next to a mountainous area.

The main source of energy for the steel industry in Norway has always been hydropower and Scana Steel Stavanger AS was no exception. The company used energy harnessed from the rivers in the area, particularly the river which reaches the see near Jørpeland itself. Moreover, the water from this river is supplemented by water from other rivers in the area that are diverted so that they can be exploited along with the water from the Jørpeland river.

Recently, Norwegian steel companies have become less profitable, due in great part to increased foreign competition and a significant increase in cost of operation associated with this type of industry in Norway.\footnote{Salary costs, in particular, have become prohibitive. See, e.g., \emph{Information Booklet about Norwegian Trade and Industry}, published by the Ministry of Trade and Industry in 2005.} This has led to many such companies shifting their attention away from labour-intensive steel production, focusing instead on producing electricity, selling it directly on the national grid. Jørpeland Kraft AS was established as part of such a move to exploit the energy resources in Jørpeland, and the role played by Lyse Kraft AS is an important one.

As I discussed in Chapter \ref{chap:3}, Norwegian law favours companies where the majority of the shares are held by public bodies. Lyse Kraft AS is publicly owned, with the city of Stavanger as the main shareholder. Hence, it is a very valuable partner. Moreover, Lyse Kraft AS  is also responsible for the electricity grid in the region, so is well-positioned to provide easy access to the electricity market. 

Lyse Kraft AS was established as a merger between several local monopoly companies in the Stavanger region which were reorganised following liberalisation of the sector in the early 1990's. As discussed in Chapter \ref{chap:3}, there is little doubt that old monopolists still enjoy considerable power and influence, particularly in relation to the state agencies that regulate the industry.\footnote{In fact, Lyse Kraft AS is good example, particularly since their power appears to have \emph{increased}, as the restraints imposed both by the non-commercial nature of former monopolists, and the local, political, anchoring of such companies, have disappeared.} This is another reason why they can serve as valuable partners for private companies wishing to make a profit from Norwegian waterfalls.

As attention shifts from harnessing rivers for the purpose of industrial production to the purpose of producing electricity to sell on the national grid (and, increasingly, to export abroad), the main variables that determines the profitability of the undertaking also changes. On the cost side, what matters the most is the one-time investment required to construct the hydropower plant.\noo{\footnote{For an overview of the considerations made when assessing the commercial value of hydropower, I point to \cite{jensen07}. In fact, due to the importance that small-scale hydropower has assumed in recent years, investigating models for investing in such projects has become an active field of research in Norway, see for instance \cite{investment}.}} Maintaining and operating a hydropower station tends to be comparatively inexpensive. On the income side, what matters is the price of energy on the electricity market, a market that is no longer anchored in local conditions of supply and demand.

Importantly, as long as energy production is the sole focus, the business no longer depends in any significant way on the local labour force. Hence, large-scale exploitation becomes much more profitable than the medium or small-scale power plants that would otherwise be suitable for local industrial exploits. Indeed, it was in keeping with a general trend in Norway when Jørpeland Kraft AS, following  their new commercial strategy, proposed to undertake measures to increase their energy output. This could be achieved relatively cheaply, by further constructions aimed at diverting water away from nearby waterfalls into dams that were already built to collect the water from the Jørpeland river.

One relatively small waterfall from which Jørpeland Kraft AS suggested to extract water is owned by Ola Måland and five other local farmers. This waterfall is not located in Jørpeland kommune and it does not reach the sea at Jørpeland. Rather, it runs through the neighbouring municipality of Hjelmeland, on the other side of a mountain range, until it eventually reaches the sea at Tau, another neighbouring municipality. The plans to divert the water would deprive original owners of water along some 15 km of riverbed, all the way from the mountains on the border between Hjelmeland and Jørpeland, to the sea at Tau. Far from all the water would be removed, but the flow of water would be greatly reduced in the upper part of the river known as {\it Sagåna}, the rights to which is held jointly by Ola Måland and five other local farmers from Hjelmeland.

The water in question stems from the \emph{Brokavatn}, located 646 meters above sea level, where altitude soon drops rapidly, making the waterfall suitable for hydropower development. Plans were already in place for such a project, which would use the water from just below the altitude of Brokavatn, to the valley in which the original owners' farms are located, at about 80 meters above sea level. A rough estimate of the potential of this project was made by the NVE itself, stating that the energy yield would be 7.49 GWh per annum.\footnote{See \cite{jorpeland09}.} This is about five times more than the water from Brokavatn would contribute to the project proposed by Jørpeland Kraft AS.\footnote{See \cite{jorpeland09}.}

Importantly, the estimate was not made in relation to the expropriation case, but as part of a national project to survey the remaining energy potential in Norwegian rivers.\footnote{The survey was carried out in 2004 and its results are summarised in \cite{jensen04}.} Ola Måland and the other owners of the waterfall were not identified as significant stakeholders in this regard and they were not notified of the assessment that had been made. Moreover, when Jørpeland Kraft AS submitted an application to the NVE for permission to divert the water, the owners were not notified by the NVE.\footnote{However, a generic orientation letter was apparently sent by Jørpeland Kraft AS, a letter that the owners themselves could not remember having received.}

Rather, the approach to the case was the traditional one, with an assessment directed at the environmental impact. Many interests groups were called on to comment on consequences in this regard, and public debate arose with respect to the balancing of commercial interests and the desire to preserve wildlife and nature.

Despite not being asked to do so, one of the owners, Arne Ritland, commented on the proposed project. He did this in an informal letter sent directly to Scana Steel Stavanger AS. In this letter, he inquired for further information and protested the proposed diversion of water from Brokavatn. He also mentioned the possibility that an alternative hydropower project could be undertaken by original owners, but he did not go into any details regarding this, stating only that such a locally owned hydropower plant had previously been in operation in the area. The plant he was referring to dates back to the time before there was a national grid and was directed at local supply of electricity. It has since been shut down, in keeping with the general trend mentioned in Chapter \ref{chap:3}.\footnote{See \cite{jorpeland09}.}

Arne Ritland received a reply from Scana Steel Stavanger AS, in which it said that more information on the project and its consequences would soon be provided. Ritland did not pursue the matter further at this time. Meanwhile, Scana Steel Stavanger AS submitted his letter to the NVE, who in turn presented it as a comment directed at the application as part of the hearing. This prompted Jørpeland Kraft AS to undertake their own survey of alternative hydropower in Sagåna. The conclusions, but not the report itself, were sent to the water authorities. The original owners were not informed, and they were not asked to comment, or even told that such an investigation of the commercial potential in their waterfalls was being conducted by the expropriating party, as a response to Ritland's letter.

Despite being presented with the issue, the water authorities did not take steps to investigate the commercial potential of local hydro power on their own accord. Moreover, the conclusion presented by Jørpeland Kraft AS did not go into details, but merely stated that if the local owners decided to build two hydropower plants in Sagåna, then one of them, in the upper part of the river, would not be profitable, neither with nor without the contested water. The other project, in the lower part, could apparently still be carried out, even after the diversion. No mention was made of what the original owners stood to loose, nor was there any argument given as to why it made sense to build two separate small-scale power plants in Sagåna. Nevertheless, the NVE handed the expropriating party's findings over to the Ministry, without conducting their own assessment and without informing the original owners. 

In addition to the report made by Jørpeland Kraft AS, Hjelmeland kommune, the local municipality government, also commented on the possibility of local hydropower. In their statement to the NVE, they directed attention to the data in the NVE's own national survey, which suggested that a single hydropower plant in Sagåna would be a highly profitable undertaking. On this basis, they protested the diversion, arguing that original owners should be given the possibility of undertaking such a project. This statement was not communicated to the original owners, and in their final report it was dismissed by the NVE, who stated that the most efficient use of the water would be to transfer it and harness it at Jørpeland.

In addition to the statement made by Ritland, one other property owner, Ola Måland, commented on the plans. He did so without having any knowledge of the commercial potential of the waterfall. Moreover, he had not been informed of the statement made by Hjelmeland Kommune. On this basis, he expressed his support for Jørpeland Kraft's plans, citing that the risk of flooding in Sagåna would be reduced. He also phrased his letter in such a way that it could be interpreted as a statement on behalf of the owners as a group. However, Måland was the only person who signed.

In the final report to the Ministry, the NVE refer to Måland's letter and state that the original owners are in favour of the plans. For this reason, the NVE argues, the opinion of Hjelmeland kommune should not be given any weight. The NVE neglects to mention Arne Ritland's statement in this regard. Moreover, earlier in the report, where all incoming statements are reported, Ritland is referred to as a private individual, while Ola Måland is referred to as a property owner who speaks on behalf of the owners as a group.

The report made by the NVE was not communicated to the affected local owners at all, so the owners had no chance of correcting the mistakes. However, the report was sent to many other stakeholders, including Hjelmeland kommune. In light of NVE's conclusions, the municipality changed their original position and informed the Ministry that they would not press for local hydropower, since this was not what the original owners wanted.

All of this happened without the owners' knowledge. However, while the case was being prepared by the water authorities, the original owners had begun to seriously consider the potential for hydropower on their own accord. In late 2006, Jørpeland Kraft's application reached the Ministry and a decision was imminent. At the same time, the owners were under the impression that they would receive further information before the case progressed to the assessment stage. 

As the owners had now come to realise the commercial value of the water from Brokavatn in their own project, they approached the NVE, inquiring about the status of the plans proposed by Jørpeland Kraft AS. They were subsequently informed that an opinion in support of plans had already been offered to the Ministry. This communication took place in late November 2006, summarised in minutes from meetings between local owners, dated 21 and 29 of November. On 15 of December 2006, the King in Council granted a concession for Jørpeland Kraft AS to transfer the water from Brokavatn to Jørpeland.

At this point, it had become clear to the original owners that the water from Brokavatn would be crucial to the commercial potential of their own project. They also retrieved expert opinions that strongly indicated that the NVE was wrong when they concluded that diverting the water would be the most efficient use of the water for hydropower. In light of this, the owners decided to question the legality of the concession (with the corresponding permission to expropriate). They argued, in particular, that the administrative decision was invalid.

According to the owners, the expropriation was materially unjustified. Moreover, they contended that the administrative process had not fulfilled procedural requirements. The district court, Stavanger tingrett, held in favour of the owners.\footnote{See \cite{jorpeland09}.} The court emphasised that the authorities were obliged to properly take into consideration the fact that the waterfalls could have been exploited commercially by the original owners themselves. Moreover, the court noted that in order to ensure a proper assessment, the authorities should have ensured that the owners were kept informed about the progress of the case, so that they would have an opportunity to comment in a timely fashion. The court also noted that the administrative decision was based on factual mistakes, regarding the owner's opinion of the plans, as well as the energy potential of Jørpeland Kraft's plans compared to an owner-led project in ``Sagåna''. This, the court held, meant that the concession had to be declared invalid.

This view was rejected by the court of appeal, Gulating lagmannsrett, which held that sufficient steps had been taken to clarify the commercial interests of the owners.\footnote{See \cite{jorpeland11a}.} Moreover, the court held that established practices regarding the preparation and evaluation of such cases -- dating from a time when it was not feasible for original owners to undertake hydropower schemes -- still provided adequate protection. The owners appealed the decision to the Supreme Court, who decided to hear the case. However, the appeal was eventually rejected, and the Supreme Court went even further than the court of appeal when they declared that established practices were beyond reproach.\footnote{See \cite{jorpeland11}.}

In the following section, I present the main legal arguments relied on by the parties, as well as a summary of how the three national courts judged the case.

\subsection{The Legal Argument}\label{view}

First, they owners argued that procedural mistakes had been made in preparing the case. Second, they argued that according to Norwegian expropriation law, it was not permissible to expropriate in a situation such as this, when the loss of energy and commercial potential would outweigh the gain to those same interests. Moreover, it seemed to the original owners that allowing expropriation would only serve to benefit the commercial interests of Jørpeland Kraft AS, and that it would do so to the detriment of both local and public interests. 

For this reason, the owners also contended that the concession should be regarded as an abuse of power, a manifestly ill-founded decision which could not be upheld.\footnote{There are at least two different ways in which to argue such a point under Norwegian law. One is with respect to water law and general administrative law, whereby clearly ill-founded decisions can be overturned by the courts, even when they involve discretion on part of the executive, which is otherwise not subject to review by the courts. Secondly, an argument can be made with respect to the Norwegian Constitution, s 105, which gives property a protected status. The former is usually more effective, but in both cases, quite a severe transgression will have to be established before courts consider it within their competence to overturn discretionary decisions.} 

Third, the owners argued that the government had not fulfilled its duty to consider the case with due care. In particular, the assessment made with respect to the interests of the local community at Hjelmeland, and the local owners residing there, was not adequate. Particular attention was directed at the fact that local owners had not been informed about the progress of the case, and had not been told of, or asked to comment on, those preparatory steps that explicitly sought to assess their interests.

Fourth, the owners argued that irrespectively of how the matter stood with respect to national law, the expropriation was unlawful because it would be in breach of the provisions in P1-1 of the ECHR regarding the protection of property.

Jørpeland Kraft AS protested all these objections to the expropriation, arguing that it was the responsibility of the owners to actively provide information about possible objections against the project and that the process had therefore been in accordance with the law. Unfortunate misunderstandings, if any, should be attributed to the fact that original owners had neglected their responsibilities in this regard. Moreover, Jørpeland Kraft AS argued that it was not for the courts to subject the assessment of public and private interests to any further scrutiny, since this was a matter for the government to decide.

Indeed, according to Norwegian national law, it is traditionally held that unless the exercise of power it clearly unjustified, the courts do not have the authority to overturn decisions based on discretion, unless it can be demonstrated that the government has made procedural mistakes. While this view has become somewhat more relaxed in recent years, with a standard of \emph{reasonableness} becoming increasingly important, the inadmissibility of court interference in administrative discretionary decisions is still very much a part of Norwegian national law.\footnote{See \cite{eckhoff14}, in particular, Chapters 24 and 29.}

Finally, Jørpeland Kraft AS argued that there was no issue of human rights at stake in the case. They argued for this by simply pointing to the fact that the procedural rules had been followed and that  the material decision was beyond reproach. Hence, no human rights issues seemed to arise. Moreover, Jørpeland Kraft AS suggested that since the owners would be compensated financially by the courts for whatever loss they incurred, no human rights issues could possibly arise.

The matter went before Stavanger Tingrett who decided in favour of the owners on 20 May 2009.\footnote{See \cite{jorpeland09}.} In the following, I offer a presentation of the reasons given by this court, leading to the conclusion that the expropriation was unlawful and that the diversion of Brokavatn could not be carried out.

\subsubsection{Stavanger Tingrett}

Stavanger tingrett agreed with the original owners that the decision to grant concession was based on an erroneous account of the relevant facts. Moreover, the court concluded that it was evident, from the assessment carried out by the NVE themselves (in their national survey of small-scale hydropower potential), that allowing the applicants to use the water from Brokavatn in their own hydroelectric scheme would be the most efficient way of harnessing the hydropower potential. This, the court noted, directly contradicted what the NVE had stated in their report.

In addition, the court noted that Hjelmeland kommune had in fact referred to the national survey in their objection against the application. Hence, the court found that it was a procedural error that the NVE had not considered the assessment of hydropower potential contained herein.

The court substantiated their decision by giving several direct quotes from the report made by the NVE. For instance, from p 199, as quoted by Stavanger tingrett (my translation):
%\begin{quote}Hjelmeland kommune ser helst at kraftressursene i vassdraget blir utnyttet av lokale %grunneiere. 
%Dette står i kontrast til uttalelsen fra grunneierne selv som ønsker at overføring blir gjennomført, 
%slik at flom og erosjonsskader kan bli noe redusert. NVE mener at den beste utnyttelsen med tanke 
%på kraftproduksjon vil være å tillate overføringen da en slik løsning vil innebære at vannet utnittes i 
%størst fallhøyde. Når dette samtidig er grunneiernes eget ønske har vi ikke tillagt Hjelmeland 
%kommunes synspunkt på dette noen vekt
%\end{quote}
%Our own translation follows below: 
\begin{quote}
Hjelmeland kommune would like the hydroelectric potential in the waterfall to be exploited by 
local property owners. This contrasts with the statement given by the property owners 
themselves, who wish that the transfer of water takes place, so that damage due to flooding can be 
somewhat reduced. NVE thinks that the best use of the water with respect to hydroelectric 
production is to allow a transfer, since this means that the water can be exploited over the greatest
distance in elevation. When this is also the property owners' own wish, we will not attribute any 
weight to the views of Hjelmeland kommune.
\end{quote}

Stavanger tingrett concluded that as this was a factually erroneous account of the situation, the decision made to allow transferral of the water could not be upheld. Summing up, the court offered the following assessment of the case (my translation):

\begin{quote}
It is the opinion of the court, having considered how the case was prepared by the authorities, that the factual basis for the decision made by the government suffers from several significant mistakes and is also incomplete.
\end{quote}

In light of this, Stavanger tingrett concluded that the decision to grant concession for diversion of water was invalid. Here the court relied on a well-established principle of administrative law: while the exercise of discretionary powers is usually not subject to review by court, a decision based on factual mistakes is nevertheless invalid if it can be shown that the mistakes in question were such that they could have affected the outcome. This is not provided for explicitly in statue, but it is one of the core unwritten legal principles of administrative law.\footnote{See \cite{eckhoff14}.}

Concerning the second requirement, that the factual mistakes could have affected the outcome, Stavanger tingrett found that it was clearly fulfilled in this case since the hydropower project suggested by original owners was in fact a \emph{better} use of the resource, even with respect to public interests. In any event, the requirement with regards to factual and procedural mistakes is only that the mistakes \emph{could} have affected the outcome; in the presence of mistakes, the burden of proof is shifted onto the party seeking to defend the decision.

Since Stavanger tingrett held that the license to allow diversion was invalid because it was based on factual mistakes, there was no need to consider claims regarding the legitimacy of the decision with respect to human rights law. Stavanger tingrett did conclude, however, by making a more overarching assessment of the case. Here they state that the procedure followed in preparing the case had not taken sufficient regard of owners' interests and that this was the likely cause of the mistakes that had been made. The court also stated that the standard of assessment when considering owners' interests had to be interpreted more strictly now that local hydropower was an option available to original owners.

\noo{In this regard, t also seems that Stavanger Tingett found some additional support in its interpretation of Norwegian law that was based on human rights concerns, especially the fact that expropriation, in circumstances such as those of this case, appeared to be a major interference in the rights of owners, and that established practice developed under a different regulatory regime was therefore no longer able to provide adequate protection.}

Jøpeland Kraft AS appealed the decision and the case then went before the court of appeal, Gulating lagmannsrett, who found in favour of Jørpeland Kraft AS. 

\subsubsection{Gulating Lagmannsrett}

In their argument, the court of appeal do not rely on direct assessment of the report made by NVE, nor do they mention the expert statements retrieved by the opposing parties. Instead, the court of appeal base their decision on general considerations concerning the need for efficient procedures in hydropower cases. Such reasoning provides the apparent grounds for making the following crucial observation concerning the facts:

\begin{quote}... It was not a mistake to take Ola Måland's statement into consideration, as he was, and still is, a significant property owner. NVE's statement to the effect that granting the concession will facilitate a more effective use of the water seems appropriate, as it refers to a current hydroelectric plant that exploits a waterfall of 13.5 meters.
\end{quote}

The court of appeal do not mention the statement made by Hjelmeland kommune, nor do they comment on the fact that alternative hydropower, as suggested by the municipality, would amount to exploiting the waterfall over a difference in altitude of some 550 meters. In fact, the hydroelectric plant that they do mention has nothing to do with Ola Måland and the other owners, but exploits the same water further downstream, on property owned by persons that were no longer parties to the dispute (since their interests were obviously quite negligible).

This waterfall was only brought up in the testimony of a representative from the NVE. When pressed on the matter, this representative claimed that the reasonable way to interpret the assessment made by the NVE was to see it as an assessment concerning the existing hydropower plant further downstream. In light of the statement provided by Hjelmeland kommune, to which the report explicitly refers, this is a manifestly ill-founded interpretation. But the regional court adopted it, without further comment.

The court of appeal seems to have assumed, quite generally, that the practices traditionally adopted by the water authorities in hydropower cases would still provide adequate protection for original owners. In this regard, the court of appeal seem to base their stance on an overarching appeal to the need for efficient procedures.

The court of appeal's decision was appealed by Ola Måland and the other owners. The Supreme Court decided to consider the juridical aspects of the case.\footnote{The appeal concerning the assessment of facts would not be considered. Hence, the Supreme Court would base themselves on the presentation of facts given by the court of appeal.}

\subsubsection{The Supreme Court}

The Supreme Court ruled in favour of Jørpeland Kraft AS.\footnote{See \cite{jorpeland11}.} In their decision, they comment on the relevant facts on p 9 of their decision. There, they mention that Jørpeland Kraft AS had considered the possibility that a hydroelectric scheme could be undertaken by local property owners. As I have already mentioned, a statement on this was provided to the NVE by Jørpeland Kraft AS themselves. This statement mentioned one possible project that was deemed not to be commercially viable. However, recall that in the same statement another project was also identified -- in the same river, using the same water -- that Jørpeland Kraft AS claimed was such a good project that it could be carried out even after the water from Brokavatn had been diverted.

The statement does not say anything about what the property owners would stand to loose when the water from Brokavatn disappeared, and the Supreme Court also remains silent on this. Nor do they mention that the statement was never handed over to the applicants, and that the details of the calculations were never handed over to, or considered by, the NVE. In fact, the full report first appeared during the hearing at Gulating lagmannsrett. But this fact was not considered relevant by the Supreme Court.

Moreover, the Supreme Court remained silent on the fact that the conclusion concerning efficiency of exploitation contradicts both the NVE's own assessment, the statement made by Hjelmeland Kommune, and also all subsequent assessments made both on behalf of the applicants and on behalf of Jørpeland Kraft AS. I mention that all of the above were presented to all national courts, including the Supreme Court.

As to the legal questions raised by the case, the Supreme Court makes a more detailed argument than the regional court, culminating in the conclusion that established practices still provide adequate protection. Interestingly, the Supreme Court base their arguments in this regard on the premise that the case does \emph{not} involve expropriation of waterfalls. A similar sentiment was also expressed by Gulating lagmannsrett. However, the true force of this point of view did not become apparent until the case reached the Supreme Court.

The Court first concludes that a legal basis for the concession to transfer the water is to be found in the \cite[16]{wra17}. Moreover, they conclude that while this provision alone does not provide a right to expropriate the waterfall, it does give the applicant a right to divert the water away from it. While the Supreme Court notes that this amounts to an interference in property rights, they take it as an argument in favour of regarding the rules in the \cite{wra17} as the primary source of guidance also in relation to procedural rules. 

Hence, the provisions in the \cite{ea59} are regarded as relatively unimportant compared to the rules of the \cite{wra17} and established practice with respect to the provisions in this Act. Moreover, the main reason they the Court gives for this is that the diversion of water is \emph{not} to be considered as an expropriation of a waterfall. As I have mentioned earlier, there is no rule in the \cite{wra17} which states that the authorities are required to consider specifically the question of how the regulation affects the interests of property owners. Such a rule is only found in \cite[2]{ea59}. But according to the Supreme Court, this rule does not apply at all in cases where water is being diverted away from a river.\footnote{It follows, by implication, that the Court regards this rule as ``in conflict with'' or ``unsuitable'' in such cases, c.f., \cite[31]{ea59}.} This is so, according to the Supreme Court, because transferral of water is not regarded as a case of expropriation of a right to the waterfall, but merely an expropriation of a right to deprive the waterfall of water.

This is significant in two ways. First, it is important with respect to the legal status of owners who are affected by projects involving transferral of water. In Norwegian law after {\it Jørpeland}, it seems that established practice with respect to the assessment of such cases, focusing on environmental aspects and the positions taken by various interest groups, is beyond reproach already because such cases do not involve expropriation of waterfalls. 

However, the Norwegian water authorities themselves indicate that they follow similar practices for all large-scale hydropower cases, by relying on an expropriation presumption when the automatic right to expropriate does not apply directly.\footnote{See \cite{flatby08}.} Hence, it remains to be seen if cases involving explicit expropriation of the waterfalls themselves will be treated differently by the courts. Is the conclusion regarding the admissibility of current administrative practices supposed to apply only to those cases when water is subject to diversion? If it is, then it leads to the peculiar situation that the level of protection for owners depend solely on the way in which the developer propose to gain control over the water.

The difference appears arbitrary from the point of view of owners. But of course, it will soon cease to be arbitrary for developers, who must be expected to favour gutter projects, collecting water from many small rivers and diverting it, since this mode of exploitation makes it easier to acquire necessary rights. On the other hand, if the Supreme Court is to be understood as saying that traditional practices are adequate in general, the consequences of the decision seem fairly dramatic for local owners. It appears that it is not possible, in cases involving expropriation of waterfalls, to solicit any kind of judicial review, not even in circumstances when the factual basis of the decision is manifestly erroneous, and local owners are not informed of assessments that the expropriating party makes regarding their interests.

To illustrate that a lack of consultation is a general problem, and not confined to the particular case of {\it Jørpeland}, I will conclude by offering a quote from Harald Solli, director of the Section for Concessions at the Ministry of Petroleum and Energy, who submitted written evidence to the Supreme Court regarding the practices followed in cases involving expropriation of waterfalls. Below, I give one of several exchanges that seem to indicate that under current practices, local owners are left in a rather precarious position.

\begin{quote}
Q: In cases such as this, should owners affected by a loss of small scale hydro-power potential be kept informed about the factual basis on which the authorities plan to base their decision? I am thinking especially about those cases in which the authorities make an assessment regarding the potential for small scale hydropower on affected properties. \\
A: Affected owners must look after their own interests. The assessments made by the NVE in their report is a public document, and it can be accessed through the homepage of the NVE.
\end{quote}

By their reasoning in \emph{Jørpeland}, it appears that the Supreme Court gave this dismissive attitude towards local owners a stamp of approval. In light of this, I believe the study of the law in a socio-legal setting becomes all the more relevant. For while this attitude might be a reflection of correct national law, as decided in the final instance by the Supreme Court, it seems pertinent to ask if it is \emph{reasonable} law. Also, it seems that one must ask if a case can not be made with respect to human rights, by arguing that the protection awarded is insufficient in this regard. This point, while it was raised by the original owners in \emph{Jørpeland}, did not receive any separate treatment in the Supreme Court. In the following section, I briefly describe some further questions raised by the case.

\subsubsection{Assessment}

Following \emph{Jørpeland}, I conclude that the liberalisation of the energy sector does not imply that original owners are entitled to increased protection or participation during decision-making processes regarding the use of waterfalls. At least, this is the view held by the Norwegian judiciary. Of course,  one should not overlook the possibility that the water authorities themselves will eventually adopt new practices regarding the assessment of such cases. So far, however, it seems that they stick quite closely to the established routine.

Hence, it seems reasonable to ask about the sustainability of these practices. In fact, I believe the case of \emph{Jørpeland} itself illustrates why the current system is inadequate, and how it can lead to decisions that appear ill-founded and leave the affected communities marginalised. The likelihood of \emph{factual mistakes}, in particular, might well increase greatly when the involvement of the local population in the decision-making process is not ensured.

More importantly, it seems that when decisions are made following a traditional process, it can often be hard or impossible to see any legitimate reason why the project proposed by the developer would be a better form of exploitation than allowing the local owners to carry out their own projects. In the case of \emph{Jørpeland}, it seemed that small-scale hydro-power would be a better way of harnessing the water in question, even in the sense that it would be more efficient, and would provide the public with more electricity at a lower cost. More generally, unless the issue of alternative exploitation in small scale hydropower is considered during the assessment stage, one risks making decisions that are not in the public interest at all. 

Additionally, this can send out the signal that expropriation of owners' rights is undertaken solely in order to benefit the commercial interests of the energy company applying for a development license. At this point, it seems appropriate to recall the concerns expressed by US Justice O'Connor in the case of {\it Kelo}, regarding the disproportionate influence of powerful commercial actors in takings cases.\footnote{As discussed in Chapter 1.} In relation to these concerns, a major point of contention is  whether or not Justice O'Connor's grim predictions about the fallout of the {\it Kelo} decision did indeed reflect a realistic analysis. 

Surely, anyone who agrees with Justice O'Connor that the powerful will usurp the power of eminent domain to the detriment of the poor, would also agree with her conclusion that it is perverse. However, whether her pessimism is warranted by empirical fact seems less clear. In this context, I believe the case of Norwegian waterfalls serves an important broader purpose, since it sheds light on the hypothesis that a loose interpretation of the public interest requirement will indeed lead to a transfer of property from those with fewer resources to those with more.

I believe the \emph{Jørpeland} case illustrates that her concern must be taken seriously. More generally, the Norwegian case of hydropower seems to show that we need to be clear about the fact that property has a social and political function that goes beyond the financial interests of individuals. For the Norwegian case at least, it seems particularly relevant to ask if local people, by virtue of their right to property and their original attachment to the land, have a legitimate expectation \emph{both} that their commercial interests should be protected, \emph{and} that they should be granted a say in decision-making processes. 

Protection of individual financial interests does not necessarily imply social protection, and the right to participate might be both more significant, and harder won, than the right to be compensated according to whatever the courts regard as the market value of the property in question.

Another perspective, which I will pursue in more detail in the next chapter, is the question of how property rights relate to the overreaching goal of sustainable development of natural resources. Rather than seeing property rights as a means towards securing sustainable development, it is presently very common, particularly in a social-democratic context, to see it as an impediment. This, indeed, has shaped much of the Norwegian discourse regarding environmental law and policy, including the law relating to waterfalls.\footnote{For example, such a sceptical view of property rights appear to provide an overarching perspective on the law of sustainability in \cite{backer12} (a widely used textbook on environmental law in Norway, by one of the most influential jurists over the past 25 years).}

Moreover, a typical justification given for interference in property is that an equitable and responsible management of natural resources requires it. It seems to me, however, that an egalitarian system of private ownership of resources -- as we find in Norway for the case of waterfalls -- could itself serve as a sustainable basis for management of these resources. In particular, I believe that  private property rights are one of the most robust ways in which local communities can be given a degree of self-determination concerning how to manage local resources. 

This is in itself considered desirable from the point of view of sustainability, but often, it appears that the favoured approach to securing it is through administrative arrangements that is supposed to empower local people. The voice that locals get, under such an arrangement, is heard at the discretion of the administrative branch. Hence, it is easily drowned in the context of large-scale commercial development, particularly when a narrative is established whereby this development is carried out for the ``common good''. I believe the case of Norwegian hydropower illustrates this point, and also shows that there is every reason to remain sceptical when commercial interests claim that they embody the public interest.

If property becomes a privilege for the few, rather than an obligation for the many, there is reason to expect that the level of vigilance will be diminished in this regard. In this way, takings for profit serve a double purpose in that they establish a culture of property which threatens to transform it from an egalitarian institution focused on virtue, to an elitist institution focused on entitlement.
The repercussions of this, I believe, extend well beyond the local communities that are adversely affected in the first instance. For example, it seems to me that an egalitarian property regime is one of the paramount guarantees we have that the state will be able to effectively and rationally exercise its regulatory powers, without bias. 

In a system where property no longer acts as an equaliser, but rather as a marker of political and commercial power, one must expect that the government itself will be more easily intimidated by commercial interests. This becomes particularly likely if such interests permeate the political system itself. The case of hydropower in Norway illustrates how this can happen, as a result of naive policies of state-ownership, aiming to protect the public interest, but effectively serving to render the public will subservient to the imperatives of the market. 

Moreover, a capitalist elite which commands significant regulatory power may not take lightly to what they perceive as undue political interference in their business practices. A company which is partly owned by the state, operates the electricity grid on behalf of the public, and is accustomed to expropriating property without meeting with much resistance, is likely to act with great confidence and boldness also when it faces political opposition. Much more so, one would presume, than a group of peasants, the original owners of waterfalls.

I think the case of \emph{Jørpeland} suggests that these perspectives are all relevant when considering takings for profits and their consequences. In the next chapter, I will turn to the question of how to ensure commercial development via compulsion by a different route, {\it without} incurring the negative effects associated with expropriation in such circumstances. Here I believe the fundamental challenge is one of setting up a decision-making structure that can ensure a balance between the various stakeholders. Importantly, the decision-making structure itself should recognise that local owners are those who have the most to loose, and gain, from undertakings involving their property. 

Indeed, I believe that the appropriate form of democratic decision-making involving property interests needs to reflect both the obligations and rights associated with those interests. The owners must not be conceptualised as an impediment to development, but as a potential driving force. Moreover, they cannot be conceived of merely as interested bystanders, but must be recognised as {\it the} primary actor in any collective action that seeks to transform their property so that it may serve a more valuable social function. No doubt, to achieve such a transformation may require compulsion. But it need not require outright expropriation, as I will attempt to demonstrate in the next chapter. The overriding concern is to develop strategies whereby the state can impose new property uses without undermining existing distributions of property rights and obligations among its citizens.

\section{Conclusion}\label{conc}

In this Chapter, I have argued that the law relating to expropriation of waterfalls in Norway is based on a tradition that sees owners as profit-maximising and the state as welfare-seeking. Hence, the question of striking a balance between private and public interests is approached under the presumption that private property rights embody mainly private values, while public values are pursued through regulation that ensures public ownership and control. I observed how this perspective shaped the law of expropriation of waterfalls, so that expropriation could only take place for narrowly defined public purposes and only to the benefit of public bodies.

I noted, however, how the increasing centralisation of the energy sector and the increasing scale of projects following the Second World War led to increased worry about the legitimacy of interference in property and the natural environment. I concluded that the ensuing conflicts, while severe, largely failed to make an impact on the law relating to hydropower, as long as the sector was organised as a public service. In particular, the perceived level of political control over the sector meant that courts shunned away from adopting a strict view on legitimacy. 

This did not mainly apply to the question of the authority to expropriate, which was hardly raised at all in the period between the reversion controversy of the early 20th century and the market-reform of the early 1990s. It applied mainly to procedural rules. Here the Supreme Court adopted a stance whereby these rules were themselves considered largely ``discretionary'' in nature. Hence, it would fall under the authority of the executive to determine their scope and application in concrete cases.

I went on to note that this perspective has been maintained by the courts and the executive even after the market-reform, when the sector lost its characteristics as a public service. I argued that today, expropriation of waterfalls for hydropower development must be regarded as takings for profit, typical economic development takings. I discussed how the law came to be changed on this point, with a dramatically widened expropriation authority introduced in conjunction with the \cite{wra00}.

I concluded with a description of the fallout from this, as expressed concretely in the case of {\it Jørpeland}. This case served to illustrate that administrative practices developed and sanctioned during the monopoly days are now applied in a context of competing commercial interests. As a result,  expropriation has become an important tool that the powerful market players can use to gain the upper hand in competition with locally based companies or smaller companies that rely on cooperation with owners. I noted that the law as it stands is unprepared for dealing with this dynamic. 

Still, in the case of {\it Jørpeland}, the Supreme Court explicitly denied that established practices were in need of revision. Moreover, it refused to reconsider the established interpretation of the scope of procedural rules in hydropower cases, rejecting arguments to the effect that these must now be understood to provide protection for waterfall owners that matches the protection offered to other affected parties.

\noo{ In the next Chapter, I will consider an aspect of the law were the Supreme Court {\it has} taken the view that a revision of established practices is in order, namely in relation to the question of compensation. I note, however, that the Court's emphasis on the compensation issue serves to reinforces the idea that private property rights pertain mainly to financial entitlements. As I have already argued, this perspective hardly does justice to the role of private ownership of waterfalls in Norway. }

In the next chapter, which is the last chapter of the thesis, I consider land consolidation as an alternative to expropriation. I believe it has great potential. However, as demonstrated in the present chapter, Norwegian courts do not seem to recognise the shortcomings of the current system. Until they do, or are directed to do so by political bodies or international tribunals, it is unlikely that expropriation law will evolve much from its current fixation on the compensation issue and its unshaken belief in public-private partnerships as arbiters of the common good.
