\chapter{Conclusion}\label{chap:7}

\begin{quote} \small
Proudhon got it all wrong. Property is not theft -- it is fraud.\footnote{\cite[252]{gray91}.}
\end{quote}

\begin{quote} \small
That's what makes it ours -- being born on it, working on it, dying on it. That makes ownership, not a paper with numbers on it.\footnote{\cite[33]{steinbeck39}.}
\end{quote}
\noo{
\begin{quote} \small
Gode bønder nå til motstand reises. Enhver som må gi fra seg farsarvi til kongens grever mener det er det rene ran. \\
Good farmers prepare to resist. Those who must give up their father's inheritance to the lords of the king consider it plain robbery.\footnote{From the so-called {\it Bersogliviser}, an old Norse poem addressed to King Magnus (c. 1024 – 25 October 1047), who was accused of treating his subjects badly, including taking their properties, see \cite[259]{titlestad12}. The King listened to the farmers' protest and was thereafter referred to as Magnus ``the Good''.}
\end{quote}
}
%There is reason to think that fraudulent behaviour requires some, but not too much, in the way of intellectual refinement. Arguably, therefore, the case can be made that 

%The concept of property has received its share of criticism, but it remains omnipresent.

%
\noo{ t. Further to this, legitimacy-enhancing alternative to expropriation were considered. Specifically, the thesis argued that ideas taken from the research on common pool resources could inspire novel solutions for local self-governance that could obviate the need for using eminent domain to ensure economic development. 



%A concrete proposal along this line, due to Heller and Hills, was considered in some depth. 

%An important tension was identified within this proposal, between the ideal of self-governance and the danger of abuse at the local level. This, in turn, reinforced one of the key design principles of common pool resource management, namely that institutional arrangements need to be sensitive to the local context. This set the stage for the next part of the thesis, where the issue of economic development takings was looked at concretely, from the point of view of waterfall expropriation foor hydropower development in Norway. 

%Hence, to formulate a one-size-fits-all institution to replace eminent domain for economic development is unlikely to work. At least, such a solution would have to be far more flexible than the proposal made by Heller and Hills, which in the end offers owners very little in the way of participation in decision-making regarding economic development on their properties.


In such cases, it might well be that the balancing of different reasons for and against the taking has taken place prior to the decision to interfere with property. The plans for development themselves may well precede any specific property-oriented implementation steps, such as the use of eminent domain. It might even be that democratically accountable bodies responsible for land use planning have already concluded that some local community interests must give way to other interests.

In these cases, it might be tempting to argue that a narrow takings narrative is appropriate because it pertains only to the final implementation step, which is the only one that involves property rights. But this argument, I believe, rests on a flawed perception of what property is, and should be, in a democratic society. Invariably, property has to do with decision-making and power. If the decision-making process does not grant significant self-determination rights to affected property owners, a taking is already in progress. It might be justified, but it is still a taking. 

More worryingly, it is clear that this kind of taking carries with it a great potential for differential treatment, discrimination, and corruption. The traditional takings narrative does a good job of setting up a framework that makes it difficult to simply pay higher compensation to certain kinds of people, without offering any justification. But with respect to the aspects of taking not recognised, e.g., pertaining to what role the owner has during the planning stages, differences in treatment will not even be notices. But if property is owned by the right sorts of people, then invariably it {\it will} come with considerable decision-making power. 
}


\noo{
This has been a thesis in two parts that has addressed economic development takings from two distinct angles. In the first part, a theoretical discussion was provided, which started from the notion of property itself and gradually made its way towards the question of legitimacy of takings by exploring a broader meaning of property than that typically embraced by the law. The aim was to argue {\it why} property should be protected, while also providing a template for recognising and discussing special issues that arise for economic development takings. 

Following up on this, the thesis explored different approaches to legitimacy, culminating in a concrete proposal for a legitimacy test, inspired by the work of Kevin Gray. Furthermore, the thesis presented a template for formulating alternatives to outright expropriation, inspired by the work of Elinor Ostrom and others on local self-governance of common pool resources. The aim here was to suggest possible ways to restore legitimacy in cases when an either-or approach to legitimacy of property interference is not appropriate. Better than that it should be interference bottom-up than interference top-down, provided adequate safeguards are put in place to protect local minorities from abuse.

The discussion remained quite abstract in the first part of the thesis. By contrast, the second part of the thesis approached the question from the opposite angle, through an in-depth case study of takings for hydropower development in Norway. The thesis first presented and discussed the social, economic, and legal context of such takings, before proceeding to study practices and rules relating to expropriation of waterfalls in greater depth. This also provided an opportunity to apply the legitimacy test proposed in the first part of the thesis. The conclusion was that current expropriation practices in Norway fail in this regard, as the property rights of local people are very weakly protected in situations when large commercial companies wish to undertake hydropwoer development.

However, the thesis went on to observe that the land consolidation procedure represents a possible alternative to expropriation, one that is already being used extensively in Norway to facilitate owner-led hydropower development. Here the emphasis is on organising joint use of property, possible involving compulsion whereby owners must partake in development against their will. The procedure is judicial in nature, moreover, so also provides safeguards against abuse by local elites. Moreover, it is primarily a service to the owners and their properties, but is required to actively promote solutions that are in the public interest. After recent changes in the law, the scope of obligations that can be imposed on owners for the common good is also likely to increase, making the land consolidation alternative appear like a realistic option even in large-scale development situations that will necessarily also involve non-local actors. The thesis argued that land consolidation is a good example of the kind of institution that can function as an alternative to expropriation in hard cases. It has already proven itself in a setting of egalitarian property, in communities where property rights are held by local people. The possibility of employing it in situations where this is not the case remains more uncertain, but is an interesting prospect. Arguably, this would require including non-owners in the process as well, on the basis of connections to property that might not otherwise be recognised as property interests in the law. Specifically, it seems that one might want to rely on broader social function ideas of property in relation to consolidation, even if a more traditional account is maintained in other areas of the law. In relation to financial law or tax law, one might well wish to entertain an artificially narrow concept of property for efficiency reasons, even if a much broader concept is required in the context of compelled economic development. Moreover, the land consolidation procedure appears attractive because it fills a gap between planning and property, a gap that otherwise appears susceptible to infiltration by powerful commercial interests.

%The thesis made the case that these issues are important enough to suggest that economic development takings should be approached as a special category, also in the law.

%From this, the theoretical discussion continued by an exploration of different ways to approach the legitimacy question for such takings. 
}

%In this final section of my thesis, I would like to take a step back to briefly follow two broader threads that I believe run through my thesis. 
This has been a thesis in two parts, each of which have approached the issue of economic development takings. The first part took a theoretical approach, starting from the notion of property itself, to answer the question of {\it why} it should be protected. This, in turn, gave rise to a framework for assessing the legitimacy of economic development takings, and for formulating alternatives to it that could obviate the need for dispossessing current owners.

The second part of the thesis approached the issue of legitimacy concretely, by giving a case study of takings of waterfalls for hydropower development in Norway. The political, social and economic context was also analysed, leading to an application of the Gray test formulated in the first part of the thesis. Moreover, the case study considered the possibility of alternatives to expropriation, by assessing the Norwegian institution of land consolidation, which is now used extensively by local owners who wish to undertake hydropower development themselves.

% The current approach to takings for waterfalls in Norway were found wanting, with the current takings practices appearing to fail several, if not all, the points set out by the Gray test. However, the final chapter of the thesis studied a possible alternative to expropriation, which paradoxically is also actively used in the context of hydropower development in Norway. This framework, however, is so far used only when some of the local owners themselves wish to carry out development, and need to sort out their internal disagreements, possible even by compelling unwilling neighbours to join them in their endeavours. 

%The thesis argued that this alternative, although not necessarily applicable in other contexts, provides an interest

%More generally, and especially in its proposal for a possible solution, I hope the thesis has made a valuable contribution to the study of economic development takings. 

In the following, I offer a brief summary of the main points discussed in each chapter. While doing so, I hope to shed further light on a broader thread that runs through the work done in this thesis, pertaining to the importance of social justice considerations in takings law and the function of private property as an anchor for local self-governance and sustainable resource management in democratic societies.

%To conclude the thesis, I will take a step back to consider two broader threads that I believe run through my work, pertaining to the nature of property and how to ensure that it functions as a force for good in democratic societies.

%The first concerns the many senses of taking that have been brought into focus throughout the analysis, while the second concerns ways in which the law can help to give back some of the legitimacy that is typically lost when eminent domain is used to facilitate economic development.

%\section{Summary of main points}

\subsubsection*{Property theory}

%To motivate the theoretical work done in chapter 2, the thesis used the example of Donald Trump coming to Scotland to build a golf resort on the Aberdeen coast, on a site of special scientific interest. This looked like it was going to result in the compulsory acquisition of land, as the local authorities were asked to considered issuing compulsory purchase orders in support of Trump's project. However, in the end, the plans did {\it not} require takings, as Trump was able to make creative use of property rights he acquired voluntarily, against the complaints of his neighbours.

%Importantly, this turn of events did not make the example less relevant to the thesis. Rather, it served to highlight that the question of economic development takings cannot be looked at in isolation from the surrounding social and political context. Clearly, the grievances of the local people in Balmedie who were marginalised by Trump did not turn on whether or nor he formally condemned property, but rather on the manner in which the social functions of property in the local area conclusively changed in his favour, at the expense of others. Trump's actions, then, arguably still amounted to takings,  even if not in a formal legal sense of the word. Moreover, the example of Trump coming to Scotland allowed me to emphasise the importance of context when assessing both the nature of property, the many ways of taking, and the meaning of protecting owners against predation.

%The protection sought by those who opposed Trump's golf course did not target their entitlements as individuals. Rather, it targeted the community, as the owners felt it would be detrimental to the community, and to their lives, if Trump was allowed to redefine the social functions of local property. After Trump decided not to pursue compulsory purchase, protecting the property of these members of the community became a question of {\it restricting} the degree of dominion that Trump could exercise over his own property. Hence, under a conventional and overly simplistic way of looking at these matters, protecting property became tantamount to restricting its use, a seeming paradox.

To arrive at a theoretical framework for discussing economic development takings, chapter 2 considered various theories of property. The chapter noted that there are differences between civil law and common law theorising about ownership, but concluded that these differences are not particularly relevant to the questions studied in this thesis. In particular, the chapter observed that neither the bundle theory, dominant in the common law world, nor the dominion theory, taught to many civil law jurists, helps to clarify the distinguishing features of economic development takings. Specifically, traditional thinking both in civil and common law jurisdictions has a tendency to abstract away from the sensitive social and political context of such takings; there is a clear tendency, especially in takings cases, that private property is approached as a set of individual entitlements rather than an interconnected web of social functions. This, in turn, means that the broader societal effects of takings are often not considered when addressing the legitimacy of property interference against general protection principles found in constitutional and human rights law.

The thesis set out to arrive at a theoretical foundation for thinking about property that would support a more comprehensive approach to takings in the context of economic development. The chapter focused especially on property theories that emphasise that property plays a crucial role in many social and political relations within a society. Such social function theories, it was argued, provide us with crucial {\it descriptive} insights into the workings of property and its role in the legal order. In this regard, the thesis advanced a position different to that adopted by many previous scholars in the social function tradition, by arguing that we should actively try to decouple descriptive insights from normative claims about property. It was argued that it we succeed in doing this, the social function theory could serve as  common ground for further value-driven debates that cross ideological divisions.

Following up on this argument, the chapter further clarified the normative starting point adopted in this thesis, by expressing support for the human flourishing theory proposed by Alexander and Pe\~{n}alver. This theory is based on the premise that property rights {\it should} be integrated in the legal order in such a way that they enable -- possibly even compel -- individuals and their communities to participate in social and political processes. Specifically, the chapter argued that property and its associated social functions should be approached as a crucial anchor for democracy, since private property can potentially provide a powerful anchor for participatory decision-making on the basis of mutual obligation and respect, targeting owners and non-owners alike. Specifically, the human flourishing theory rightly emphasises the {\it duties} associated with private property, especially those duties that are directed at non-owners and which therefore necessitate inclusive institutions for collective decision-making in cases when conflicting interests are at stake.

Importantly, the human flourishing theory emphasises how such institutions can arise from the structure of property itself, rather than as a purely external framework imposed by the authority of the state. For this reason, the human flourishing theory contains a further important insight, concerning the scope of the state's obligation to protect property. In particular, the human flourishing theory asks us to acknowledge that protecting property also implies a commitment to protecting the right to some degree of self-governance for local communities, where the stake an individual has in decision-making processes are reflected in their opportunity to influence those processes. Being an owner, with all the rights and responsibilities this entails, then places an individual right at the center stage of decisions pertaining to the use of property, including in situations involving plans for large-scale economic development. This highlighted the relevance of property theory to the question of legitimacy in economic development takings.

Following up on the theoretical argument, chapter 2 sketched out what a social function perspective could imply in practice by considering {\it Kelo}, the paradigmatic case of a taking for economic development in the US. It was observed that the disagreement within the Supreme Court seemed to turn, in part, on the willingness of the justices to adopt a social function perspective on the case. Justice O'Connor, who led the dissenting minority, departed from a formalistic, entitlements-based approach, when she argued that takings for economic development should be prohibited because such takings would systematically bestow benefits on powerful commercial interests at the expense of owners from weaker social groups. In addition, both Justice O'Connor and Justice Kennedy, who voted with the majority to uphold the taking, emphasised the danger that the commercial incentives associated with using eminent domain to facilitate economic development could distort decision-making processes, creating a democratic deficit within government institutions endowed with eminent domain powers. These two themes, pertaining to the social fairness and procedural legitimacy of takings for economic development, has remained in focus throughout this thesis.

\subsubsection*{Testing for legitimacy and looking for alternatives to eminent domain}

In chapter 3, I gave a positive-law presentation of the legitimacy question that arises for economic development takings.
I considered existing approaches from the UK, the US, and at the ECtHR. As explained in the introduction to the thesis, this choice led me to consider a set of jurisdictions that adopt distinctly different approaches to the legitimacy question, yet remain easily comparable both to each other and to Norway, which was investigated in more depth in the subsequent case study. 

In the UK, the debate on legitimacy is traditionally structured around procedural rules, while in the US the traditional starting point, at least in before the higher courts, is a substantive assessment of the merits of takings. In the end, the chapter argued in favour of an approach that combines procedural and substantive standards, with the intention being that the latter should be used also to assess the fairness and democratic merit of the decision-making procedure, not merely the outcome. Furthermore, it was argued that such an institutional approach to fairness has started developing at the ECtHR in response to the newly introduced framework for pilot judgements applied in cases that might indicate systemic problems at the state level.

The chapter went on to propose a concrete heuristic for assessing the legitimacy of economic development takings against any standard that implies a commitment on the state's part to ensure a ``fair balance'' between the opposing interests involved. The chapter referred to this heuristic as the Gray test, since it is strongly influenced by previous work done by Kevin Gray, a leading UK scholar who specialises in land law. Indeed, the proposed heuristic includes a list of legitimacy indicators provided by Gray in his work on the legitimacy of takings. In addition to these original indicators, the thesis went on to add three additional points, based on the work done in this thesis. The resulting heuristic should be able to identify many, if not most, cases of eminent domain abuse, especially those that occur in the context of economic development. Furthermore, the Gray test incorporates several aspects of the social function theory of property, looking at the legitimacy issue from a broad vantage point of how the interference in property will affect communities and local democracies, not merely individual owners. This makes the Gray test stand out in the literature on takings so far, especially in the UK and the US, where individualistic perceptions of what property is have tended to dominate.

Following up on the discussion of legitimacy testing, chapter 3 considered the question of how to increase legitimacy without giving up on the idea that the collective should be entitled to push adamantly for economic development on private land. It was argued that the work done by Elinor Ostrom and others on common pool resources provides a suitable starting point for making institutional proposals to achieve a better framework for making decisions about economic development on privately owned land. Specifically, the chapter briefly presented Ostrom's design principles for local self-governance of natural resources, which can be used as a starting point also for designing procedures to replace eminent domain for economic development.

The chapter went on to consider a proposal that has already been made along these lines, namely the idea of Land Assembly Districts, due to Heller and Hills. This proposal was analysed in depth, and the chapter pointed out some potential problems with it, including a tension between the overarching goal of self-governance and the perceived danger of tyranny by local elites and abuse of local minorities. In the end, the chapter concluded that a single institutional framework is unlikely to work in all settings; institutions for self-governance need to be attuned to local conditions, so they cannot be made too general or justified in overly theoretical terms. Indeed, sensitivity to local conditions is one of the key design principles proposed by Ostrom, and should influence also the proposals we come up with for institutional reform in the law of takings.

This observation marked the end of the first part of the thesis. In the second part, the thesis considered the case of Norwegian hydropower. This led to an analysis of legitimacy of takings for this purpose along the lines of the Gray test, as well as a case study of the land consolidation courts and their power to set up local institutions for self-governance that can obviate the need for eminent domain. In this way, the second part went on to apply key aspects of the theory developed in the first part, while exploring further the idea that social functions run as a common thread through individual property rights.

\subsubsection*{Norwegian waterfalls and their social functions}

Chapter 4 introduced the case study and provided background information that placed it in a broader context with respect to Norwegian law. The legal and regulatory framework surrounding hydropower development was discussed, and its history was traced back to pre-industrial times. The chapter emphasised that local rights to hydropower have a long tradition in Norway, with communities of smallholders typically holding the rights to harness power from local rivers in common, as incidents of their shared ownership of the surrounding outfields. This arrangement dates from a time when grist mills and saw mills were important to many local communities, whose proprietary rights provided them with an opportunity to benefit from local resources.\footnote{As noted in chapter 4, Norwegian tenant farmers also enjoyed such rights before they started buying back their farms in the 17th and 18th century. In particular, tenants would typically have quite extensive rights to natural resources found in the outfields. This highlights that private ownership of land was always imbued with egalitarian social functions in Norway, not simply an encapsulation of individual entitlements for the privileged classes.} 

After the advent of the industrial age, and particularly following the Second World War, the state took the view that hydropower was a public good that should be exploited for industrial development in the interest of the general public. This resulted in a tension between local self-governance, rooted in private property, and central management, rooted in  the authority of the state, that now permeates the law of hydropower. This tension became particularly severe after the liberalisation of the electricity sector in the early 1990s. This reform reorganised hydropower development as a commercial pursuit, meaning that when the state uses its regulatory power, it often bestows benefits on commercial companies at the expense of local resource owners. At the same time, local owners themselves were empowered by the liberalisation reform, since they could now themselves engage in commercial hydropower development. This was made possible by the fact that a market for electricity was set up, founded on the idea that all actors should have access to the electricity grid on non-discriminatory terms.

Chapter 4 discussed the resulting system in some depth, addressing also the question of whether or not the liberal market functions as intended. It was argued that the energy reform has marginalised the municipality governments, who used to play an important role because they were in charge of public utilities for the supply of electricity in their own local district. The result, it was argued, has been a dramatic concentration of power in the hands of commercial companies partly owned by the state, a centralisation effect that threatens to undermine the market-stimulating intentions behind the reform. Specifically, the chapter provided a critical assessment of the extent to which the regulatory framework is able to accommodate new actors and facilitate competition on non-discriminatory terms. Special attention was directed at the position of waterfall owners and the companies that specialise in cooperating with them. It was argued that owners and small-scale development companies currently suffer under a regime that tolerates, and sometimes encourages, discrimination against less dominant market actors.

The chapter went on to present how smaller market actors now tend to organise themselves. First, the chapter presented an early organisational model that emphasised respect for property rights, local communities, and the environment. According to this model, owners and local communities would typically retain controlling stakes in development projects, with external partners providing capital and technical expertise either as a paid service or in return for a minority stake in the enterprise. Many owner-led hydropower plants have since been built according to this model, demonstrating its commercial viability. 

However, later developments have led to an increased concentration of power and ownership also in the small-scale segment of the electricity production sector. Specifically, external partners now typically demand controlling stakes in development projects and do not concede to organisational provisions meant to protect local communities and environmental interests. Effectively, the owners and their communities are increasingly asked to remain on the sideline. In many cases, local people have agreed to this in return for a promise of higher compensation in the future, calculated as a percentage of income from the generation of electricity. However, it has turned out that such promises have often been made with little realism. Indeed, as shown in chapter 4, there have been some high-profile cases of speculation in this market, where upstart energy companies have entered into a large number of agreements with local owners without carrying out much actual development. %Recently, large portfolios of agreements have been sold on to third parties, including multi-national corporations.  who have then been known to re-negotiate the terms with local owners under the threat that no development will take place unless the agreements are made more favourable to the external partner.

In general, chapter 4 argued that recent developments in the small-scale sector marks a sharp departure from initial visions of this sector, visions that emphasised values such as local self-governance and environmental sustainability. Instead, the values and practices of leading small-scale actors have become increasingly similar to those that dominate among more established market actors. The chapter went on to argue that this might be a contributing reason why small-scale development now appears to be falling out of political favour. Today, critical voices claim that large-scale development is better, not only because it is more commercially optimal, but also because it is more environmentally friendly. The specific argument provided for this claim is that small-scale projects affect a greater number of square meters per energy unit they produce. As noted in chapter 4, this is no doubt true, since many small-scale plants will typically be required to match the energy output of a single large-scale plant. Unfortunately, issues relating to more substantive notions of sustainability, as well as issues relating to property rights, benefit sharing, and local participation in decision-making, appear only at the fringes of the present debate. Hence, policies are now at risk of being formulated on the basis of an incomplete picture of what is at stake. As suggested in chapter 4, this makes the present thesis a timely scholarly contribution that could also inform policy making on hydropower in Norway.

%By offering this diagnosis, chapter 4 foreshadowed many of the issues that was brought into focus in the subsequent chapter. There, the thesis looked specifically at expropriation of waterfalls, by tracking the position of owners under the current regime. It was argued that the law as it stands is based on a perspective that blocks out both the significant commercial interests of the taker, as well as the significant social functions and obligations of the original owners. The main point made was that the issue of expropriation invariably raise questions that seem difficult to address without adopting a broader view, which also takes into account the owners' communities and their role within it.
\subsubsection*{Expropriation as an automatic consequence of a development license}

In chapter 5, the legal framework surrounding expropriation of waterfalls was studied in depth. In addition to presenting the law, the chapter also discussed administrative practices developed by the water authorities. The chapter argued that the present system is based on the presumption that private property embodies private values, while public values are to be pursued through regulation, if necessary also by redistributing or negating property rights. In effect, chapter 5 tracks how this perspective has shaped the law of expropriation of waterfalls. In particular, the chapter made clear that the notion of property presupposed by Norwegian expropriation law is not based on a social function understanding of what property is and why it should be protected against interference. 

At the same time, the traditional view on public values implied a sharp distinction between commercial and public uses of property. Hence, the law of expropriation initially tended to restrict the takings power to narrowly defined purposes that clearly served the common good. This was the case also in the context of water law, where the government did not initial have any authority to expropriate waterfalls for the purpose of developing hydropower. However, as noted in chapter 5, the increasing focus on electricity production as a public service resulted in the introduction of new authorities to expropriate waterfalls for public utilities. Still, expropriation could not take place for commercial purposes or in favour of private companies.

The restrictions placed on the power to expropriate waterfalls gave legitimacy to the legal framework. Still, the increasing centralisation of the energy sector and the increasing scale of projects seen after the Second World War led to increased tension surrounding new hydropower projects. Tensions came to a high-point in the case of {\it Alta}, when the indigenous Sami population from the north of Norway objected to a project that would have detrimental effects on their livelihoods and, as they argued, their very way of life. In collaboration with environmental groups, they launched a legal challenge directed at the development license, resulting in one of the most comprehensive cases ever dealt with by the Supreme Court. In the end, the development interests triumphed, and the regulatory framework surrounding hydropower and expropriation was given a stamp of approval. 

Even though the development license was upheld, the {\it Alta} conflict contributed to increased awareness of indigenous rights in Norway, starting a development that has since resulted in better protection of Sami rights within the legal order. No similar effect was observed with respect to the law of hydropower; the precedent set by {\it Alta} remains leading in disputes over the legitimacy of licensing decisions and expropriation orders. However, as noted in chapter 5, the context of waterfall expropriation has changed dramatically since the liberalisation of the electricity sector. Unlike before, expropriation is now regularly ordered for commercial purposes, to the benefit of limited liability companies.\footnote{For a discussion of how the law was changed to make this kind of expropriation possible, see chapter 5, section ....}

This change of context has failed to make any discernible impact on the administrative practices adopted by the waer authorities. Rather, the traditional approach, which evolved when electricity generation was still organised as a public service, remains in force. As discussed in chapter 5, the key feature of these practices is that they render expropriation a {\it de facto} automatic consequence of obtaining a development license; if an energy company manges to obtain a development license for a large-scale project, the right to expropriate waterfalls is granted to it by default. As shown in chapter 5, this leaves the owners and their local communities in a precarious position in such cases. Essentially, they enjoy very limited protection under administrative law, not substantially more than arbitrary members of the public, and in some cases far less than the members of organised interest groups.

%This did not mainly apply to the question of the authority to expropriate, which was hardly raised at all in the period between the reversion controversy of the early 20th century and the market-reform of the early 1990s. It applied mainly to general procedural rules. Here the Supreme Court adopted a stance whereby these rules were themselves considered largely ``discretionary'' in nature. Hence, it would fall under the authority of the executive to determine their scope and application in concrete cases.

Chapter 5 discussed the practical fallout from this in depth when considering the case of {\it Jørpeland}. This case served to illustrate that administrative practices currently in place serve to make expropriation available as an effective tool for powerful market players that wish to gain the upper hand in competition for resources owned by weaker groups. Hence, the current state of expropriation law in Norway adds weight to Justice O'Connor's prediction in {\it Kelo}, where she predicted permitting economic development takings would give powerful commercial interests the opportunity to take property from weaker members of society.

Chapter 5 also noted that in the case of {\it Jørpeland}, the Supreme Court explicitly denied that established practices were in need of revision. Moreover, the Court refused to reconsider the established interpretation of the scope of procedural rules in hydropower cases. Effectively, a range of general rules of administrative law do not apply in hydropower cases, since the special legislation used to regulate licensing of hydropower take precedence. In a dramatic departure from the situation as it had been before commercial expropriation was permitted, this hydropower-specific legislation is now also understood to cover the expropriation proceedings, which are not considered separate from the licensing proceedings at all.

\subsubsection*{Land consolidation as an institutional framework for local self-governance and sustainable resource management}

In chapter 6, the Norwegian system of land consolidation was presented. It was shown to be a unique institution that combines a property-based approach to land management with a broad authority for the courts to intervene in order to organise collective decision-making and promote sustainable resource management at the local level. The chapter went on to show that the system of use directive has become widely used in recent years to facilitate hydropower projects organised by local owners. In some cases, it is also used to deprive some owners of their holdout power by compelling them to participate in a development project that they oppose.

In consolidation cases, interference in property does not take place because property rights must give way to public interests. Rather, consolidation relies on proof that benefits will outweigh harms at the local level, with respect to each affected property. However, this requirement targets the property as a functional unit, irrespective (in principle) of the specific interests of its current owner. Hence, depending on what functions are regarded as more important, the stated desires of the owner might have to give way to other priorities. At the same time, the compensatory perspective is abandoned; the owners' financial entitlements are subordinated to the interests found to be inherent in their properties. If the property as a functional unit does not suffer a loss, no compensation is payable to the owner as an individual. 

As demonstrated in chapter 6, this perspective diverges from a perspective that sees private property as a way to encapsulate the financial entitlements of individuals. Instead, the perspective on property inherent in the land consolidation model is much closer to that postulated by the social function theorists discussed in chapter 2, especially those that focus on human flourishing as the underlying purpose of private property rights. Indeed, land consolidation relies on a highly functional perspective on property: beneficial resource uses, not individual entitlements, take center stage throughout the process. This might limit the power of the owners to do as they please, but it does not marginalise them. After all, it is hard to deny that one of the primary functions of private property is to bestow rights and obligations on its owners. Moreover, in normal circumstances, it would be safe to assume that when a property benefits, then so does its owner.

For this reason, it also seems that consolidation can be used to address the democratic deficit of economic development takings in an elegant way. Chapter 6 addressed this possibility in depth and argued that the land consolidation courts can be seen as providing an institution authorised to design and implement self-governance arrangements that approach private property in land as a common pool resource. This connected the case study of Norwegian hydropower with the discussion provided at the end of chapter \ref{chap:2}, regarding institutional alternatives to eminent domain for economic development. In particular, it connects the system of land consolidation courts with theories about self-governance and common pool resource management.

Because it combines great flexibility with a well-structured judicial procedure including many safeguards, the land consolidation option appears attractive as a vehicle for sustainable self-governance of local resources found in a local community. Moreover, it provides the public with a means to impose their will on owners and communities, without having to resort to the use of expropriation. While ensuring that legitimacy policy goals can be pursued more effectively, this also raises the worry that consolidation abuse will emerge to replace eminent domain abuse. Chapter 6 argued that in order to address this worry, the land consolidation alternative needs to function as a service to owners and local communities, as a means of helping them to manage their properties in accordance with public interests. Arguably, this requires a clear commitment on part of the state to prevent abuse of consolidation measures by commercial interests and public-private partnerships that seek access to property rights held by weaker parties.

Assuming that such a commitment is made, chapter 6 argued that use directives issued by a land consolidation court can be empowering to local owners, who are presented with a direct sense in which their property rights contributes to collective decision-making with a democracy guided by the rule of law. Specifically, chapter 6 argued that land consolidation can be used to deal with many of the challenges that arise at the intersection between private property, local community, and economic development in the public interest. The Norwegian model might also inspire similar solutions elsewhere, particularly in jurisdictions that are committed to an egalitarian ideal of property ownership.

\noo{ 

\section{On the legitimacy of interference in property, and the social structures it helps sustain}

\section{Substance and procedure -- protecting the social functions of property by reviewing decision-making processes leading to interference} 

\section{Norwegian waterfalls -- properties, assets, and common goods}

\section{Property taken by default}

\section{Property given following participation}

\section{Conclusion}

%\section{Property Lost -- Taking and Excluding}\label{sec:7:1}

%According to some, the law does not like it when things get too academical; after all, the law is not in the business of settling philosophical debates, but rather used to resolve disputes.
%Rather, its task is to deliver effective management of disputes, involving concrete legal persons or governments.
%In fact, many leading legal theorists adopt a closely related position, when they argue that the law 

According to many legal scholars, the law is an instrument of order, grounded in social facts, not an arbiter of justice, grounded in moral theory.\footnote{This is a terse formulation of a position typically associated with legal positivism; it can be elaborated and restated in a multitude of ways, giving rise to many theoretical variations. The merits of the positivist account of the relationship between morality and law is a contested issue in the so-called Hart-Dworkin debate in legal philosophy, see \cite{hart12,dworkin86,shapiro07}.} Building on this, a pragmatist might be tempted to think that the law should be sceptical of embracing the complexities of equity when goals such as efficiency,  certainty, and control appear to be better served by simplification.\footnote{For a theory that emphasises pragmatism and chides moral theory, see \cite[109-110]{posner99} (``Holmes warned long ago of the pitfalls of misunderstanding law by taking its moral vocabulary too seriously. A big part of legal education consists of showing students how to skirt those pitfalls.'' (citations omitted)).} This way of thinking appears to have played a significant role in shaping the traditional approach to the legitimacy issue in the law of takings.

% would be in the interest of efficiency.
%According to positivist thought, the law is rightly sceptical of allowing things to get too theoretical; after all, the law is not in the business of settling ethical issues. Indeed, many believe that its main responsibility is to deliver effective management of disputes, involving concrete legal persons or governments.

It is clear that by focusing on individual owners and their losses, the law makes things easier for those called to administer it. Moreover, by assuming that all losses can be quantified in financial terms, the law's standard can be at least partly automated. The potentially broad question of legitimacy has been reduced to the question of when and how to award compensation, for which the ostensibly neutral idea of ``fair market value'' appears to provide a practical starting point, well suited for maintaining order in a capitalist society.\footnote{See, e.g., \cite[510-511]{acres79} (``In giving content to the just compensation requirement of the Fifth Amendment, this Court has sought to put the owner of condemned property `in as good a position pecuniarily as if his property had not been taken.' However, this principle of indemnity has not been given its full and literal force. Because of serious practical difficulties in assessing the worth an individual places on particular property at a given time, we have recognized the need for a relatively objective working rule. The Court therefore has employed the concept of fair market value to determine the condemnee's loss. Under this standard, the owner is entitled to receive `what a willing buyer would pay in cash to a willing seller' at the time of the taking.'' (citations omitted).)} A large chunk of the remaining work, in turn, can be delegated to the appraisers, allowing the courts to get on with other business. In addition to being effective, this comes with the added bonus of allowing the courts to distance themselves very clearly from the political overtones of the legitimacy question.

This approach to legitimacy is prevalent, but I believe it needs to be rejected. The reason, as argued in the first part of this thesis, is that property itself cannot be drawn up as narrowly as the compensation approach presupposes.\footnote{See the discussion in chapter \ref{chap:2}, section \ref{sec:2:4}.} The legitimacy of property interference cannot be understood in mechanical terms, as a matter for the appraises, regardless of how much weight we want to place on the value of efficiency and the ideal of deference to political decision-making. Specifically, unless the issue of legitimacy is recognised as being far more complex than pertaining solely to the question of when and how to calculate market values, there will be a significant mismatch between what property is and what the law pretends it to be.\footnote{As contended in Part I of this thesis, this much appears to be a descriptive insight, which does not seem to depend on one's moral philosophy, see the discussion in chapter \ref{chap:2}, section \ref{sec:2:4:3}.}
\noo{In a setting where takings are rare and happen only in extraordinary situations, such a mismatch might be tolerable. Arguably, the strong commitment to the sanctity of property by early writers such as Blackstone, apparently conflicting with historical records, could be sustained because takings were exceptional, ordered only after careful deliberation by a legislative body with an authority over property to match or exceed even that of an owner.\footnote{See chapter \ref{chap:3}, section \ref{sec:3:2}.}

Such a narrative is no longer plausible in a world where the state has expanded its activities so much that interference in private property, rather than being exceptional, have become a normal occurrence. This also means that the law cannot pretend not to interfere also with the complexities of property as a social phenomenon and an anchor for participation and human flourishing. This broader narrative of property must then also be considered in the law, especially in the law of takings.}
This is harmful at a structural level, especially if the idea of property as a fundamental right is to have a future. If the law continues to insist that property is nothing more than a form of entitlement protection, there might even be a case to be made that the notion should just be done away with in its entirety.\footnote{See generally \cite{grey80}.}

In the first part of this thesis, I presented a theory of property which suggests that this would be a tragedy, particularly for marginalised groups who are in need of protection against economic and social elites. Importantly, while international law and human rights conventions offer important clarifications and protections at the level of principles, what property provides is an imperfect, yet very powerful, framework for implementing such principles at the local level. This is property's promise, which it can only keep if it is recognised as having social functions going well beyond the protection of individual entitlements.

Arguably, principles of human rights should even be recognised as inhering in property as such, not only as mediated by the power of states.\footnote{See chapter \ref{chap:2}, section \ref{sec:2:5:1}.} A worry often voiced as a counterargument for direct horizontal application of human rights is that it can serve as an excuse for the state to do nothing.\footnote{See \cite[110]{manisuli07}.} For this reason, it might well be worth emphasising that if the owners fail to deliver on basic rights, the state is still responsible. However, the converse is equally true: if states fail, then owners still have obligations.

If there is no distinction between owners and states, by contrast, failure in one is also failure in the other. This seems dangerous, particularly if proprietary power is exercised only by a small group of people, regardless of whether they are commercial leaders or powerful government officials. Property should therefore be widely distributed among the population, to be rendered as a provider of basic rights and an anchor of democracy.

The first chapter of this thesis explored this idea in depth and argued that a broader notion of property needs to be acknowledged also by the law, particularly in the law of takings. If property serves a broad social function by sustaining a community and aiding in the delivery of basic rights to all its members, transferring that property to a non-local commercial owner is not merely a highly dubious redistribution of entitlements. In the end, it will also be the destruction of property, as it undermines property's most significant functions in relation to the overarching goal of human flourishing. In such cases, therefore, property is not only taken, it is lost.

\noo{
t is worth pausing to recognise that the bundle of rights theory did us a favour in this regard, in that it directed our attention at the multifaceted nature of property. However, to make progress, it was necessary to further unpack the property bundle, to get at the substantive content of property in life: social functions as opposed to legal abstractions.

\noo{ Moreover, I argued that while private property might often be found wanting, it remains a potentially powerful force for good in the world. As discussed in chapter 1, its roles as a building block of democracy and a protector of communities is particularly important in this regard. In addition, I discussed the importance of social obligations arising from property, and how they can potentially function as a guarantee that the basis rights of non-owners will be delivered at the local level.

If these aspects are recognised and embraced as a crucial part of the concept of property in the law, it should hopefully go some way towards restoring confidence that property is neither theft nor fraud, but a promise to work hard for a better future for all. By contrast, the idea of property as a financial entitlement do not appear to offer any such relief. If anything, dismissive attitudes to property will take their fuel from the idea of property as entitlement; entitlements, after all, can often appear undeserved, especially when they are not checked by corresponding obligations. 
}
%In this way, a threat emerges to the stability of the concept of property itself, as a legitimate part of the social and political order.
%In reality, of course, owning property has nothing to do with what one deserves, but rather what task one has been allotted on this earth, to pursue in keeping with one's beliefs and convictions.

%In cases involving regulation of property use, it might still be possible to keep this aspect away from undermining the concept of property as such. At least, it might ensure that only those well versed in the technical details of the law are able to recognise property for the ``phantom'' that it appears to be.

%However, when property is taken outright, even this containment strategy is bound to falter. This is particularly clear if taking become increasingly prevalent not only in situations of pressing public need, but also as a means for companies to turn a profit. In such a setting, even the most naive observer would surely be tempted to think that property as a concept must be altogether rather vacuous.  


%The impression that private property, in the end, is nothing but a shorthand to describe a special class of liability rules, leaves property open to further attack. Indeed, if property and ownership has only such a thin content, why worry about interfering with it in the public interest? At this point, however, it seems prudent to take a step back, to reconsider the origin of the feeling that property is no more than theft, or no better than fraud. Specifically, it might be appropriate to note that unlike property as a concept, the act of taking it without its owner's consent is quite likely to involve both theft or fraud as a matter of fact, not merely a manner of speech.

%In the second part of this thesis, I have explored this in further depth, by analysing the law and practices relating to the expropriation of waterfalls in Norway.

%However, as I noted in the first chapter of this thesis, property itself is highly multifaceted, serving a range of social and individual functions. 

The first chapter set out to do this, in order to get at the multitude of different ways in which a taking can impact on society and its members. The economic consequences of a taking might be the most easily recognisable, particularly in the economic development cases. But as I have argued in this thesis, other consequences can be just as important, particularly those pertaining to property as a building block of communities. If jointly owned property is taken from a community, with full compensation paid to all individual owners, the community suffers a distinct uncompensated loss, namely the loss of future self-governance opportunities. %In light of work done by Elinor Ostrom and others on the strength of self-governance solutions for sustainable resource management, the law of takings should arguably offer protection directed specifically at the potential loss of community.

%In the traditional narrative on takings, social and political effects are typically only recognised on one side of the takings equation, namely the side of the taker, particularly the public interest. This has also influenced the debate on economic development takings. In order to make sense of the broader sense of unfairness often associated with such takings, critics tend to focus on the taker rather than the owner, by questioning the legitimacy of the motives behind the taking.

%However, this might be tantamount to shifting a variable to the wrong side of the takings equation. In particular, the feeling of unfairness associated with economic development takings clearly arise from a sense in which the owners are victims of an abuse of power. So why shift attention to the taker? 

%Perhaps it is tempting to do so simply because the sense of unfairness at work here pertains to a broader notion of justice than that normally associated with property interests. If so, the entire narrative points to a shortcoming of the liberal idea of property. If even property's staunchest defenders must turn to notions of ``public interest'' (and the lack thereof), then why do we need property as a concept at all? Why not simply say that a licence to undertake economic development should not be granted unless all affected parties agree, or the public interest is sufficiently strong to go ahead against some of their wishes? What makes property special in this picture, if all that is at stake is the strength of the public interest used to justify imposing the state's will on private individuals?

%Clearly, the gaping hole in the opposition to economic development takings in the US has been a {\it positive} account of 

%This part of the thesis focused on coming up with an answer as to why property is worthy of protection in the first place, in cases where economic rationality appears to dictate that it should be put to more profitable uses. If there is a reason to resist this, it must be because there is something valuable in property that the law should protect, irrespective of the current owner's financial entitlements.

Moreover, the thesis argued that the dynamics of power in takings cases need to receive more attention: the practice of economic development takings can result in local communities being deprived of highly valuable political capital in order for politically powerful commercial interests to make a profit. I noted that this was also the main concern raised by Justice O'Connor in her {\it Kelo} dissent.\footnote{\cite{kelo05}.} In this way, the social function view is arguably also implicit in one of the most forceful voices that have spoken out against economic development takings on the basis of constitutional property law.

However, the thesis also noted a weakness with the typical approach to legitimacy in the US, via the public use requirement. Specifically, this requirement does not appear to get us very far towards a justiciable restriction on the takings power along the lines of reasoning adopted by Justice O'Connor. Unlike the majority in {\it Kelo}, building on recent precedent, and Justice Thomas, building on the original meaning of the public use restriction, Justice O'Connor's more institutionally oriented reasons for rejecting the taking appeared to lack a firm basis in law.

To address this, I pointed to recent developments at the ECtHR, where an institutional perspective on fairness appears to be developing, which might be more likely to embrace the social function perspective on property as a basis that can support a justiciable restriction on the states' takings power. At the same time, the position of the Court in Strasbourg might be conducive to an approach that can adopt broad scrutiny in controversial cases without becoming too entangled with the politics of those cases at the state level. Specifically, the value of deference could be given a firmer expression as a norm that compels recognition of diversity and local democracy, not a norm that calls for passivity or loyalty to governments in politically sensitive situations.

Following up on this, the thesis went on to formulate a legitimacy test based on a set of conditions formulated by Kevin Gray. Three additional points were added to this list, emphasising the regulatory context, the position of non-owners, and the broader issue of democratic merit. 
}

Arguably, takings of this kind are currently being carried out in Norway, to the benefit of large energy companies. Waterfalls are still nominally considered private property, belonging to members of the rural communities in which the water resources are found. However, as shown in Part II of this thesis, the practice of taking waterfalls from local communities has become so insensitive to the plight of owners that the law itself is ambiguous about whether private ownership of water resources has much content at all.

%It is important to note that there is nothing inevitable about this state of the law. Indeed, the historical context shows that in Norway, where water is anything but scarce and most rivers are non-navigable, property rights in waterfalls (as opposed to water as a substance), was long recognised as on par with property rights in land. Expropriation to pursue hydropower development was not permitted under any circumstance until the early 20th century. 

But this does not mean that proprietary power is no longer exercised over the power of water. Quite the contrary. The commercial companies that acquire waterfalls are quick to turn them into commodities whose primary purpose is to turn a profit. At the same time, water resources have now been encapsulated in development licenses; today, licenses from the government, not waterfalls as such, are the key assets in the hydropower production sector. The nature of property has thus transformed; ownership of hydropower has become a legal fiction, arising from a bundle of papers with numbers on them, not from physical and social proximity to the underlying resource.

The result is that the notion of property backing up this regime is now so thin that it arguably cannot be distinguished at all from the political and economic power that backs it up. As such, it is also no wonder that narratives of egalitarian property give way to narratives based on thinking about water resources as though they belong to the ``public''.\footnote{See the \indexonly{ica17}\dni\cite[1]{ica17}.} This is a politically defensible way of talking, to mask the reality that water resources are managed according to the will of dominating market players and powerful interest groups, both striving to protect their dominion over the fruits of the land.\footnote{I mention that state-owned companies are currently extending their dominion elsewhere as well, with potentially deleterious consequences. In Nepal, for instance, Norwegian hydropower companies generate large profits from their aid-funded cooperation with the Nepali government, apparently at the expense of local populations and the rights of the poor. See \cite{gaarder15} (report from one of the largest environmental organisations in Norway, discussing how the Nepali government entered into an agreement where the price of electricity would be set at an extortionate level, denominated in US dollars, and adjusted upwards proportionally to US consumer prices; apparently, the project involves a transfer of money from Nepal to Norway that far exceeds the flow of money going the other way); \cite[644]{peris12} (mentioning the presence of Norwegian actors, discussing the marginalisation of local people, and arguing that conventions on indigenous rights are unable to deliver social justice due to the ``democratic deficit'' of decision-making regarding hydropower development in Nepal).}

Due to this dynamic, the case of Norwegian waterfalls appears to be an example of how illegitimate takings can do more harm than to deprive owners of valued resources. Specifically, the case study shows how a lack of legitimacy can effectively deprive property of its meaning, as the sticks of the 
bundle are bent to suit the interests of the commercial and political elites. Such a system might well give us the impression that property is little more than theft, maintained in the law only as a fraud.

\noo{ However, given the historical context Perhaps, then, the nature of property itself has changed, so that there is nothing left except those financial entitlements that Norwegian expropriation law recognised. If so, the change has not come about by any legislative move, nor has it been preceded by any kind of debate. It has simply emerged, gradually and unplanned, as a result of sector-based regulation and administrative practices. The process, therefore, meets neither the requirements of land reform or expropriation. It is an unacknowledged process about which the law in Norway has had nothing much to say at all, for which silence still persists. 

%Property can be an elusive concept, especially to property lawyers. Indeed, in the law of property, the word itself typically only functions as a metaphor -- an imprecise shorthand that refers to a complex and diverse web of doctrines, rules, and practices, each pertaining to different ``sticks'' in a ``bundle'' of rights. Indeed, this bundle perspective dominates legal scholarship, especially in the common law world. Some even go as far as to argue that words such as ``property'' and ``ownership'' should be removed from the legal vocabulary altogether. 

%So is property as a unifying concept lost to the law? It certainly seems hard to pin it down. In the words of Kevin Gray, when a close scrutiny of property law gets under way, property itself seems like it ``vanishes into thin air''.\footnote{See \cite[306-307]{gray91}.} %Indeed, one may argue that different ideas of property, practical and theoretical, are behind most, if not all, the major conflicts and confrontations that have shaped the society in which we live.
%Responding to this, some prominent philosophers have taken the view that property is not a concept suitable for philosophical study at all.
%According to some philosophers, property as a concept is a lost cause, not suitable for conceptual analysis at all. Instead, these scholars have suggested that property should be taken as a pragmatic and contingent derivative of other notions, such as the social order, or, on a normative account, by regarding it as an expedient of {\it justice}.
%Moreover, legal scholars are usually content with theories of property that remain largely descriptive, settling for the more modest aim of exploring how best to think of property given the prevailing legal order, rather than trying to come up with theories to explicate its nature as a pre-legal concept. Indeed, even legal philosophers are sometimes found doubting that there even is such a thing as property.
%Arguably, however, property never truly disappears. Indeed, there is empirical evidence to suggest that humans come equipped with a {\it primitive} concept of property, one which pre-exists any particular arrangements used to distribute it or mould it as a legal category.\footnote{See\cite{stake06}.} Perhaps most notably, humans, along with a seemingly select group of other animals, appear to have an innate ability to recognise {\it thievery}, the taking of property (not necessarily one's own) by someone who is not entitled to do so.\footnote{See \cite[11-13]{brosnan11}.}

%Taken in this light, Proudhon's famous dictum ``property is theft'', might be more than a seemingly contradictory comment on the origins of inequality. It might point to a deeply rooted aspect of property itself, namely its role as an anchor for the distinction between legitimate and illegitimate acts of taking.

%But what is a taking, and when is it legitimate? In this thesis, I will aim to make a contribution to this question. I will study takings of a special kind, namely those that are implemented, or at least formally sanctioned, by a government. In legal language, especially in the US, such acts of government takings are often referred to as takings {\it simpliciter}, while talk of other kinds of ``takings'' require further qualification, e.g., in case of ``takings'' based on contract, tax or occupation. 

%The US terminology brings the issue of legitimacy to the forefront in an illustrative manner. We are reminded, in particular, that under the rule of law, taking is not the same as theft. Rather, the default assumption is that the takings that take place under the rule of law are legitimate. If they are not, we may call them by a different name, but not before. At the same time, it falls to the legal order to spell out in further detail what restrictions may be placed on the power to take. 

%Indeed, restrictions appear implicit in the very notion of taking. The idea that someone might find occasion to resist an act of taking, and may or may not have good grounds for doing so, appears fundamental to our pre-legal intuitions. But how should we approach the question of legitimacy of takings from the point of view of legal reasoning, and what conceptual categories can we benefit from when doing so? This is the key question that is addressed in this thesis.
}

\section{Property Regained -- Giving and Participating}\label{sec:7:2}

The converse of a taking is a {\it giving}. In the US, this term is sometimes used to refer to situations when private property owners benefit from state actions involving property.\footnote{See generally \cite{bell01}.} For instance, it might be characterised as a giving if the state allows someone to purchase property cheaply, or if regulation makes some properties appreciate in value. Arguably, and analogously to the case of takings, there is a case to be made that the state should sometimes charge people for disproportionate givings.\footnote{See \cite[590-604]{bell01}.}

The issue of when this is appropriate, if at all, will not be discussed here. Instead, I wish to direct attention at the terminology itself, and its subtle conceptual commitment to a top-down way of thinking about both takings and givings. Indeed, consider what happens if we turn the terminology on its head. This is not an implausible conceptual shift. After all, if the state takes property, the current owners will have to give it up.

The owners' act of giving, however, is rarely given much recognition or attention when we approach takings. Plainly, the owners' active participation as a giver -- not merely an injured party -- is typically considered irrelevant. Why is that? The obvious answer is that since the giving takes place under compulsion, it does not express any intention to give. However, there are many situations in life where actions are compelled, but where the person taking that action still gets some credit for it.\footnote{Paying taxes, for instance, is typically associated with being a good citizen.} Moreover, the owners can clearly decide to be more or less cooperative when faced with the government's wishes for their property. 

By shifting attention towards the choices that the owners have in this regard, we can hope to find a path towards increased legitimacy by recognising the owners as active participants. Indeed, even a purely symbolic recognition of the owners' role and the importance of their choices in dealing with a takings request, can serve to enhance subjective legitimacy. However, quite apart from recognising the constructive role that owners can play in the existing system, thinking about takings as givings also suggests the possibility that owners should be granted more choices and asked to take part more actively in the proceedings.

%The owners make an active contribution to the public purpose, instead of being regarded as obstacles to it. This, in turn, can become a starting point for coming up with arrangements where the owners are permitted to take up a more lasting interest also in the new use of the property that the public desires. 

In cases involving economic development, this way of thinking seems particularly appropriate. As they contribute to the project by giving their property, it seems only fair that the owners should have a stake also in the planning and the continued use of their property for economic gain. For instance, it might be appropriate to offer owners shares in the development company, and to make sure that the property is taken under a leasehold rather than by a full transfer of title. Better yet, one could allow the owners themselves to deliberate on how they wish to honour their commitment to the public, to formulate their own plans and implement their own solutions, based on continued interaction both among themselves and with representatives from the relevant bodies of government.

Interestingly, as shown in chapter \ref{chap:5} of this thesis, such an abstract and highly idealistic idea is in fact (partly) implemented through the system of land consolidation found in Norway. As noted, this institution is even equipped with the power to compel owners to come together and participate in specific endeavours. If a proposed development is judged as being beneficial to the properties in question, the owners may be stripped of their holdout power, not only as individuals but even as a group, without any property having to be condemned. The owners will retain their ownership even after their properties have been put to new and more productive uses.

In light of this, land consolidation can indeed replace expropriation, provided our understanding of what property is, and should be, is broad enough to say that the purpose we wish to pursue is also in the interest of the properties involved. This limitation, expressly encoded in the law of consolidation in Norway, is interesting also on the theoretical level, because it makes the {\it purpose} of property directly relevant to determining the extent of the government's power to interfere with it. At the same time,  in the context of economic development, the restriction is rarely going to prevent consolidation from going ahead. If development is both economically beneficial and represents a sustainable use of resources, the land consolidation courts should have no difficulty justifying consolidation measures on the basis that development is also in the interest of the properties involved. At the same time, the context of land consolidation, with its emphasis on problem solving and owner participation, means that the process will usually be far more inclusive towards owners than traditional expropriation proceedings. 

If property is in the hands of the few, while many property dependants are without formal ownership rights, the consolidation model might be less appropriate. However, in these cases, it might be possible to adapt it by allowing a larger group of local people to partake in the proceedings. In complex cases, most of them might find themselves unable to participate effectively in the process. However, a consolidation approach will still serve an important function in that it gives marginalised groups new opportunities for engaging with the democratic process.

%Under consolidation, it is truly more appropriate to speak of givings than to speak of takings. Moreover, what is given in these cases is not normally the property as such, but the right to determine how it is to be used, with the public having to turn to the current owners for help in realising the public purpose. It would still be possible to expropriate, or to rely on a mix between expropriation and consolidation. In Norway, the relationship between the two remains to be clarified in the law. A general rule of consolidation first, expropriation second, 
%should arguably be introduced in order to better promote the consolidation alternative.

%This thesis went on to explore how this alternative works in practice for hydropower development, noting that the current system works best when the owners are in reasonable agreement with each other and society about how development should proceed. Hence, the broader applicability of the idea can not be taken for granted. However, it seems to have great potential for being fine-tuned, to make it more effective in situations involving deeper disagreements and conflicting interests. There is a danger here, however, namely that increasing the power of the institution will also undermine its role as a democracy-on-demand for owners and the community. However, if safeguards can be put in place, the land consolidation model might well be a highly attractive alternative to expropriation, also in cases involving deeper conflicts about property uses.

The system can also serve an empowering and educational role. Local elites might often dominate the process in practice, but the fact that the procedure is judicial in nature means that abuses can be curbed while the disadvantaged may be granted access to more fruitful forms of citizenship. At least, access to the decision-making process should become easier for those who wish to challenge the local leadership.

More generally, the flexible, issue-focused, and transient nature of a consolidation court might have significant advantages. Specifically, concentrations of power and participatory fatigue are unlikely to become entrenched, given the context of decision-making; the group of people involved, the area covered, and the agenda to be deliberated on, will all change from situation to situation. This keeps participants engaged while diffusing power, thereby minimising the risk of elite tyranny and expert rule, otherwise easily enabled by rigid institutions and apathetic majorities.\footnote{This also relates closely to the worry that ``rational ignorance'' can thwart attempts at meaningful eminent domain reform, see chapter \ref{chap:4}, section \ref{sec:3:3:2}. See also \cite{somin09}.}

Further exploration of the transferability of the consolidation framework to other contexts will have to be left for future work. What this thesis has hopefully shown is that fully fledged institutional alternatives to expropriation for economic development already exist and that they can work in practice. Hence, my research can hopefully inspire more work in this direction both at the theoretical and empirical level. In my opinion, the ideas underlying the land consolidation system found in Norway are worth exploring further, also in other jurisdictions that rely on expropriation as a tool to facilitate economic development. In this context, it would be a great victory for both property and equity if public and private interests could come together in a way that will give rise to givings in the future, clearly distinguished from the takings of the past.
}