%\newcommand{\isr}[1]{{#1}}

\chapter{Possible Approaches to the Legitimacy Question}\label{chap:2}

\section{Introduction}\label{sec:intro}

In the previous chapter, I introduced the social function perspective of property and argued in favour of a normative approach to property based on the notion of human flourishing. Moreover, I argued that economic development takings make up a separate category of interference with private property, deserving of special attention. I also placed this category in the theoretical landscape, by relating it to the theory of property presented in the first part of the chapter. Specifically, I argued that economic development takings raise questions that require us to depart from the individualistic, entitlements-based narrative that has tended to dominate in property theory.

This chapter considers the question of legitimacy of economic development takings in more depth. First, it presents there existing approaches, based on evidence from England and Wales, the United States, and the European Court of Human Rights. I start by considering the idea that parliament itself imbues each instance of a taking with legitimacy, as the result of a decision made in a legitimate manner within a democracy. Such a narrative leaves little room for judicial scrutiny of takings, and the doctrine of deference becomes the norm. In England and Wales, this approach carries great weight, particularly historically, when parliament itself would authorise most takings directly through so-called private Acts. 

I argue that the expanding state and the increasingly expanding use of eminent domain puts this perspective on legitimacy under strain. To some extent, it can be upheld by a well-organised executive, compelled to remain faithful to parliament and the ideas of democracy. However, a threat to the stability and success of such a procedural approach can arise from the lack of any clearly defined safeguards to protect against institutional failure and substantive abuse. If property as an institution begins to falter, for instance because takings for profit become too prevalent, the courts might find themselves unable to intervene on behalf of those democratic ideals that motivated the idea of deference in the first place. I will argue that recent cases of economic development takings in England illustrate this danger, suggesting that we should also consider substantive approaches to legitimacy.

Following up on this, I go on to consider the US, where the public use restriction is considered to be an important substantive limitation on the government's power to take property. I track the history of public use scrutiny in some depth, showing that it was widespread and extensive at the state level, at least until a contrasting position of almost unconditional deference to the legislature was adopted by the Supreme Court in the case of {\it Berman}. After this, at the federal level at least, the public use restriction was effectively stripped of its content.\footnote{See \cite{berman54}.} The eventual backlash of this came with {\it Kelo}, which was decided in keeping with precedent, but with severe doubts arising among the justices, particularly those who looked at the history of the public use doctrine and how it had worked prior to {\it Berman}.

I argue that a contextual approach to the public use requirement, based on broad assessments of local conditions, was prevalent among the state courts in early public use cases. This arguably also reflects a social function understanding of property, connecting the public use test to the property theorising in chapter 1. Moreover, I note that there has been a resurgence of extensive public use scrutiny after {\it Kelo}, particularly at the state level. I note, however, that this change appears to have been largely ineffective at curbing dubious uses of eminent domain. Specifically, it has been argued that recent reforms, and broad substantive standards such as the public use requirement, are likely to become only symbolic nods to the danger of abuse: hidden within the complex arrangements of modern government, there is business as usual regardless.

This in turn raises the issue of how to combine the institutional and the procedural perspective on legitimacy, to ensure that substantive standards actually translate into effective protection. This brings me to the third approach to legitimacy, which I call the institutional fairness approach. I argue that this approach has been adopted recently by the ECtHR in Strasbourg, as they have developed a system of pilot judgements to deal with their vastly increasing case load. The idea of such judgements is that the Court will focus on systemic problems, to determine whether they should order the state to take general measures to improve their own institutions. By doing this, the Court will protect itself from having to deal with many similar cases. Instead, it can move on to novel issues of principle that need to be considered.

Quite apart from the practical motivation behind this development, I argue that the institutional perspective on fairness that it conditions is the way forward towards testing for legitimacy in takings cases. It should work well because it allows courts to adopt a middle ground between the procedural and the substantive approach. I consider the case of {\it Hutten-Czapska v Poland} in some depth to argue for the merits of this approach.\footnote{See \cite{hutten06}.}

Following up on this, I consider the question of how the courts should proceed to assess the legitimacy of economic development taking against such an institutional fairness perspective.\footnote{There is not yet any case law on this from the ECtHR.} Building on a list of conditions due to Kevin Gray, I propose a concrete heuristic for this purpose. In addition to the original points made by Gray, I add three of my own, inspired by the discussion in this and the previous chapter. 

A legitimacy test can never provide more than a partial solution to the legitimacy problem. Specifically, in cases when the desire for economic development is a genuine reflection of democratic decision-making, the follow-up question is how to better enable the collective to communicate this desire to private owners, without resorting to eminent domain. I address this question by looking to the theory of governance for common pool resources, developed by Elinor Ostrom and others. Specifically, I note how the connection between this theory and property law suggests the possibility that new institutions should be introduced to allow the collective to push for economic development involving privately owned property. In fact, such a proposal has already been made, by Heller and Hills, who proposed that so-called {\it Land Assembly Districts} could replace the use of eminent domain in many cases when holdouts make economic development on privately owned land hard to implement.  

I analyse the proposal in some depth, pointing out problems to suggest that Land Assembly Districts are not the final answer to the legitimacy question for economic development takings. Specifically, I note the context dependence of their proposal, and the underlying tension between the ideal of self-governance and the fear of tyranny by local elites. This goes to show that the cure for illegitimacy, much like its diagnosis, depends on the circumstances, and what one regards as the property's proper function. In light of this, I believe the critical examination of Land Assembly Districts marks a natural end to this chapter, as well as to the theoretical part of this thesis as a whole.

\noo{

Hence, what is the right approach for a prosperous nation with an egalitarian distribution of property rights might not be right for a nation marked by inequality and poverty.

I will return to this in Chapter 5, by presenting an alternative institution, already operational in Norway, which I believe might be preferable to Land Assembly Districts in many situations. However, 
In the second part, I will draw on the work done in this and the previous chapter to analyse the particular problems and possible solution that arise with respect to takings for hydropower development in Norway.



, might be quite different from what should 



 as a solution, unlike what is inappropriate, seems to depend on the totality of the legal order in which the 

Following up on this, I 



the system becomes susceptible to 

 develops this idea further, by considering how economic development cases are dealt with in England, where such takings have yet to be widely recognised as a separate category, and the US, where they first began to attract special attention. In addition, the chapter considers case law from the ECtHR and asks what it tells us about how to approach economic development takings under European human rights law.\footnote{So far, the issue of economic development takings have been brought into focus at the Court in Strasbourg.}
Finally, the chapter considers recent proposals for reform that focus on how to increase legitimacy by developing new institutions for self-governance to replace the traditional takings procedure in economic development cases.

I begin in Section \ref{sec:lgppp} by commenting briefly on the importance of economic development takings on the global stage. Specifically, I note that the core issues raised by such takings appear relevant also in the context of developing economies, even when property rights as such are an unstable basis on which to reason about the rights and obligations of individuals and communities. Specifically, I propose that the social function theory might offer a conceptual bridge between the study of economic development takings and the study of {\it land grabbings}, large-scale land acquisitions in the developing world. In both cases, the worry is often that local communities, who might lack formal title to the land, suffer as a result of a dramatic change in property's social function.

In Section \ref{sec:contrast}, I move on to consider the status of economic development takings in  English law. This also serves to introduce the topic of my thesis from the point of view of an important jurisdiction in Europe, where the issue of economic development takings has attracted far less attention than in the US. It appears to be gaining importance, however, as public-private partnerships and a market-oriented approach to public services has become influential in many jurisdictions, including in England.

In Section \ref{sec:echr}, I elaborate on a practically significant pan-European property clause, namely Article 1 of Protocol 1 (P1(1)) of the European Convention of Human Rights (ECHR). I argue that this clause provides an interesting perspective on the legitimacy issue, asking us to focus on the proportionality of the interference, judged relatively to its social and political context. I also consider some possible objections against the human rights approach, including the worry that the court in Strasbourg is not well-placed to be the arbiter of social and individual justice throughout Europe. At the same time, I point to some recent decisions at the Court that I believe signal hope that the case law on property is moving away from ill-conceived ``micro-management'', towards a more open-ended jurisprudence that seeks to force member states to address systemic problems that they might otherwise be reluctant or incapable of raising to the national agenda. Here the involvement of a (hopefully) politically neutral institution like the ECtHR can serve an important purpose, particularly if it manages to tailor its own case law in such a way as to leave room for local institutions of the member states to work out for themselves how to concretely resolve human rights issues flagged by the Court in Strasbourg.

In Section \ref{sec:us}, I return to the US setting, by presenting in detail how the perspective on economic development takings, mediated through case law on the public use restriction, has evolved since the 19th century until today. I structure the presentation as a story in two parts, describing the situation before and after the {\it Kelo} case. For the pre-{\it Kelo} presentation, I begin by pointing out that the case law on the public use restriction was initially developed by state courts, who would adjudicate legitimacy cases against the respective state constitutions (which typically also contain some sort of public use restriction on the takings power). 

The Supreme Court adopted deference to state {\it courts} initially, before changing their perspective by adopting a policy of deference directed rather at the state {\it legislature} (in practice also the administrative branch). I argue that this shift in Supreme Court jurisprudence can be pin-pointed to the case of {\it Berman}.\footnote{See \cite{berman54}.}

I go on to argue that this shift in case law at the federal level had the effect of destabilizing the established state approach to economic development takings, resulting in increased tension and controversy, paving the way to {\it Kelo}. In essence, my argument is that the Supreme Court was right in taking a deferential stance with respect to local institutions, but wrong in stripping the public use restriction of content, a move that threatened to undermine the authority of state courts. In effect, the federal takings jurisprudence threatened to weaken a very sensible {\it local} judicial constraint on executive power, a constraint that was also important to the proper division of power at the state level.

In Section \ref{sec:postkelo}, I follow this up by a discussion of developments after {\it Kelo}, which has seen a resurgence in state court scrutiny of the public use requirement, often backed up by state legislation that explicitly seeks to limit the scope of takings for economic development. According to some, such state reforms have been largely ineffective. %Ilya Somin, one of the most prolific writers on economic development takings in the US, has argued that this is partly due to so-called ``rational ignorance'' of political decision-makers and voters regarding the subtleties of the public use issue. The idea is that the distance between policy makers and communities affected by economic development takings is too great, so that policy makers have no incentive to consider the finer details of the takings equation. 
In principle, the US public is almost unanimously on the side of the local communities in cases like {\it Kelo}, but in practice, the great distance between political cause and effect makes effective reform policies hard to formulate. The danger is that reform proposals come to rely on oversimplified narratives tailored to centralised processes of decision-making.

In Section \ref{sec:ir}, I consider a proposal due to Heller and Hills that serves as a possible answer to this concern.\footnote{See \cite{heller08}.} This proposal focuses on the need for new frameworks for collective action, institutions that can replace the top-down dynamics of eminent domain in cases of economic development. The goal is to ensure a greater level of self-governance for the communities directly affected by the development, the individual members of which have a rational incentive to invest time and effort in reaching sophisticated compromises that can replace the use of black-white solutions (be it in the form of an economic development taking or a politically sanctioned top-down {\it ban} on such takings).

I argue that this idea embodies both a natural and necessary counterpart to increased judicial scrutiny of the public use restriction. In particular, I argue that the two ideas are mutually conducive to each other, when properly conceived. This argument will set the stage for the case study in the second part of the thesis, where I explore the tension between takings and self-governance in the context of hydropower development in Norway.

\noo{ In the following, I will present a comparative background for my case study. I will begin by considering English law, where courts have generally been reluctant to broadly scrutinize the use of economic development as a justification for state interference in property. After this, I turn to the ECHR and the proportionality test that is now at the core of property adjudication at the ECtHR. I note that while states are considered to have a wide margin of appreciation with regards to the legitimacy of the purpose underlying interference, the balancing required under the proportionality test can still become a powerful basis on which to scrutinize the broader negative effects of economic development takings.

Following this, I move on to consider the US in greater depth, both the historical debate that led to {\it Kelo} and the suggestions for reform that have emerged following its backlash. There has been much written about this issue in the US. Moreover, while much of it is repetitive and coloured by the tense political climate, I believe some historical points, as well as some recent suggestions for reform, are highly relevant also to the international setting. To single out and analyse those aspects is the main aim of this part of the chapter. Indeed, the current debating climate in the US might be an indication of what is to come also in Europe, if concerns about the legitimacy of economic development takings are not taken seriously. }

%I also highlight what I believe to be a connection between the situation in the US leading up to {\it Kelo} and the present situation in Europe, illustrated by the fact that the European Court of Human Rights is now explicitly endorsing ``stronger protection'' of property rights.  I attempt to identify the reasons behind calls for a stricter approach, arguing that it is connected to the fact that interferences in property under modern regulatory regimes is sanctioned in wide a range of different circumstances, serving to undermine their status as a necessary burden imposed on owner's according to the will of the greater public. In some cases, rather, takings appear to both owners and the public as improperly motivated and socially and politically unfair. I note that this happens particularly often in economic development cases, when commercial actors benefit to the detriment of local communities. I go on to list some concrete issues that arise with respect to such takings and that have been flagged as problematic in the literature.
%
%Following up on this, I consider various proposals that have been made to resolve tensions and limit the possibility of abuse in economic development cases. The differences of opinion that have been expressed in this regard have been quite substantial, and proposals have ranged from suggesting an outright ban on economic development takings  (Somin 2007; Cohen 2006) to suggesting that the best way forward is to reassess principles for awarding compensation in such cases (Householder 2007; Lehavi and Licht 2007).

%Much of the current theory focus on assessing traditional judicial safeguards that courts can rely on to prevent abuses, pertaining primarily to the material assessment of proportionality, public purpose, and compensation. 

%In the last part of the chapter, I will focus on a very interesting strand of recent work in the US, which shifts attention towards procedural rules that can help address the worry that economic development takings tend to suffer from a democratic deficit. The core concern is that the manner in which eminent domain decisions are typically made, and the way in which owners are compensated, might be unsuitable for economic development cases. Importantly, the need for special procedures has been noted, to restore legitimacy.\footnote{See generally \cite{lehavi07,heller08}.} This ties the US debate even closer to the European context, where proportionality, not public use, has become the key notion in property protection. Several recent suggestions from the US can be conceptualized as suggestions that aim to secure fairness and proportionality, while paying less attention to the formalistic question of what constitutes a ``public use''.
%
%%Also, it allows us to be very clear about a special concern that arises for economic takings cases: under current regulatory regimes, the government and the developer together often dominate the decision-making process completely, leaving the property owners marginalized. Hence, there is often a {\it democratic deficit} in such cases, resulting in discontent and a feeling that the taking is not in the public interest at all. Importantly, some recent writers hypothesize that if the proper balance can be restored in the decision-making process, so will the decision reached appear more legitimate, also with respect to the public use clause. In my opinion, this idea is crucial, and together with the question of compensation, which raises a similar structural problem, it will guide the rest of the work done in this thesis. 
%

In response to that worry, this chapter aims to bring into focus the key question of how to ensure meaningful participation for owners and their local communities in decision-making pertaining to economic development on their land. The tentative answers provided in Section \ref{sec:ir} will set the stage for the remainder of the thesis, where these answers will be assessed in depth against the case study of Norwegian hydropower.

%In particular, I will consider two special semi-judicial procedural systems used in such cases in Norway, one targeting compensation following expropriation, and another used as an alternative to expropriation, particularly in cases when development requires cooperation among many owners.

%I conclude by arguing that approaches along procedural lines represent the best way forward in relation to addressing issues associated with economic development takings. This raises the following problem, however: what procedural principles can be used to ensure meaningful participation, without hindering socially and economically desirable development projects? This question sets the stage for the remainder of my thesis, where I conduct a case study of expropriation for the development of hydro-power in Norway. In particular, I will consider two special semi-judicial procedural systems used in such cases in Norway, one targeting compensation following expropriation, and another used as an alternative to expropriation, particularly in cases when development requires cooperation among many owners.

}

\section{England and Wales: Legitimacy through Parliament}\label{sec:contrast}

%Economic development takings have not become as controversial in Europe as they are in the US, but there have been cases where the issue has come up, in several different jurisdictions.\footnote{For instance, in the UK, Ireland and Germany, as well as in Norway and Sweden. See \cite[466-483]{walt11}; \cite{stenseth10}.} \noo{ The P1(1) of the ECHR protects property, but the legitimacy of economic development takings has not yet been discussed in case law from the European Court of Human Rights (ECtHR). However, it is interesting to analyse cases like {\it Kelo} against P1(1), particularly since the ECtHR has developed a doctrine that focuses on ``proportionality'' and ``fairness'' rather than the purpose of interference.\footnote{See generally, \cite[Chapter 5]{allen05}. This approach may become even more significant as a source of property protection in the future, as the ECtHR have indicated that there are ``jurisprudential developments in the direction of a stronger protection under Article 1 of Protocol No. 1'', see \cite[135]{lindheim12}.}}

%In this section, I address economic development takings from the point of view of English law.\noo{ I then go on to give a more detailed presentation of the unifying property clause in P1(1). The case law from the ECtHR is presented and analysed in some depth, in an effort to assess how the ECtHR would be likely to approach an economic development case such as {\it Kelo}. In particular, I argue that the proportionality doctrine offers an interesting approach to such cases. Importantly, the doctrine stipulates that a ``fair balance'' must be struck  between the interests of the property owner and the public.\footcite[Chapter 5]{allen05} I argue that such a perspective could make it easier to get to the heart of why economic development takings are often seen as problematic, without getting lost in theoretical discussions about the meaning of  terms like ``public use'' or ``public purpose''. However, I also raise the concern that the ECtHR is not the appropriate institution for applying the proportionality test. Indeed, its remoteness to most of Europe suggests that we should look for more locally grounded legitimacy-enhancing institutions. Such institutions will likely be better able to assess the fairness of interference in context.

%I go on to discuss whether existing government institutions can serve this purpose, arguing that local courts may well be the best candidates. However, I also note that active application of the proportionality test in property cases might not be found at the local level. In this regard, the ECtHR could play a crucial role, by focusing on the systemic question of what issues local courts need to consider when assessing legitimacy of property interference. 

%However, quite apart from this, there is reason to worry that judicial bodies are not ideally suited to carry out the kind of assessment that is required. Hence, new institutional proposals might be in order. I conclude by arguing that once the need for local grounding is recognised and met, the ECtHR has the potential to play an important and constructive role in providing oversight and developing basic principles, also with respect to new institutions that aim to deliver increased legitimacy at the local level.

In England and Wales, the principle of parliamentary sovereignty and the lack of a written constitutional property clause has led to expropriation being discussed mostly as a matter of administrative law and property law, not as a constitutional issue.\footnote{See generally \cite{taggart98}.} Moreover, the use of compulsory purchase -- the term used to denote takings in the UK -- has not been restricted to particular purposes as a matter of principle.\footnote{See, e.g., \cite[48-49]{waring09}.} The uses that can justify taking property by compulsion are those uses that parliament regard as worthy of such consideration.\footnote{See \cite[48-49]{waring09}.} However, as private property has typically been held in high regard, the power of compulsory purchase has traditionally been exercised with caution.\footnote{See \cite[47-48]{waring09}.}

In his {\it Commentaries}, William Blackstone famously described property as the ``third absolute right'' that was ``inherent in every Englishman''.\footnote{See \cite[134-135]{blackstone79}. The first right, according to Blackstone, is security, while the second is liberty.} Moreover, Blackstone expressed a very restrictive view on the appropriateness of expropriation, pointing out that it was only the legislature that could legitimately interfere with property rights. He warned against the dangers of allowing private individuals, or even public tribunals, to be the judge of whether or not the common good could justify takings. Blackstone went as far as to say that the public good was ``in nothing more invested'' than the protection of private property.\footcite[134-135]{blackstone79}

In terms of historical accuracy, Blackstone's description conveys an impression of takings practice in England and Wales that is arguably only an idealisation. Specifically, it has been argued that his description of property might be shaped not so much by practical reality as by political values gaining ground among the bourgeoisie after the decline of the feudal system.\footnote{See \cite[34-35]{waring09} (describing Blackstone's account as a ``myth'').} However, the fact remains that compulsory purchase powers appear to have been granted relatively infrequently during his time, with no great increase in prevalence until the industrial revolution and the birth of the modern state.\footnote{See \cite[15]{allen00}. That said, recent scholarship has pointed out that expropriation appears to have taken place more frequently than previously thought, particularly following the glorious revolution, see \cite{hoppit11}.} Moreover, the conferral of such powers would typically require parliamentary involvement on a case-by-case basis, a practice reflecting that takings of private property, although far from unheard of, were indeed considered draconian.\footnote{See \cite[43-46]{nulty12}.}

Interestingly, the procedure followed by parliament in takings cases often resembled a judicial procedure; the interested parties were given an opportunity to present their case to parliament committees that would then effectively decide whether or not compulsion was warranted.\footnote{See \cite[13-16]{allen00}.} On the one hand, the direct involvement of parliament in the decision-making is suggestive of a fundamental respect for property rights. But at the same time, parliamentary sovereignty meant that the question of legitimacy was rendered mute as soon as compulsory purchase powers had been granted. The courts were not in a position to scrutinize takings at all, much less second-guess parliament as to whether or not a taking was for a legitimate purpose.\footnote{See, e.g., \cite[643]{nulty12}.}

During the 19th Century, as an industrial economy developed, so-called {\it private} acts, granting compulsory purchase powers to specific legal persons, grew massively in scope and importance.\footnote{See \cite[204]{allen00}.} Railway companies, in particular, regularly benefited from such acts.\footnote{\cite[204]{allen00}. See generally \cite{kostal97}.} During this time, the expanding scope of private-to-private transfers for economic development led to high-level political debate and controversy.\footnote{See \cite[204]{allen00}.} Usually, it would attract particular opposition from the House of Lords.\footcite[204]{allen00} Interestingly, this opposition was not only based on a desire to protect individual property owners. It also often reflected concerns about the cultural and social consequences of changed patterns of land use.\footcite[204]{allen00}

Hence, the early {\it political} debate on economic development takings in the UK shows some reflection of a social function approach to property protection. At the same time, as society changed following increasing industrialisation, a more expansive approach to compulsory purchase would eventually emerge as the norm.\footnote{Arguably, the social function perspective helps explain why this happened. Indeed, the expanded use of private takings in England during the 19th century, particularly in connection with the railways, might have served a more easily justifiable social function than that commonly associated with economic development takings today. Waring, in particular, notes how railway takings tended to affect aristocratic landowners rather than marginalised groups (``unlike private takings today, the railway legislation was most likely to affect those who could best defend their property rights from attack''), see \cite[111]{waring09}.} The idea that economic development could justify takings gradually became less controversial.

Today, the law on compulsory purchase in England is regulated in statute. Hence, parliament rarely gets involved on a case-by-case basis, and the role of the courts is largely limited to the application and interpretation of statutory rules.\footnote{See \cite[116-121]{waring09}. Some common law rules still play an important role, such as the {\it Pointe Gourde} rule, which stipulates that changes in value due to the compensation scheme itself should be disregarded when calculating compensation to the owner. The rule takes it name the case of \cite{gourde47}. The underlying principle, including also statutory regulations with a similar effect, is referred to as the ``no scheme'' principle, see \cite{lawcom01}. The principle is found in many jurisdictions, see \cite{sluysmans14}. It is often quite contentious, and notoriously hard to apply in practice. For a recent clarification of (some aspects of) the principle, see \cite{waters04}. I note that a strict interpretation of the no-scheme principle effectively precludes benefit sharing between takers and owners, a phenomenon that is also relevant in the context of economic development takings. See generally \cite{dyrkolbotn15}.} Moreover, with respect to the question of legitimacy of takings more broadly, the starting point for English courts is that this is a matter of ordinary administrative law.\footnote{See \cite{taggart98}.} More recently, the \cite{hra98} adds to this picture, since it incorporates the property clause in P1(1) into English law. Even so, the usual approach in England is to judge objections against compulsory purchase orders on the basis of the statutes that warrant them, rather than constitutional principles or human rights provisions that protect property.\footnote{See \cite[121-132]{waring09}. The important statutes are the \cite{ala81}, the \cite{lca61}, the \cite{cpa65}, the \cite{tcpa90} and the \cite{pcpa04}.} It is typical for statutory authorities to include standard reservations to the effect that some public benefit must be identified in order to justify a compulsory purchase order, but the scope of what constitutes a legitimate purpose can be very wide. For instance, to justify a taking under the \cite{tcpa90}, it is enough to argue that it will ``facilitate the carrying out of development, redevelopment or improvement on or in relation to the land''.\footcite[226]{tcpa90} 

While various governmental bodies are authorised to issue compulsory purchase orders (CPOs), a CPO typically has to be confirmed by a government minister.\footnote{See \cite[48]{waring09}.} The affected owners are given a chance to comment, and if there are objections, a public inquiry is typically held. The inspector responsible for the inquiry then reports to the relevant government minister, who makes the final decision about whether or not it should be granted, and on what terms. The CPO may then be challenged in court, but then on the basis of the statute authorising it, not on the basis of whether or not the purpose appears legitimate as such.\footnote{See, e.g., \cite[48-49]{waring09}. The typical way to launch an attack on a taking would be to argue that it serves a purpose that falls outside the scope of the statute authorising it, or, more subtly, that the administrative decision-maker took irrelevant purposes into account when granting the power. See, e.g., \cite{sainsbury10}.} 

That said, the idea that property may only be compulsorily acquired when the public stands to benefit permeates the system. Indeed, this has also been regarded as a constitutional principle, for instance by Lord Denning in {\it Prest v Secretary of State for Wales}.\footnote{See \cite[198]{prest82} (``I regard it as a principle of our constitutional law that no citizen is to be deprived of his land by any public authority against his will, unless it is expressly authorised by Parliament and the public interest decisively so demands.'').} Moreover, in {\it R v Secretary of State for Transport, ex p de Rothschild}, Slade LJ spoke of ``a warning that, in cases where a compulsory purchase order is under challenge, the draconian nature of the order will itself render it more vulnerable to successful challenge''.\footcite[938]{rothschild89}

In keeping with the principle of parliamentary sovereignty, this warning targets judicial review of administrative decision-making, not legislation. Despite this limitation, the English approach to legitimacy has traditionally proved quite effective in preventing controversy from arising with respect to the use of eminent domain.\footnote{See generally \cite{allen10}.} An underlying respect for private property, as well as the idea that the authority to interfere with it rests on the authority of parliament, appears to have influenced the decision-making framework and the surrounding administrative practices. Hence, legitimacy has become an objective to be pursued through legislation, regulation, and administrative practice, not judicial scrutiny.\footnote{For a more detailed analysis of how this works, noting, among other things, that higher levels within the executive are also meant to act as safeguards of private property, filling -- to some extent -- the possible role of courts in this regard, see \cite[85-100]{allen08}.}

%\begin{quote}
%It is clear that no minister or public authority can acquire any land compulsorily except the power to do so be given by Parliament: and Parliament only grants it, or should only grant it, when it is necessary in the public interest. In any case, therefore, where the scales are evenly balanced – for or against compulsory acquisition – the decision – by whomsoever it is made – should come down against compulsory acquisition. I regard it as a principle of our constitutional law that no citizen is to be deprived of his land by any public authority against his will, unless it is expressly authorised by Parliament and the public interest decisively so demands. If there is any reasonable doubt on the matter, the balance must be resolved in favour of the citizen.\footcite[198]{prest82}
%\end{quote}  
%Lord Denning also supported the doctrine of necessity, as expressed by Forbes J in {\it Brown v Secretary for the Environment}:\footcite{brown78}

%\begin{quote}It seems to me that there is a very long and respectable tradition for the view that an authority that seeks to dispossess a citizen of his land must do so by showing that it is necessary, in order to exercise the powers for the purposes of the Act under which the compulsory purchase order is made, that the acquiring authority should have authorisation to acquire the land in question.\footcite[291]{brown78}
%\end{quote}

%In practice, however, the public interest principle remains mostly implicit in legal reasoning, as a factor that influences the courts when they interpret statutory rules and carry out judicial review of administrative decisions. %As Watkins LJ stated in {\it Prest}:
%\begin{quote}
%The taking of a person's land against his will is a serious invasion of his proprietary rights. The use of statutory authority for the destruction of those rights requires to be most carefully scrutinised. The courts must be vigilant to see to it that that authority is not abused. It must not be used unless it is clear that the Secretary of State has allowed those rights to be violated by a decision based upon the right legal principles, adequate evidence and proper consideration of the factor which sways his mind into confirmation of the order sought.\footcite[211-212]{prest82}
%\end{quote}
%Plainly, England has not seen many cases that have proved as controversial or looked as offensive as the cases of economic development takings that have arisen in the US.\footnote{For a more thorough exploration of why this is so, also emphasising the importance of checks on administrative power, see generally \cite{allen08}.}

However, England and Wales have also seen controversial cases of takings for economic development. Indeed, such cases appear to have become more frequent.\footnote{See generally \cite{gray11}.} For instance, in the case of {\it Alliance}, many properties were taken in order to facilitate the construction of a new stadium for the football club Arsenal.\footcite{alliance06} Some owners who stood to lose their business premises protested on the basis that the purpose was dubious, pointing also to the fact that the inspector in charge of the public inquiry had recommended against the takings.\footcite[6-7]{alliance06} Their arguments also invoked P1(1) of the ECHR, to overcome the limitations of traditional judicial review in England and Wales. However, these argument were all quite summarily rejected by the Court.\footnote{See \footcite[6-7]{alliance06}. For a critical discussion, describing the Court's assessment against P1(1) as ``worryingly brief'', see \cite{gray11}.}

Arguably, the {\it Alliance} case reflects a weakness of the English approach to legitimacy. This weakness, moreover, appears to go beyond whatever doubts one might have about the principle of parliamentary sovereignty applied to property as a constitutional and/or human right. Specifically, if the framework laid down or condoned by parliament greatly empowers the administrative branch, while failing to appropriately regulate administrative practices, the deference due to parliament might effectively become undue deference to the executive branch. If the {\it practice} of using compulsory purchase continues to expand in relation to for-profit undertakings, there appears to be a significant risk of abuse associated with broad powers granted to the executive to take property for economic development. Plainly, if values such as those expressed by Blackstone are discredited further, there appears to be a lack of alternative sources for legitimacy in a system so reliant on a narrative of pure procedure.

To some extent, it would be possible for the Supreme Court to develop a more restrictive stance on compulsory purchase to address this, within the established constitutional order. In fact, there are some signs that this might be about to happen, specifically with respect to the broad powers granted under the the \cite[226]{tcpa90}. In the case of {\it R (Sainsbury's Supermarkets Ltd) v Wolverhampton City Council}, Lord Walker cited {\it Kelo} and went on to comment that ``economic regeneration brought about by urban redevelopment is no doubt a public good, but ``private to private'' acquisitions by compulsory purchase may also produce large profits for powerful business interests, and courts rightly regard them as particularly sensitive''.\footnote{See \cite[82]{sainsbury10}.}

However, the outcome of the {\it Sainsbury} case arguably also underscores the weaknesses of an indirect approach to legitimacy through administrative law. Instead of relying on Lord Walker's observations about the sensitivity of economic development takings, the majority of the Court quashed the compulsory purchase order on the basis that the local government had taken into account promises that the taker had made regarding a regeneration project in a different part of town. This was regarded as contravening section 226 of the \cite{tcpa90}, which only directs attention at the potential for improvements on or in relation to ``the land'', i.e., the land that is subject to compulsory purchase. The reasoning behind the decision, therefore, rests largely on a technicality, not any substantive assessment of legitimacy.

On a more purposive assessment, the taking in {\it Sainsbury} should arguably even have been upheld: the owner and the taker were both large commercial companies, they each owned a share of a plot of land suitable for joint development, they both wanted to develop at the expense of the other party, and the taker appeared to have the best overall plan for the community. Ironically, the English approach resulted in such a taking being struck down as illegitimate, while the taking in {\it Alliance}, involving the displacement of local people in favour of a football club, received little or no scrutiny at all. In light of this, it seems that alternatives to the traditional idea of legitimacy should be considered, at least if one agrees with Lord Walker's characterisation of economic development takings as ``particularly sensitive''.\footnote{See \cite[82]{sainsbury10}.}

\noo{ %This case nicely illustrates how English courts approach objections to the legitimacy of takings.% is found in the recent case of {\it Regina (Sainsbury’s Supermarkets Ltd) v Wolverhampton City Council}.\footcite{sainsbury10} 
%In this case, a CPO was granted to allow the company Tesco to acquire land from its competitor Sainsbury, in a situation when they were both competing for licenses to undertake commercial development on the same land, owned partly by both. The decisive factor that had led the local authorities to grant the CPO was that Tesco had offered to develop a different property in the same local area, which was currently in need of regeneration.

%Sainsbury protested, arguing that the local council could not strike such a deal on the use of its compulsory purchase power. It was argued, moreover, that taking the land for incidental benefits resulting from development in a different part of town was not legitimate under the Town and Country Planning Act 1990. The UK Supreme Court agreed 4-3, with Lord Walker in particular emphasising the need for heightened judicial scrutiny in cases of private-to-private takings for economic development.\footcite[80-84]{sainsbury10} Lord Walker even cited {\it Kelo}, to further substantiate the need for a stricter standard in such cases.\footcite[81]{sainsbury10} 

However, the main line of reasoning adopted by the majority was based on an interpretation of the Town and Country Planning Act itself. In particular, the majority held that it was improper for the local council to take into consideration the development that Tesco had committed itself to carry out on a different site.\footcite[73-79]{sainsbury10} This, in particular, was not ``improvement on or in relation to the land'', as required by the Act.\footcite[336]{tcpa90} In addition, Lord Collins, who led the majority, said that ``the question of what is a material (or relevant) consideration is a question of law, but the weight to be given to it is a matter for the decision maker''.\footcite[70]{sainsbury10} These comments reflect the traditional approach to judicial review of CPOs under English law, demonstrating how the underlying statutory authority tends to be at the center of attention.

However, it is interesting to see how the purpose of the interference featured in the background of the Supreme Court's interpretation and application of the statutory rule. The opinion of Lord Walker is particularly interesting, since he stresses that ``the land is to end up, not in public ownership and used for public purposes, but in private ownership and used for a variety of purposes, mainly retail and residential.''\footcite[81]{sainsbury10} He goes on to state that ``economic regeneration brought about by urban redevelopment is no doubt a public good, but ``private to private'' acquisitions by compulsory purchase may also produce large profits for powerful business interests, and courts rightly regard them as particularly sensitive.``\footcite[81]{sainsbury10}

Lord Walker then makes clear that he does not think it is impermissible, as such, for the local council to take into account positive effects on the local area, even when these do not directly result from the planned use of the land that is being acquired. Instead, he relies explicitly on the for-profit character of the taking, by arguing that ``the exercise of powers of compulsory acquisition, especially in a ``private to private'' acquisition, amounts to a serious invasion of the current owner's proprietary rights. The local authority has a direct financial interest in the matter, and not merely a general interest (as local planning authority) in the betterment and well-being of its area. A stricter approach is therefore called for.''\footcite[84]{sainsbury10} 

Lord Walker's opinion might indicate that the narrative of economic development takings is about to find its way into English case law. Moreover, a more critical approach might be adopted in the future, when compulsory purchase powers are made available to commercial companies wishing to undertake for-profit schemes. However, for schemes where the commercial aspect appears less dominant, English courts still appear very reluctant to quash CPOs, also when the purpose is economic development. This is so even in situations when the owners have requested a stricter standard of review on the basis of human rights law.\footnote{See \cite{smith08,alliance06}. See also the commentary in \cite[]{gray11}.} }

\noo{ \subsection{Some Other Cases from England} %{\it Smith \& Others v Secretary of Stare for Trade and Industry}}

In the case of {\it Smith \& Others v Secretary of State for Trade and Industry}, a caravan site was compulsorily acquired for development in connection with the London Olympic Games.\footcite{smith08} Some of the owners protested, including Romany Gypsies who used the caravans as their primary residence. A public inquiry was held, after which the inspector recommended that the CPO should not be confirmed until adequate relocation sites had been identified. However, due to the ``urgency, timing and importance'' of the project, the Secretary of State decided to go ahead before a relocation scheme was put in place (although he expressed commitment to ensuring satisfactory relocation).\footcite[10]{smith08} The owners argued that without satisfactory relocation plans, the interference in the property rights was not proportional and had to be struck down on the basis of human rights law, in particular Article 8 in the ECHR regarding respect for the home and private life.\footcite[27-51]{smith08}

The Court of Appeal considered the matter in great depth, applying the doctrine of proportionality developed at the ECtHR. Importantly, this doctrine was understood to go beyond the standard form of judicial review required under English law. However, the Court still concluded that the taking was proportional. This was largely based on the finding that ``the issue of proportionality has to be judged against the background that everyone accepts that an overwhelming case has been made out for compulsory acquisition of the sites for the stated objectives and that compulsory purchase is justified.''\footcite[42]{smith08} 

Justice Williams arrived at this conclusion after noting that the owners' {\it only} substantial objection against the CPO was that it was confirmed before adequate relocation measures had been agreed on.\footcite[42]{smith08} Hence, the question, as he saw it, did not concern the validity of using compulsory purchase powers, but merely the timing with which it had been ordered. On this basis, he framed the question of legitimacy as one relating to the ``necessity'' standard, according to which an infringement of Convention rights is only permissible when the public interest cannot be served in some other way.\footcite[43]{smith08} A strict reading of this standard holds that an interference must be the {\it least intrusive means} of achieving the stated aim.\footnote{Such a standard has been adopted in some Convention cases, for instance in \cite{samaroo01}.}

Justice Williams argued against such a strict reading, subscribing instead to a view expressed as an {\it obiter} in the case of {\it Pascoe v The First Secretary of State}. According to this view, an interference need not be the least intrusive means. Rather, it is sufficient that the measure is ``reasonably necessary'' to achieve that aim.\footnote{See \cite[74-75]{pascoe06} (quoting \cite[25]{clay04}).} However, while noting his agreement with this approach, Justice Williams went on to also apply the stronger necessity test, and found that even if this was applied the CPO in question would still be a proportional interference.\footcite[41-50]{smith08}

It seems clear that while the taking in question was for economic and recreational development purposes, the case was marked by a preliminary finding to the effect that the legitimacy of the aim of interference -- to facilitate the London Olympics -- was beyond reproach. Hence, there was no need for, or even room for, more detailed purposive reasoning of the kind that would later be applied by Lord Walker in {\it Sainsbury}. The fact that the taking was for economic development and recreation, not for a pressing public need, was not considered relevant. Moreover, since the case was construed to be solely about the extent to which the CPO was ``necessary'' to further its stated aim, the proportionality test that was carried out, despite being detailed, was very narrow in scope. It concerned only proportionality of the means, not of the aim itself. The question of how to weigh the public interest in a multi-billion dollar sporting event against the security of someone's home was not considered.

In later cases, a dismissive attitude towards substantive review has been adopted even in situations when the owners have argued against takings by explicitly questioning the proportionality of the interference against the importance of the aim. 

%\subsection{{\it Alliance Spring Co Ltd v The First Secretary of State}}

In the case of {\it Alliance Spring Co Ltd v The First Secretary of State}, a large number of properties were expropriated to build a new football stadium for the football club Arsenal.\footcite{alliance06} Some owners who stood to lose their business premises protested, pointing to the fact that the inspector in charge of the public inquiry had recommended against the takings.\footcite[6-7]{alliance06} According to Justice Collins, the main argument that the owners relied on when protesting the taking was that it did not serve a ``proper purpose''.\footcite[19]{alliance06} This argument was not held to be valid, however, with Justice Collins concluding as follows: 

\begin{quote}
There is nothing in the material put before and accepted by the Inspector which persuades me that that decision was ill founded or was one which the Secretary of State was not entitled to reach. Developments which result in regeneration of an area are often led by private enterprise. Mr Horton perforce accepts that that is so, but submits that this is not the sort of situation where, for example, a private development is the anchor for a particular scheme. I disagree.\footcite[19]{alliance06}
\end{quote}

Hence, unlike the case of {\it Smith}, where the Court did in fact carry out its own assessment of proportionality, the {\it Alliance} Court was content with deferring to the assessment carried out by the executive branch.\footnote{This has been criticized, e.g., by Kevin Grey who describes the reference to Convention Rights in Alliance as ``worryingly brief''. See \cite{gray11}.} As such, the case appears to follow the pattern of judicial review of CPOs established before the Human Rights Act 1998. This means that the decision also contrasts with how English courts have approach the Convention in relation to other  rights, such as those of Article 8 addressed in {\it Smith}.

Whether the approach taken in {\it Alliance} is good law after {\it Sainsbury} is unclear; judging from Lord Walker's opinion, it seems that a more substantive assessment might be required for similar cases in the future. While this might not imply a different outcome for a case like {\it Alliance}, it would mean that courts would have to engage in independent review of the purpose and merits of contested CPOs that benefit commercial actors. In particular, English courts would have to change the way they approach such cases, by being better prepared to assess for themselves whether a fair balance is struck between the interests of the developer and the property owners. Hence, it is not unlikely that the category of economic development takings will become an important point of reference in the future, both for the law and those who study it. }

\noo{ \subsection{Germany}\label{sec:germany}

In German law we find an explicit constitutional property clause. In particular, Article 14 of the Basic Law ({\it Grundgesetz}) reads as follows:

\begin{quote}
(1) Property and the right of inheritance shall be guaranteed. Their content and limits shall be defined by the laws. \\
(2) Property entails obligations. Its use shall also serve the public good. \\
(3) Expropriation shall only be permissible for the public good. It may only be ordered by or pursuant to a law that determines the nature and extent of compensation. Such compensation shall be determined by establishing an equitable balance between the public interest and the interests of those affected. In case of dispute concerning the amount of compensation, recourse may be had to the ordinary courts.\footcite[14]{basic49}
\end{quote}

Apart from the fact that the property clause is explicit, I note two further characteristic features of the protection of property in Germany. First, the constitution explicitly stresses that property comes with social obligations as well as rights. The use of property should ``serve the public good''. On the other hand, it is also made clear that expropriation is only permissible when it is ``for the public good''. Hence, it follows immediately that the purpose of expropriation is a relevant factor when determining the legitimacy of a taking, \isr{irrespective} of the specific statute used to authorise it. Importantly, it is clear already from the outset that the question of legitimacy is a \emph{judicial} question, one which the courts can only answer if they form an opinion about that constitutes the ``public good''. 

This means that it is quite natural to approach the question of economic development takings from the point of view of constitutional law. Unlike in England, disputes over the legitimacy of such takings can be comfortably adjudicated directly against a ``public good'' restriction. While this sets Germany apart on the theoretical level, it is unclear how much of an effect it has had in practice. To shed some light on this question, we can look to the two major authorities on the legitimacy of economic development takings, the cases of {\it D\"{u}rkheimer Gondelbahn} and {\it Boxberg}.\footcite{durkheimer81,boxberg86} 

In both cases, the German Constitutional court found that expropriation to the benefit of commercial interests was illegitimate. However, the Court argued for this result on the basis that there was insufficient statutory authority for such takings in the concrete circumstances complained of. That is, the Court did not directly address the question of whether the relevant statutes were compliant with Article 14 of the basic law. Instead, they interpreted statutory authorities on the assumption that they had to be, following a pattern of reasoning that appears to be rather close to the approach followed by English courts in similar cases.\footnote{Although in {\it Dürkheimer Gondelbahn}, Böhmer J gave a separate concurring judgment where he argued for this result on the basis of the public good requirement of the basic law.} It seems, in particular, that even in Germany, the public purpose restriction is primarily relevant as a factor guiding the interpretation of statutory authorities.

That said, the cases of {\it D{\"u}rkheimer Gondelbahn} and {\it Boxberg} show that in situations when the public purpose of a taking is unclear, German courts seem inclined to \isr{favour} a narrow interpretation of the relevant statute. In {\it Bloxberg}, several properties were expropriated \isr{in favour} of the car company Daimler Benz AG, for commercial purposes. The affected local communities suffered from high unemployment rates and a slow economy, so a {\it prima facie} reasonable \isr{case} could be made that allowing Daimler to acquire the land was in the public interest, as it would facilitate economic growth. However, the Federal Constitutional Court agreed with the owners that the expropriation was invalid. This, it held, was because the taking was outside the scope of the relevant statute, which authorised expropriations for ``planning purposes''. The owners had argued extensively using Article 14 of the Basic Law and the constitutional ``public good'' restriction clearly did play a role in the Court's reasoning. But at the same time, the Court stressed that private-to-private transfers that bestow financial benefit on the acquiring party may well satisfy the ``public good'' requirement. The important issue was whether a sufficiently strong public interest could be identified, \isr{irrespective} of any windfall benefits that might fall on private parties.

In light of this, I think it is wrong to exaggerate the importance of the explicit formulation of the public use test offered in the German constitution. Its importance seems to rest mainly in the fact that it provides a particularly authoritative expression guiding the national courts' application of statutory provisions regarding expropriation of property. But developments in common law, where the public use requirement is stressed as a guiding constitutional principle, might well point in the same direction. In principle, both German and English Courts are in a good position to respond to increased tension regarding economic development takings by developing a stricter standard of judicial review in such cases.

A different aspect of German law deserves special attention, however, since it does not appear to have any clear counterpart in the common law tradition. This is the  ``social-obligation'' norm in Article 14 (2), which points to a different \isr{conceptualisation} of property rights as such. As argued by Alexander, the distinguishing feature of the property clause in the German Constitution is that the value of property is thought to relate more strongly to its importance for human dignity and flourishing in a social context, rather than the protection of individual financial entitlements. As Alexander notes regarding the Germans' own \isr{conceptualisation} of their property clause:

\begin{quote}
This theory holds that the core purpose of property is not wealth maximization or the satisfaction of individual preferences, as the American economic theory of property holds, but self-realization, or self-development, in an objective, distinctly moral and civic sense. That is, property is fundamental insofar as it is necessary for individuals to develop fully both
as moral agents and participating members of the broader community.\footcite[745]{alexander03}
\end{quote}

With such a starting point, it is not surprising that in cases such as {\it Boxberg}, resembling {\it Kelo}, German Courts will tend to adopt a strict view on legitimacy. These are cases when the property rights infringed on serve a fundamentally different function for the two opposing private parties. To the owner, the property is a home, an important source of self-identity, autonomy, security and membership in a community. To the taker, it represents an obstacle to commercial development which needs to be removed. In such a situation, it is in keeping with the spirit of the social-obligation norm of property to offer enhanced protection to the homeowner. To this owner, the property serves a purpose which is fundamentally different, and arguably more worthy of protection, then the property's purpose for the developer. A taking in this situation might therefore, because of Article 14, require a particularly clear and strong public interest.

But unless there is an asymmetry between owner and taker, heightened scrutiny does not necessarily follow. Hence, it is interesting to speculate what German courts would have made of a case such as {\it Regina (Sainsbury’s Supermarkets Ltd) v Wolverhampton City Council}. Here, the interests of owner and taker were strictly commercial nature. Both owned part of the contested land and neither one could develop the land according to their plans without buying out the other. The enhanced protection of property offered under German law would probably not have much significance in such a case. 

In fact, it might well be that German courts would be {\it more} likely to accept such a taking. First, their \isr{conceptualisation} of property rights appears to allow greater flexibility to adapt the level of protection to the circumstances and the purposes of the property in question. So even if is correct that private-to-private transfers for commercial projects require a ``stricter approach'' in general, as argued by Lord Walker in \textcite{sainsbury10}, the fact that the interests of the owner were also purely commercial  might make this less relevant. Second, German courts might be more inclined to have regard to socially beneficial additional commitments entered into by the applicant, even if they do not concern the property that is taken. As a tie-breaker, looking to such commitments might be as good an approach as any other.\footnote{This was the view taken by the dissenting minority in \textcite{sainsbury10}.}

Of course, objections could still be raised on the basis of general administrative law. Indeed, some might see the case as an example of government ``auctioning'' off licenses to the highest bidder. This might well be regarded as an affront to good governance. I will not delve into German law to assess the case from this perspective. My point is simply that because of the purposive and contextual nature of Article 14, it seems unlikely that a case like \textcite{sainsbury10} would turn on constitutional property law.

To sum up, German constitutional law serves to create an interesting contrast with English law regarding the question of economic development takings. On the one hand, property appears to be better protected against such takings in Germany, but on the other hand, the extent to which increased protection is offered depends more closely on the social values involved. The German system appears to look more actively at the social function of property for guidance when resolving property disputes, thereby echoing some of the ideas discussed in Chapter \ref{chap:1}. 

In the next section, I will discuss the property clause in the ECHR, which explicitly serves to set up a minimum level of property protection that provides a common standard for all member states, including Germany and the UK.
}

\section{The US: Legitimacy through Public Use}\label{sec:us}

\noo{In this section, I consider US law in more depth. First, I track the development of the case law on the public use restriction found in the Fifth Amendment and in various state constitutions. I consider the jurisprudential development from the early 19th century up to the present day.\footnote{The public use clause in the US constitution was not held to apply to state takings until the late 19th century, see \cite{chicago97}.} Many writers assert that case law from the 19th and early 20th century was \isr{characterised} by a tension between `narrow' and `broad' readings of the notion of public use.\footnote{See, e.g., \cite[483]{walt11}; \cite[203-204]{allen00}. For a more in-depth argument asserting the same, see \cite{nichols40}.} Adding to this, I argue that while different state courts expressed different theoretical views on the meaning of ``public use'', there was a growing consensus that the approach to judicial scrutiny should be contextual, focused on weighing the rationale of the taking against the social, political and economic circumstances.\footnote{A summary of state case law that supports this view is given in the little discussed Supreme Court case of \cite{hairston08}.}  In particular, early state courts did not focus unduly on the exact wording of constitutional property clauses.

Following up on this, I argue that the doctrine of deference that was developed by the Supreme Court early in the 20th century was directed primarily at state courts, not state legislatures and administrative bodies.\footnote{See \cite{vester30} (echoing and citing \cite{hairston08}).} I then present the case of {\it Berman}, arguing that it was a significant departure from previous case law.\footcite{berman54} After {\it Berman}, deference was now taken to mean deference to the (state) legislature, meaning that there would be little or no room for judicial review of the takings purpose. 

This paved the way for the infamous case of {\it Poletown}, where a \isr{neighbourhood} of about 1000 homes was razed in order to provide General Motors with land to build a car factory.\footnote{See \cite{poletown81}.} I note how {\it Berman} provided a key authority used by the state court to uphold this taking. {\it Poletown} in turn links up with the even greater controversy surrounding {\it Kelo},  the eventual backlash of the deferential stance introduced in {\it Berman}.

After the historical overview, I go on to briefly present the vast amount of research that has targeted economic takings in the US after {\it Kelo}. I devote special attention to a proposal due to Heller and Hills, who propose a new institution that can replace eminent domain as a mechanism for land assembly in case of economic development.\footcite{heller08} This proposal will serve as important reference point later on, when I consider the Norwegian land consolidation courts in Chapters 6.

\subsection{The History of the Public Use Restriction}\label{sec:hop}
}

Unlike England and Wales, the US Constitution is a basis for substantive judicial review also of the actions taken by the federal and state legislatures. However, considering its status as a basis for potentially extensive review, the Constitution is remarkably terse. The takings clause, arriving mid-sentence through the fifth amendment, reads simply: ``nor shall private property be taken for public use, without just compensation''.\footnote{See \cite{us}.}

The compensation requirement is clearly stated, but the takings clause is also understood to include the requirement that property may only be taken for ``public use''. This is the aspect of the clause that will interest me in this thesis, since it provides an anchor for legitimacy that is particularly relevant -- and contentious -- in relation to economic development takings.\footnote{The compensation requirement is also important, of course, but the problems is gives rise to are rather more technical, pertaining also more to the entitlements-aspect of property protection, not the social function dimensions I focus on in this thesis. For a more in-depth assessment of the compensation issue in the context of economic development takings, see \cite{dyrkolbotn15}.}
Specifically, the question is to what extent such takings offend against the clause: is a taking for economic development by a commercial company really a taking for ``public use''?

Going back to the time when the fifth amendment was introduced, there is not much historical evidence explaining why the takings clause was included in the Bill of Rights.\footnote{See \cite{fifth}.} Moreover, there is little in the way of guidance as to how the takings clause was originally understood. James Madison, who drafted it, commented that his proposals for constitutional amendments were intended to be uncontroversial.\footnote{See letters from Madison to Edmund Randolph dated 15 June 1789 and from Madison to Thomas Jefferson dated 20 June 1789, both included in \cite{madison79}.} Hence, it is natural to regard the takings clause as a codification of an existing principle, not a novel proposal. Indeed, several state constitutions pre-dating the Bill of Rights also included takings clauses, seemingly based on codifying principles from English common law.\footcite[See][299]{johnson11} As Meidinger notes, the Americans had never really charged the British with abuse of eminent domain, and private property had tended to be respected, also in the colonies.\footcite[17]{meidinger80} This undoubtedly influenced early US law.

Just like English scholars at the time, early American scholars emphasised the importance of private property. James Kent, for instance, described the sense of property as ``graciously implanted in the human breast'' and declared that the right of acquisition ``ought to be sacredly protected''.\footnote{See \cite[see][257]{kent27}.} Indeed, the Supreme Court itself expressed similar sentiments early on, when it spoke of the impossibility of passing a law that ``takes property from A and gives it to B''.\footnote{This was a {\it de dicta} in \cite[388]{calder98}. See also \cite[310]{vanhorne95}.}

However, just as in England, this early US attitude changed in response to industrial advances and a desire for economic development. As the 19th century progressed, eminent domain was used more frequently, now also to benefit (privately operated) railroad operations, hydroelectric projects, and the mining industry.\footcite[23-33]{meidinger80} During this time, it also became increasingly common for landowners to challenge the legitimacy of takings in court, undoubtedly a consequence of the fact that eminent domain was used more widely, for new kinds of projects.\footcite[24]{meidinger80} Controversy over the public use requirement arose particularly often with respect to the so-called mill acts.\footnote{\cite[24]{meidinger80}. See also \cite[306-313]{johnson11} and \cite[251-252]{horwitz73}.} Such acts were found throughout the US, many of them dating from pre-industrial times when mills were primarily used to serve the farming needs of agrarian communities.\footnote{A total of 29 states had passed mill acts, with 27 still in force, when a list of such acts was compiled in \cite[17]{head85}. According to Justice Gray, at pages 18-19 in the same, the ``principal objects'' for early mill acts had been grist mills typically serving local agrarian needs at tolls fixed by law, a purpose which was generally accepted to ensure that they were for public use.} Following economic and technological advances, provisions originally enacted to serve local farming purposes were now being used by developers wishing to harness hydropower for manufacturing and hydroelectric plants.\footnote{See, e.g., \cite[18-21]{head85} and \cite[449-452]{minn06}.}

It is important to note, however, that mill acts could not be used to authorise large-scale compulsory transfer of natural resources from owners to non-owners.\footnote{See the discussion in \cite{head85} (a mill case adjudicated by the Supreme Court, including a summary of mill acts and case law from several different states).} Rather, mill acts provided management tools that could be used to ensure that owners of water resources could make better use of their rights. This would sometimes involve allowing riparian owners to interfere with, or take a necessary part of, the property of their neighbours, e.g., by constructing dams that would flood neighbouring land.\footnote{See, e.g., \cite[265]{staples03}.} However, the primary purpose of most mill acts was to facilitate rational coordination among owners, to the benefit of their community as a whole. This point was frequently made by the courts to justify upholding takings on the basis of mill acts, including takings that would benefit the manufacturing industry.\footnote{See \cite{fiske31}. See also the discussion (including references to other cases) in \cite{head85}.}

As the industrial use of mill acts increased in scope, the original aim of these acts gradually became overshadowed by the strength of the commercial interests involved. This, in turn, lead to public use controversies arising in relation to provisions that had not previously raised any doubts.\footnote{See \cite{head86}.} This mechanism arose from the changing social and economic context of mill takings, underscoring therefore the appropriateness of adopting a social function perspective. Indeed, the case law on public use from state courts at this time was characterised by a highly contextual understanding of property protection and the meaning of public use.\footnote{See, e.g, \cite{scudder32} (taking upheld, but said that ``the great principle remains that there must be a public use or benefit. That is indispensable. But what that shall consist of, or how extensive it shall be to authorize an appropriation of private property, is not easily reducible to a general rule.''); \footcite[409]{seawell76} (taking for a mineral company upheld on the basis that mining was the ``greatest of the industrial pursuits'' in the state of Nevada and that the benefits of the industry were ``distributed as much, and sometimes more, among the laboring classes than with the owners of the mines and mills''.); \footcite[337]{ryerson77} (taking struck down, by a Court that was ``not disposed to say that incidental benefit to the public could not under any circumstances justify an exercise of the right of eminent domain''.\footcite[337]{ryerson77}. See also \cite{gray11} (with many references to state courts striking down takings as impermissible).} This legacy from the states was also recognised by the early Supreme Court jurisprudence on economic development takings, as discussed in the next section.

\noo{In the following, I explore this in some further detail.

\subsection{Legitimacy in State Courts}\label{subsec:state}

When considering objections to the legitimacy of takings, state courts would not look to the federal Takings Clause directly, but rather base their decisions on corresponding property clauses from their own respective state constitutions.\footnote{Not all states had such property clauses, and exact formulations varied, but a public use requirement was typically observed, see \cite[293-296]{johnson11}.} Indeed, it was not until the late 19th century that state takings came to be regularly scrutinized at the federal level.\footnote{At first, the federal scrutiny took place on the basis of the due process clause in the Fourteenth Amendment, see \cite{head75}. The federal takings clause itself was only applied to state takings after 1897, see \cite{chicago97}.}

When a state court upheld an interference that would benefit commercial interests, it would typically emphasise the broader purpose, often focusing on economic ripple effects.\footnote{See, e.g., \cite{hazen53,scudder32,boston32}. A more comprehensive list of cases adopting a broad view can be found in \cite[617]{nichols40}.} By contrast, when a court decided that an interference was unconstitutional, it would often focus on the concrete use made of the property that was taken,  pointing out that it did not directly benefit the public in the sense required by the public use restriction.\footnote{See, e.g., \cite{sadler59,ryerson77,gaylord03,minn06}. A more comprehensive list can be found in \cite{alr28}.} Sometimes, the question of legitimacy would turn on how widely the notion of `use' was understood. Should this notion be interpreted narrowly, as requiring that the property had to be literally used by the public, or could it be understood broadly, as pointing to a public purpose or benefit of some sort?\footnote{According to Nichols, the narrow view emerged as the ``majority'' opinion on public use, see \footcite[617-618]{nichols40}. But contrast this with \cite{berger78} and \cite[24]{meidinger80}, who argue that the narrow view was only dominant in a handful of states, led by New York.}

This tension between broad and narrow readings of the public use clause have received much attention from legal scholarship.\footnote{See \cite{nichols40,berger78,meidinger80,johnson11}.} However, when studying the case law in more depth, a complementary picture emerges, testifying to some cohesion in the states'
jurisprudence. Regardless of their reading of the public use requirement, state courts seem to have agreed that the question of what counted as a public use was a judicial question that should be assessed concretely, not abstractly.

%For instance, in the case of {\it Gaylord v. Sanitary Dist. of Chicago}, the Supreme Court of Illinois held the state Mill Act to be unconstitutional, as it was not limited to traditional flour mills. In doing so, the court observed that public use was ``something more than a mere benefit to the public''.\footcite[524]{gaylord03} Similar sentiments were expressed in other decisions striking down uses of eminent domain for mill construction, for instance in Vermont, Michigan and New York.\footnote{References.}

A good example is the case of {\it Dayton Gold \& Silver Mining Co v Seawell}, concerning an act that gave mineral owners a right to acquire additional rights needed to facilitate extraction.\footcite{seawell76} The Supreme Court of Nevada decided that the act was constitutional on the basis of a highly contextual reading of the public use requirement in the property clause of the Constitution of Nevada. Interestingly, the Court argued against a literal (narrow) reading on the basis that such a reading would ultimately provide {\it weaker} protection of property:

\begin{quote}
If public occupation and enjoyment of the object for which land is to be condemned furnishes the only and true test for the right of eminent domain, then the legislature would certainly have the constitutional authority to condemn the lands of any private citizen for the purpose of building hotels and theaters. [...] Stage coaches and city hacks would also be proper objects for the legislature to make provision for, for these vehicles can, at any time, be used by the public upon paying a stipulated compensation. It is certain that this view, if literally carried out to the utmost extent, would lead to very absurd results, if it did not entirely destroy the security of the private rights of individuals. Now while it may be admitted that hotels, theaters, stage coaches, and city hacks, are a benefit to the public, it does not, by any means, necessarily follow that the right of eminent domain can be exercised in their favor.\footcite[410-411]{seawell76}
\end{quote}

The quote presents an argument in favour of a broad understanding of the public use requirement. However, it also prescribes broad judicial review of takings purposes, including purposes that would appear to pass a `narrow' public use test. In this way, it asks us to  resist the temptation to think that a broad understanding of public use necessarily entails a public use test that can be passed more easily.

The Court follows up on its reading of the public use requirement by giving a highly contextual assessment of the takings purpose. Specifically, it considers the social and economic importance of mining, concluding that it is the ``greatest of the industrial pursuits'' and that all other interests are ``subservient'' to it.\footcite[409]{seawell76} Indeed, the Court goes as far as to conclude that the benefits of the mining industry are ``distributed as much, and sometimes more, among the laboring classes than with the owners of the mines and mills''.\footcite[409]{seawell76} On this basis, the Court upholds the taking.

I am agnostic as to whether or not this decision was based on an accurate description of the mining industry in Nevada in the late 19th century. The importance of the decision and the remarks above does not turn on this factual question. Rather, the importance arises from the fact that the Court felt the need to scrutinize the takings purpose very broadly. The issue of legitimacy was not approached as a linguistic exercise or an attempt at recreating the original intent of the relevant property clause. Instead, the court proceeded on the basis of their assessment of the prevailing social and economic conditions in the state of Nevada. 

The Court noted the importance of deference to the legislature on matters of policy, but qualified this by remarking that any authority to take property had to be ``enforced by the courts so as to prevent its being used as an instrument of oppression to any one''.\footcite[412]{seawell76} 
Furthermore, the Court was convinced that its contextual approach in this regard would generally offer {\it increased} protection of private property compared to more formalistic approaches. The Court summarised its view on this as follows:

\begin{quote}
Each case when presented must stand or fall upon its own merits, or want of merits. But the danger of an improper invasion of private rights is not, in my judgment, as great by following the construction we have given to the constitution as by a strict adherence to the principles contended for by respondent.\footcite[398]{seawell76}
\end{quote}

The {\it Seawell} case is not unique. For another example, I mention {\it Ryerson v Brown}, a case often cited as an authority for a narrow view of public use.\footcite{ryerson77} Here the taking in question was held to be unconstitutional. However, the Supreme Court of Michigan qualified also made clear that it was ``not disposed to say that incidental benefit to the public could not under any circumstances justify an exercise of the right of eminent domain''.\footcite[337]{ryerson77}

The case concerned the constitutionality of a taking under a mill act, and while the court argues that public use should be taken to mean ``use in fact'', it is clear that ``use'' is understood rather loosely, not literally as physical use of the property that is taken.\footnote{The court explains its stance on the public use restriction by stating (emphasis added) ``it would be essential that the statute should require the use to be public in fact; in other words, that it should contain provisions entitling the public to {\it accommodations}.'' The court continues with an illustrative example: ``A flouring mill in this state may grind exclusively the wheat of Wisconsin, and sell the product exclusively in Europe; and it is manifest that in such a case the proprietor can have no valid claim to the interposition of the law to compel his \isr{neighbour} to sell a business site to him, any more than could the manufacturer of shoes or the retailer of groceries. Indeed the two last named would have far higher claims, for they would subserve actual needs, while the former would at most only incidentally benefit the locality by furnishing employment and adding to the local trade''. See \cite[336]{ryerson77}.} Moreover, when clarifying its starting point for judicial scrutiny, the court explains that ``in considering whether any public policy is to be subserved by such statutes, it is important to consider the subject from the standpoint of each of the parties''. Following up on this, the court finds, with respect to the case in question, that ``the power to make compulsory appropriation, if admitted, might be exercised under circumstances when the general voice of the people immediately concerned would condemn it''. On thi basis, the Court strikes down the taking, summing up its factual assessment as follows: ``what seems conclusive to our minds is the fact that the questions involved are questions not of necessity, but of profit and relative convenience''.\footcite[336]{ryerson77}

Hence, far from nitpicking on the basis of the public use phrase, the court adopts a contextual approach to takings that is rather similar to the approach of {\it Dayton Gold \& Silver Mining Co. v. Seawell}. The outcome \isr{is} different, but it is also based on a different assessment of the context and the consequences of the takings complained about. Importantly, the case does not rest on any {\it a priori} assumption that economic development takings of the kind in question could not meet a public use test -- no general rule is relied on at all. Hence, it is somewhat strange that later commentators have focused on the case for its ambiguous comments on public use as ``public in fact'' rather than its broad and well-reasoned assessment of actual legitimacy.\footnote{See, for instance, Justice Thomas' dissent in {\it Kelo}, \cite[513]{kelo05} (using {\it Ryerson} as a reference to support an `actual use' interpretation of the public use requirement in the fifth amendment).}

Many of the important cases from the late 19th century, on both sides of the public use debate, share crucial features with the two cases discussed above.\footnote{See, e.g., \cite{scudder32} (Eminent domain power upheld, but said: ``The great principle remains that there must be a public use or benefit. That is indispensable. But what that shall consist of, or how extensive it shall be to authorize an appropriation of private property, is not easily reducible to a general rule. What may be considered a public use may depend somewhat on the situation and wants of the community for the time being.''), \cite{fallsburg03} (Eminent domain struck down, on holding that ``the private benefit too clearly dominates the public interest to find constitutional authority for the exercise of the power of eminent domain''), \cite[538]{board91} (Eminent domain struck down, qualified by ``not only must the purpose be one in which the public has an interest, but the state must have a voice in the manner in which the public may avail itself of that use'').} Hence, a shared trait appears to have emerged among state courts during this period, namely a willingness to engage in broad judicial scrutiny of the legitimacy of economic development takings. Indeed, state courts appear to have been conscious of the special legitimacy questions that arise when eminent domain is used to facilitate economic development through commercial enterprise. The question of how to understand public use terminology was an important part of this, but it was not considered in isolation from other aspects.

This observation is relevant when considering the takings doctrine that later developed at the federal level. In particular, the broad scrutiny offered by state courts suggests that the doctrine of extreme deference that was about to be adopted by the Supreme Court resulted from a completely new development, not a continuous broadening of the public use requirement.\footnote{This contrasts with the argument given by the majority in {\it Kelo}, see \cite[479-480]{kelo05} (placing the doctrine of deference in a tradition emerging from how the narrow view of some early state courts ``steadily eroded'' because of the ``diverse and always evolving needs of society'').} 
}

\subsection{Legitimacy as Discussed by the Supreme Court}\label{subsec:US}

Initially, the Supreme Court held that the takings clause in the US Constitution did not apply to state takings at all.\footcite{barron33} Federal takings, on the other hand, were of limited practical significance since the common practice was that the federal government would rely on the states to condemn property on its behalf.\footcite[30]{meidinger80}

This changed towards the end of the 19th century, particularly following the decision in {\it Trombley v Humphrey}, where the Supreme Court of Michigan struck down a taking that would benefit the federal government.\footcite{trombley71} Not long after, in 1875, the first Supreme Court adjudication of a federal taking occurred, marking the start of the development of the federal doctrine on public use and legitimacy.\footcite{kohl75} 

At the same time, the Supreme Court began to hear takings cases originating from the states, first on the basis of the due process clause of the fourteenth amendment, introduced after the civil war.\footnote{See, e.g, \cite{head85}.} Later, in 1897, the Supreme Court held that state takings could be scrutinized also against the takings clause of the fifth amendment.\footnote{See \cite{chicago97}.}

\noo{ The early 20th century was a period of great optimism about the ability of {\it laissez faire} capitalism to ensure progress and economic growth, a sentiment that was reflected in the federal case law on eminent domain. A particularly clear expression of this can be found in {\it Mt Vernon-Woodberry Cotton Duck Co v Alabama Interstate Power Co}.\footcite{vernon16}  This case dealt with the legitimacy of condemnation arising from the construction of a hydropower plant. The Supreme Court held that it was legitimate, with the presiding judge arguing briskly as follows:

\begin{quote}The principal argument presented that is open here, is that the purpose of the condemnation is not a public one. The purpose of the Power Company's incorporation, and that for which it seeks to condemn property of the plaintiff in error, is to manufacture, supply, and sell to the public, power produced by water as a motive force. In the organic relations of modern society it may sometimes be hard to draw the line that is supposed to limit the authority of the legislature to exercise or delegate the power of eminent domain. But to gather the streams from waste and to draw from them energy, labor without brains, and so to save mankind from toil that it can be spared, is to supply what, next to intellect, is the very foundation of all our achievements and all our welfare. If that purpose is not public, we should be at a loss to say what is. The inadequacy of use by the general public as a universal test is established. The respect due to the judgment of the state would have great weight if there were a doubt. But there is none.\footcite[32]{vernon16}
\end{quote}

On the one hand, the Court notes the importance of deference to the {\it state} judgement (not specifically the judgement of the state legislature). On the other hand, it prefers to conclude on the basis of its own assessment of the purpose of the taking. This assessment, however, is not grounded in the facts of the case or the circumstances in Alabama. Rather, it is based on sweeping assertions about ``all our welfare'' and the desire to ``save mankind from toil that it can be spared''. This marks a contrast with the approach of state courts, as discussed in the previous subsection.
}

%The contrast was even greater in cases when the takings in question had been authorised by the federal government itself.
In federal takings cases, the Supreme Court showed little willingness to enforce a strict public use requirement. In {\it United States v Gettysburg Electric Railway Co}, a case from 1896, deference to the legislature in federal takings cases was referred to as a principle that should be observed unless the judgement of the legislature was ``palpably without reasonable foundation''.\footcite[680]{gettysburg96} 

Importantly, however, such a deferential stance was not adopted in cases originating from the states. In {\it Cincinatti v Vester}, a case from 1930, the Supreme Court commented that ``it is well established that, in considering the application of the Fourteenth Amendment to cases of expropriation of private property, the question what is a public use is a judicial one''.\footcite[447]{vester30} %In this judgement, Chief Justice Hughes also describes how the judicial assessment of the public use question should be carried out:

%\begin{quote}
%In deciding such a question, the Court has appropriate regard to the diversity of local conditions and considers with great respect legislative declarations and in particular the judgments of state courts as to the uses considered to be public in the light of local exigencies. But the question remains a judicial one which this Court must decide in performing its duty of enforcing the provisions of the Federal Constitution.\footcite[447]{vester30}
%\end{quote}

%Notice how this echoes the contextual approach developed at the state level, while explicitly prescribing deference to state {\it courts}. 
In the earlier case of {\it Hairston v Danville \& W R Co}, from 1908, the same was expressed by Justice Moody, who surveyed the state case law and declared that ``the one and only principle in which all courts seem to agree is that the nature of the uses, whether public or private, is ultimately a judicial question.''\footcite[606]{hairston08} Justice Moody continued by describing in more depth the typical approach of the state courts in determining public use cases:

\begin{quote}
The determination of this question by the courts has been influenced in the different states by considerations touching the resources, the capacity of the soil, the relative importance of industries to the general public welfare, and the long-established methods and habits of the people. In all these respects conditions vary so much in the states and territories of the Union that different results might well be expected.\footcite[606]{hairston08}
\end{quote}

Justice Moody goes on to give a long list of cases illustrating this aspect of state case law, showing how assessments of the public use issue had been inherently contextual.\footcite[607]{hairston08} %He then cites three further Supreme Court cases, pointing out that all of them express support for state case law on this issue.\footnote{{\it Falbrook, Clark} and {\it Strickley}.} 
Following up on this, he points out that ``no case is recalled'' in which the Supreme Court overturned ``a taking upheld by the state {\it court} as a taking for public uses in conformity with its laws'' (my emphasis). After making clear that situations might still arise where the Supreme Court would not follow state courts on the public use issue, Justice Moody goes on to conclude that the cases cited ``show how greatly we have deferred to the opinions of the state courts on this subject, which so closely concerns the welfare of their people''.\footcite[606]{hairston08}

{\it Hairston} is important for three reasons. First, it makes clear that initially, the deferential stance in cases dealing with state takings was primarily directed at state courts rather than legislatures and administrative bodies. Second, it demonstrates federal recognition of the fact that a consensus had emerged in the states, whereby scrutiny of the public use determination was consistently regarded as a judicial task.\footnote{Indeed, {\it Hariston} provides the authority for {\it Vester} on this point. See \cite[606]{vester30}.} Third, it provides a valuable summary of the contextual approach to the public use test that had developed at the state level. 

The {\it Hairston} Court clearly looked favourably on the case law from state courts. Importantly, when a deferential stance was adopted, this was clearly contingent on the assumption that state courts would continue to administer the public use test with the required vigour. Despite this, {\it Hairston} would later be cited as an early authority in favour of almost unconditional deference to legislators.\footnote{In fact, it was cited in this way also by the majority in {\it Kelo}, see \cite[482-483]{kelo05}.} 

This happened in {\it US ex rel Tenn Valley Authority v Welch}, concerning a federal taking.\footcite[552]{welch46} The Court first cited {\it US v Gettysburg Electric R Co} as an authority in favour of deference with regards to the public use limitation.\footcite{gettysburg96} The Court then paused to note that {\it Vester} later relied on the opposite view, namely that the public use test was a judicial responsibility.\footcite{vester30} The Court then attempts to undercut this by setting up a contrast between {\it Vester} and {\it Hairston}, by selectively quoting the observation made in the latter case that the Supreme Court had never overruled the state courts on the public use issue.\footnote{See \cite[552]{welch46}.} Hence, {\it Hairston} is effectively used to argue against judicial scrutiny, in a manner that is quite incommensurate with the full rationale behind the Court's decision in that case.

Later, {\it Welch} was used as an authority in the case of {\it Berman v Parker}.\footcite{berman54} This case concerned condemnation for redevelopment of a partly blighted residential area in the District of Colombia, which would also condemn a non-blighted department store. In a key passage, the Court states that the role of the judiciary in scrutinizing the public purpose of a taking is ``extremely narrow''.\footcite[32]{berman54} The Court provides only two references to previous cases to back up this claim, one of them being {\it Welch}.\footnote{The other case, {\it Old Dominion Land Co v US}, concerned a federal taking of land on which the military had already invested large sums in buildings. The Court commented on the public use test by saying that ``there is nothing shown in the intentions or transactions of subordinates that is sufficient to overcome the declaration by Congress of what it had in mind. Its decision is entitled to deference until it is shown to involve an impossibility. But the military purposes mentioned at least may have been entertained and they clearly were for a public use''. See \cite[66]{dominion25} A misleading and partial quote, to the effect that deference to the legislature is in order except when it involves an ``impossibility'', has since become commonplace. In particular, such a quote was repeated by the Supreme Court itself in the later case of \cite[240]{midkiff84}.}

Moreover, both of the cases cited were concerned with federal takings, while in {\it Berman} the Court explicitly says that deference is due in equal measure to the state legislature.\footcite[32]{berman54} It is possible to regard this merely as a {\it dictum}, since the District of Columbia is governed directly by Congress. However, {\it Berman} was to have a great impact on future cases. In effect, it undermined a large body of case law on judicial review of takings without engaging with it at all.

In {\it Hawaii Housing Authority v Midkiff}, the Supreme Court further entrenched the principle expressed in {\it Berman}.\footcite{midkiff84} Here the state of Hawaii had made use of eminent domain  to break up an oligopoly in the housing sector. Given the circumstances of the case, it would have been natural to argue in favour of this taking on the basis that it served a proper public purpose.

However, the Court instead decided to rely on the doctrine of deference, shunning away from scrutinizing the takings purpose. Justice O'Connor, in particular, observed that ``judicial deference is required because, in our system of government, legislatures are better able to assess what public purposes should be advanced by an exercise of eminent domain''.\footcite[244]{midkiff84}

Effectively, what had been a doctrine of deference towards state courts had now transformed into a doctrine of deference towards state legislatures (and, in practice, the executive branch).
\noo{
The formulation here is slightly less absolute than that given in {\it Berman}. In particular, the deferential stance is not presented as a system imperative, but rather made contingent on the fact that legislatures are ``better able'' to assess what counts as a public purpose. Moreover, Justice O'Connor also actively refers to the merits of the taking, especially when she points out that  ``regulating oligopoly and the evils associated with it is a classic exercise of a State's police powers''.\footcite[242]{hawaii84}

Despite these nuances, {\it Midkiff} reaffirmed the main principle expressed in {\it Berman}, namely that the meaning of public use is a matter for legislatures and that the room for judicial review is narrow.} In light of this, it had to be expected that {\it Kelo} would be decided in favour of the taker.\footnote{In fact, as pointed out by Somin, the {\it Kelo} case represents a slight tightening of the earlier line on public use. See \cite{somin07}.} However, the history of the public use requirement tells us that this outcome was by no means inevitable. Hence, the question arises whether legitimacy can be increased by reviving the public use test. The next section sheds some light on this, on the basis of legislative developments in the US after {\it Kel}.

\noo{ 
Formally, the case law on the federal takings clause is not binding on state courts when they assess cases against their own constitutions.\footnote{See \cite[95]{merrill86}.} Moreover, as Merrill notes, state courts have not uniformly responded by embracing deference towards their own legislatures.\footcite[65]{merrill86} Rather, many state courts continued to offer scrutiny of taking purposes, despite the signals coming from the federal level.\footcite[65]{merrill86}

It should be noted, however, that the time after {\it Berman} was also a time when many government bodies throughout the US would actively seek to condemn homes for redevelopment projects, to combat ``blight'', but often also to the benefit of commercial enterprises.\footnote{See generally \cite{pritchett03}.} Hence, continued public use scrutiny at state courts might also reflect an increased threat of eminent domain abuse. Sometimes, moreover, state courts seems to have failed in their duty to offer appropriate protection.

The case of {\it Poletown Neighborhood Council v City of Detroit} is a classic example.\footcite{poletown81} In this case, the Michigan Supreme Court held that it was not in violation of the public use requirement in the Michigan Constitution to allow General Motors to displace some 3500 people for the construction of a car assembly factory. The majority 5-2 cites {\it Berman}, commenting that the state court's room for review of the public use requirement is similarly limited.\footcite[632-633]{poletown81}

The {\it Poletown} decision was controversial, and the minority, especially Justice Ryan, was highly critical of it. He objects both to the deferential stance in general and to the majority reading of {\it Berman} in particular, pointing out that the Supreme Court's doctrine of deference outside the context of federal takings was  directed at the state courts, not state legislatures.\footcite[668]{poletown81} Hence, as he concludes, the majority's reliance on {\it Berman} was ``particularly disingenuous''.\footcite[668]{poletown81} 

Justice Ryan was not alone in his disapproval of {\it Poletown}.\footnote{Indeed, the decision would later be overturned by the Supreme Court of Michigan itself, see generally \cite{sandefur05}.} Moreover, the case is widely regarded as the prelude to an era of increased tensions over economic development takings in the US.\footnote{See \cite[664-668]{sandefur05}.} This would culminate with {\it Kelo} which, despite upholding and strengthening the deferential doctrine, also inadvertently caused a shift towards stricter public use scrutiny at the state level, as discussed in the following subsection.
}

\subsection{Economic Development Takings after {\it Kelo}}\label{sec:postkelo}

%The fact that {\it Kelo} was decided against the homeowner met with wide disapproval among the public.\footnote{See \cite[2109]{somin09}.} In addition, many scholars expressed concern that the deferential approach had been taken too far, and that economic development takings such as {\it Kelo} were in need of more substantive public use scrutiny by courts.\footnote{For a small sample, see \cite{cohen06,underkuffler06,sandefur06,somin07,gisler10}.}
%Moreover, 
Following {\it Kelo}, much attention was directed at the perceived dangers of eminent domain abuse in the US.\footnote{See generally \cite{somin09}.} %The minority opinions given in {\it Kelo}, particularly the opinion of Justice O'Connor, proved influential, causing further attention to be directed at the perceived dangers of eminent domain abuse. A massive amount of literature has since appeared devoted to studying  economic development takings. 
Many states responded by introducing reforms aimed at limiting the use of eminent domain for economic development.\footnote{For an overview and critical examination of the myriad of state reforms that have followed {\it Kelo}, I point to \cite{eagle08}. See also \cite{somin09}.} Within two years, 44 states had passed post-{\it Kelo} legislation in an attempt to achieve this.\footnote{See \cite{castle}.} Various legislative techniques were adopted. Some states, including Alabama, Colorado and Michigan, enacted explicit bans on economic development takings and takings that would benefit private parties.\footcite[See][107-108]{eagle08} In South Dakota, the legislature went even further, banning the use of eminent domain: ``(1) For transfer to any private person, nongovernmental entity, or other public-private business entity; or (2) Primarily for enhancement of tax revenue''.\footnote{South Dakota Codified Laws § 11-7-22-1, amended by House Bill 1080, 2006 Leg, Reg Ses (2006).}

In other states, more indirect measures were taken, such as in Florida, where the legislature enacted a rule whereby property taken by the government could not be transferred to a private party until 10 years after the date it was condemned.\footcite[809]{eagle08} Many states also offered lengthy lists of uses that were to count as public, designed to restrict the room for administrative discretion while allowing condemnations for purposes that were regarded as particularly important.\footcite[804]{eagle08}

As Somin has pointed out, state reforms enacted by the public through referendums tend to be more restrictive than reforms passed through the state legislature.\footcite[2143]{somin09} Many of the more radical reform proposals, moreover, were not endorsed by any of the branches of government, but were initiated by activist groups as ballot measures.\footnote{In some US states, initiative processes make it possible for activist groups to put measures on the ballot without prior approval by the state legislature. See \cite[2148]{somin09}.} As Somin observes, the reforms taking place via this route would be comparatively strict, testifying to the power of direct democracy.\footnote{See \cite[2143-2149]{somin09}.}

Indeed, the successes of popular anti-takings movements  underscores how strongly the US public opposed the decision in {\it Kelo}. Surveys show that as many as 80-90 \% believe that it was wrongly decided, an opinion widely shared also among the political elite.\footcite[2109]{somin09} 

{\it Kelo} has clearly had a great effect on the discourse of eminent domain in the US. However, the effects of the many state reforms that have been enacted are less clear. According to Somin, most of these reforms have in fact been ineffective, despite the overwhelming popular and political opposition against economic development takings.\footcite[2170-2171]{somin09} At the same time, property lawyers report a greater feeling of unease regarding the correct way to approach the public use requirement, expressing hope that the Supreme Court will soon revisit the issue.\footnote{See \cite{murakami13} (``Until the Supreme Court revisits the issue, we predict that this question will continue to plague the lower courts, property owners, and condemning authorities'').} 

Why have legislative reforms proved inadequate and ineffective? Part of the reason, according to Somin, is that people are ``rationally ignorant'' about the economic takings issue.\footnote{See \cite[2170]{somin09}.} For most people, it is unlikely that eminent domain will come to concern them personally or that they will be able to influence policy in this area. Hence, it makes little sense for them to devote much time to learn more about it. This, in turn, helps create a situation where experts can develop and sustain a system based on practices that a majority of citizens actually oppose.\footcite[2163-2171]{somin09} Indeed, Somin argues that surveys show how people tend to overestimate the effectiveness of eminent domain reform, possibly due to the fact that symbolic legislative measures are mistaken for materially significant changes in the law.\footcite[2155-2157]{somin09}

\noo{ I think Somin's analysis is on an interesting track. However, it should be noted that the notion of rational ignorance is a double-edged sword with regards to his main argument. In particular, it seems possible, in theory, that the prevailing critical attitude towards economic development takings is itself an instance of such ignorance. Perhaps people would change their opinion on economic development takings if they were better educated on the issue?

However, this possibility does nothing to detract from the main message, which is} The upshot is that the {\it Kelo} backlash have in fact caused greater insecurity about what the law is and what it delivers, often without significantly curbing those uses of eminent domain that are regarded as most problematic. Arguably, this shows that the legislative approach so far, which has focused on introducing more elaborate and detailed versions of the public use restriction, need to be supplemented by different kinds of proposals. 

In this regard, it seems important to also target governmental decision-making processes regarding the use of private land for economic development. These processes, it seems, need to be imbued with greater legitimacy. In particular, it seems crucial that owners themselves should be granted a better chance to participate in the management of their own land, even when this involves deliberating on, and possibly taking part in, large-scale development projects. After all, it is the owners' and their communities' feeling that they are being treated unfairly that tend to lie at the root of controversies surrounding takings for economic development.\footnote{For a similar perspective, see \cite{underkuffler06}.}

This points towards another perspective on legitimacy, whereby focus shifts towards the institutional setting where the relevant decisions are made. Importantly, cases such as {\it Kelo} suggest that scrutiny against institutional standards need to encompass more than assessments against administrative law and procedural due process principles. Specifically, institutional legitimacy appears to have an important quasi-substantive component whereby a decision is legitimate only in so far as it results from democratically legitimate decision-making within an administrative framework that is generally conducive to fair and proportional outcomes. Arguably, as discussed in the next section, recent developments at the ECtHR point towards a perspective on legitimacy that emphasises this interconnectedness between substantive and procedural aspects of fairness at the institutional level.

%If improved principles of governance are put in place, this alone might be enough to restore some confidence in eminent domain as a procedure by which to implement democratically accountable decisions about land use. However, it seems that eminent domain as such is often an unduly blunt instrument when society desires commercial development on private land. Instead, it might be possible to devise mechanisms for collective action that replace the use of eminent domain altogether. This way of thinking calls for an institutional perspective of legitimacy. Potentially, owners and local communities should be able to challenge takings on the basis that an alternative institutional framework granted the opportunity of \footnote{I mention that aspects of this idea can be found on continental Europe, where some jurisdictions admit something called the ``self-realisation argument'', by which owners can fend off a threat of eminent domain by committing to fulfilling the public purpose themselves. See generally \cite[...]{sluysman15}.}}
%Arguably, an example of such a shift in perspective can be found in recent developments at the ECtHR, as discussed in the following section.

%At least, it should be possible to devise mechanisms for benefit sharing in these cases, to make the imposition of a development project appear less unfair to local owners.
%\noo{
%In the next section, I will consider a proposal for reform of this kind. %The first specifically targets the question of benefit sharing by proposing a special negotiation mechanism for determining the level of compensation after an economic development taking. 
%This proposal targets the decision-making process leading to economic development by proposing a framework for land assembly that can sometimes replace the use of eminent domain.
%}

\section{Recent Developments at the ECtHR: Legitimacy as Institutional Fairness}\label{sec:echr}

It is often said that the P1(1) of the ECHR consists of three rules. The first rule guarantees a right to ``peaceful enjoyment of possessions'', while the second and the third rule expressly regulates the legitimacy of ``deprivation'' and ``control on the use'' of property respectively.\footnote{See \cite[61]{sporrong82}.}

When dealing with complaints pertaining to P1(1), the Court in Strasbourg will typically first consider which of these three rules it should apply.\footnote{See \cite[102-104]{allen05}.} However, as noted by Allen, it is not clear that this choice has any great significance for the outcome.\footnote{See \cite[104-105]{allen05}.} In practice, the evaluation of legitimacy proceeds in much the same way regardless of which rule is used, with an emphasis on the {\it fair balance} that needs to be struck when states interfere with private property rights.\footnote{See \cite[[103]{allen05}. It is also typically assumed that an interference is only legitimate when it takes place for an appropriate purpose, but here the ECtHR has consistently maintained a deferential stance, pointing to the ``wide margin of appreciation'' that the member states enjoy in this regard. See, e.g., \cite[54]{james86}.} 

%ether an interference is classified as a potential interference with the ``peaceful enjoyment of possession'', or as a potential infringement of one of the two exceptions.

%In light of this, a classification based on legitimacy principles is arguably a more instructive starting point for further analysis. Following Allen, there are three important principles to consider. First, there is the principle of {\it legality}, asserting that an interference needs to be authorized by law. Second, there is the principle of {\it appropriate purpose}, making clear that interference should only take place in pursuance of a legitimate public purpose. Third, there is the {\it fair balance} principle, requiring proportionality between the means and the aims underlying the interference.\footnote{See \cite[69]{sporrong82} and \cite[120]{james86}.} %which is applied by the ECtHR in almost all cases when it finds that there has been a violation of P1(1).

%For the issues raised in this thesis, it would appear that the {\it appropriate purpose} principle is the most relevant. However, the starting point for property adjudication at the ECtHR is that States have a ``wide margin of appreciation'' with regard to the question of whether or not an interference serves a legitimate purpose.\footcite[See][54]{james86} This question is thought to depend on democratically determined policies to such an extent that it is rarely appropriate for the Court to censor the assessments made by member states. At the same time, 

In this regard, it is important to note that the Court has gradually adopted a more active role in assessing whether or not particular instances of interference are proportional and able to strike a fair balance between the interests of the public and the property owners. As argued by Allen, this has caused P1(1) to attain a wider scope than what was originally intended by the signatories.\footcite[1055]{allen10}

In the early case law behind this development, the focus was predominantly on the issue of compensation, with the Court gradually developing the principle that while P1(1) does not entitle owners to full compensation in all cases of interference, the fair balance will likely be upset unless at least some compensation is paid, based on the market value of the property in question.\footnote{See \cite[103]{scordino06}. The case also illustrates that the Court has adopted a fairly strict approach to the question of when it is legitimate to award less than full market value.} %The focus on compensation has also been reflected in academic work on P1(1), which tends to address proportionality from a financial perspective, by investigating to what extent owners are entitled to compensation based on the market value of their property. Indeed, when considering case law and literature on the subject, one is left with the impression that ``fair balance'' with regards to P1(1) is crucially linked to financial entitlements. It seems that d as a standard that can justify a right to compensation that goes beyond what the wording of P1(1) might initially suggest.

However, the fair balance test encompasses more than the issue of compensation. In particular, the hunting cases discussed in Chapter 1 show that the Court in Strasbourg is willing to reflect broadly on the context and purpose of interference, to critically assess the social function of the taking.\footnote{See Section \ref{sec:hunt} of Chapter 1.} Moreover, institutional aspects of fairness have come to play an important role in the Court's reasoning in recent cases involving rent control schemes and housing regulation.\footnote{See \cite{hutten06,lindheim12}.} This is particularly clearly demonstrated by the case of {\it Hutten-Czapska v Poland}.\footnote{See \cite{hutten06}.}


%There are obvious financial interests at stake in such cases, for both landlords and tenants. However, the Court has addressed these cases by looking to the fairness of the underlying regulation more generally, by critically evaluating the social, economic and political context. Moreover, the Court has not shied away from using concrete cases as a starting point for providing an assessment of the sustainability of national provisions as such.

\noo{ In {\it Chassagnou and others v France} the situation was that landowners were compelled to permit hunting on their land, following compulsory membership in a hunting association which was set up to manage hunting in the local area.\footcite{chassagnou99} The owners protested this on the grounds that they were ethically opposed to hunting. The Court agreed that there had been a breach of P1(1). 

In the later case of {\it Hermann v Germany}, the circumstances were similar and the Court followed the precedent set in {\it Chassagnou}. In addition, the Court commented that they had ``misgivings of principle'' about the argument that financial compensation could provide adequate protection in such a case.\footcite[See][91]{hermann12}  In this way, the hunting cases illustrate that to the ECtHR, the right to property is more than a  financial entitlement. The fair balance that must be struck could pertain to other aspects, such as the owner's right to make use of his property in accordance with his convictions and to take part in decision-making processes regarding how it should be managed.\footnote{The assessment of proportionality should be concrete and contextual, and it is not based on a narrow or formalistic concept of property as dominion. This is demonstrated, for instance, by \cite{chabauty12}. Here the Court found no violation of P1(1) although the facts seemed close to those of {\it Chassagnou}. The case differed, however, in that the owner himself was not opposed to hunting, but wanted to withdraw his land from the hunters' association to enjoy exclusive hunting rights.}
}

\subsection{Legitimacy {\it Erga Omnes}}

The striking conclusion in {\it Hutten-Czapska v Poland}, underscoring the institutional turn at the ECtHR, was that the case demonstrated ``systemic violation of the right of property''.\footcite[239]{hutten06} The case concerned a house that had been confiscated during the Second World War. After the war, the property was transferred back to the owners, but in the meantime, the ground floor had been assigned to an employee of the local city council.\footcite[20-31]{hutten06} The state implemented strict housing regulations during this time, which eventually led to the applicant's house being placed under direct state management.\footcite[20-31]{hutten06} Following the end of communist rule in 1990, the owners were given back the right to manage their property, but it was still subject to strict regulation that protected the rights of the tenants.\footcite[31-53]{hutten06} In addition to rent control, rules were in place that made it hard to terminate the rental contracts.\footcite[20-53]{hutten06} 

After an in-depth assessment of the relevant parts of Polish law and administrative practice, the Grand Chamber of the ECtHR concluded that there had been a violation of P1(1). Importantly, they did not reach this conclusion by focusing on the owners and the interference that had taken place with respect to their individual entitlements. Rather, they focused on the overall character of the Polish system for rent control and housing regulation, as exemplified by the applicant's situation.

Specifically, the consequences for the owners were considered not in isolation, but in order to shed light on a broader question of sustainability.\footcite[60-61]{hutten06} The Court was particularly concerned with the fact that the total rent that could be charged for the house in question was not sufficient to cover the running maintenance costs.\footcite[224]{hutten06} In particular, it was noted that the consequence of this would be ``inevitable deterioration of the property for lack of adequate investment and modernisation''.\footnote{\cite[224]{hutten06}.}

In the end, the Court highlighted how three factors combined to bring both owners and their properties  to a precarious position. First, the rigid rent control system made it hard to sustainably manage rental property. Second, tenancy regulation made it hard for owners to terminate tenancy agreements. Third, the Court noted that the state itself had set up many tenancy agreements during the days of direct state management, shedding doubt on the fairness of the obligations that these contracts imposed on owners.\footcite[224-225]{hutten06} 

The Court's reasoning in {\it Hutten-Czapska} is also interesting because of how the ``social rights'' of the tenants is placed on an equal footing to the property rights of the owners.\footcite[225]{hutten06} Arguably, property rights and social rights are not considered merely as separate sets of entitlements, locked in opposition to one another. In the reasoning of the Court they also appear as mutually dependent social functions, both hampered by an unsustainable approach to property and housing during the communist era and beyond.\footnote{Specifically, the Court attached great significance to the finding that rents were too low to cover maintenance costs, see \cite[224]{hutten06}. A lack of incentives for maintenance is clearly a threat to tenants as much as to owners, illustrating the interdependence between the two groups. Despite this, when summing up their reasoning in broad strokes, the Court itself reverts back to a traditional narrative when it speaks about the ``conflicting interests of landlords and tenants''. See \cite[225]{hutten06}.}

In this regard, the Court places considerable weight on the precarious situation of the owners and ``the absence of any legal ways and means making it possible for them either to offset or mitigate the losses incurred in connection with the maintenance of property or to have the necessary repairs subsidised by the State in justified cases''.\footnote{See \cite[224]{hutten06}.} Moreover, the Court commented that the ``burden cannot, as in the present case, be placed on one particular social group, however important the interests of the other group or the community as a whole''.\footnote{See \cite[225]{hutten06}.} Importantly, however, the Court did not set out to censor the political reasoning that motivated the rent control scheme, but rather focused on the fact that it had not been implemented properly.\footnote{See \cite[224]{hutten06}.}

On this basis, the Court concluded that there had been a systemic violation of P1(1), and ordered Poland to take measures to rectify the ``malfunctioning of Polish housing regulation''.\footnote{See \cite[237]{hutten06}. The basis relied on for formulating such an order was Article 41 in the ECHR, first used in this way in the case of \cite{broniowski05}.} Hence, the lack of legitimacy was pronounced with a kind of {\it erga omnes} effect, establishing an obligation for Poland directed at all its citizens in equal measure, not merely the applicant.\footnote{There was some dissent as to whether or not this was an appropriate response, with Judge Zagrebelsky in particular arguing against it on the grounds that it would see the Court ``entering territory belonging specifically to the realm of politics''. See \cite{hutten06}.} Judgements of this kind, known as ``pilot judgements'', have now gained formal recognition as a distinct procedural form that the ECtHR can use to address systemic problems.\footnote{See generally \cite{leach10}.}

The institutional approach conditioned by the introduction of pilot judgements might point to the core function that the ECtHR is likely to serve in the future.\footnote{See, e.g., \cite{greer12} (arguing that a ``constitutional pluralism'' approach to adjudication -- better filtered, more principled, yet still context sensitive --  is the way ahead for the ECtHR).} Indeed, the ECtHR will hardly be able to protect human rights in Europe on a case-by-case basis. Nor would it seem appropriate for it to do so, given its remoteness to local conditions and its relative lack of democratic accountability. However, when the Court is able to identify systemic failures that look set to systematically give rise to imbalances and unfairness, it seems appropriate that it should take action.

This is particularly clear when, as in the case of {\it Hutten-Czapska}, the Court notes that the applicants have insufficient options available for achieving a fair balance by appealing to institutions within the domestic legal order. In such cases, it seems appropriate for the Court to demand a change at the level of the state's own institutions, giving rise to a broad duty for the state to improve those institutions. Moreover, by scrutinizing the procedures and principles that the states apply when fulfilling this duty, it is likely that the Court will still be able to steer and unify the development of the case law on human rights, at least to the extent that this is required to meet minimum standards.

Against the deferential implications of this shift of attention, it could be argued that the judicial or administrative bodies of the signatory states can easily circumvent their obligations by providing superficial reforms or biased assessments of the facts in human rights cases, to avoid embarrassment for the state's political or bureaucratic elites. However, this might then be raised as a more procedurally oriented complaint before the ECtHR, perhaps also against Articles 6 (fair trial) and 13 (effective remedy).\footnote{I note that this also fits with recent developments at the ECtHR, toward somewhat broader scrutiny under Article 6, see \cite{khamidov07}.}  

In this way, the Court can streamline its functions, by always aiming to direct attention at issues that arise at a higher level of abstraction.\footnote{A similar argument was given by Judge Zupan\u{c}i\u{c} in \cite{hutten06} (``Is it better for Poland to be condemned in this Court 80,000 times and to pay all the costs and expenses incurred in 80,000 cases, or is it better to say to the country concerned: “Look, you have a serious problem on your hands and we would prefer you to resolve it at home...! If it helps, these are what we think you should take into account as the minimum standards in resolving this problem...”? Which one of the two solutions is more respectful of national sovereignty?'').} This, in my view, seems highly desirable. The ECtHR should not aim to micromanage the signatory states, particularly not in relation to a norm such a P1(1), which the Court itself regards as highly dependent on context. By shifting attention towards institutional fairness, the Court can avoid getting stuck in deference to the states without overstepping its bounds with regards to the democratic process.

Indeed, the case of {\it Hutten-Czapska} is highly suggestive of the merits of such a perspective, not only because of the special measures ordered, but also because the Court reasoned on the basis of institutional information to identify systemic weaknesses of Polish housing regulation.\footnote{Specifically, it seems that the shift signalled by recent cases on property at the ECtHR does not end with a new take on remedies, but also signals some changes in the way the Court approaches the fair balance determination under P1(1). This seems natural; if the Court looks for systemic violations, not (only) individual transgressions, its substantive assessments of fairness will naturally be influenced.} Another example is the recent case of {\it Lindheim and others v Norway}.\footnote{See \cite{lindheim12}.} Here the applicants complained that their rights had been violated by a Norwegian act that gave lessees the right to demand indefinite extensions of ground leases on pre-existing conditions.\footcite[119]{lindheim12}

The Court agreed that this was a breach of P1(1). Moreover, it engaged in the same form of assessment as it had adopted in {\it Hutten-Czapska}. Specifically, it concluded that the Ground Lease Act 1996 as such was the underlying source of the violation. The problem was not merely that this act had been applied in a way that offended the rights of the applicants. In light of this, the Court did not only award compensation, it also ordered that general measures had to be taken by the Norwegian state to address the structural shortcomings that had been identified.\footnote{See \cite{lindheim12}.}

The Court also commented that its decision should be regarded in light of ``jurisprudential developments in the direction of a stronger protection under Article 1 of Protocol No. 1''.\footcite[135]{lindheim12} However, in light of the change in perspective that accompanies this development, it is interesting to ask in what sense exactly the protection is stronger. In particular, it is not {\it prima facie} clear that the Court's remark should be read as a statement expressing a change in its understanding of the content of individual rights under P1(1). 

Rather, it may be read as a statement to the effect that the Court has assumed greater authority to address structural problems under that provision. This might even allow the Court to conclude that a violation has occurred due to structural unfairness, even when it is not possible to trace this back to any abnormal decision that specifically targets the individual entitlements of applicants.

If this is true, it could arguably make a big difference in cases involving takings for economic development. As illustrated by Justice O'Connor's dissent in {\it Kelo}, a main concern here is that such takings are likely to have ``perverse'' consequences at the structural level, because they lack democratic merit and give rise to a gradual concentration of wealth and power.\footnote{To quote Justice O'Connor's dissent in {\it Kelo}, see \cite{kelo05}.} In light of cases such as {\it Hutten-Czapska} and {\it Lindheim}, I think the ECtHR would have been likely to approach {\it Kelo} in a manner consistent with Justice O'Connor's approach.

Whether they would reach the same conclusion seems more uncertain, particularly since confidence in the states' ability and willingness to regulate private-public partnerships might be higher in Europe than in the US.\footnote{For a discussion from the point of view of English law, arguing that the prevailing regulatory regime limits the risk of eminent domain abuse largely through regulation of the takings power rather than strict property protection, see \cite{allen08}.} However, it seems unlikely that the ECtHR would follow the majority in {\it Kelo}, by simply deferring to the determinations made by the granting authority. Rather, Justice O'Connor's predictions about the ``fallout'' of the {\it Kelo} decision would likely have been of significant interest also to the justices at the Court in Strasbourg.

To conclude, I think notions of institutional fairness can help us locate a welcome middle ground between procedural notions of legitimacy anchored in democracy and the rule of law, as found in England and Wales, and substantive notions anchored in notions such as public use and public interest, such as those found in the US. The question remains how Courts adopting such a middle ground should proceed when presented with a concrete case of alleged eminent domain abuse. In the next section, I present a possible heuristic.

\noo{
\subsection{How Would the ECtHR Approach an Economic Development Taking?}

Is the jurisprudential developments illustrated by the rent control cases relevant to the issue of economic development takings? I believe so. Indeed, I am struck by how the reasoning of the ECtHR in recent cases on hunting and rent control mirrors the kind of reasoning that Justice O'Connor engaged in when considering {\it Kelo}.\footnote{See \cite{kelo05}.} The emphasis is on structural aspects and fairness, grounded on the facts of the concrete case, but mainly interested in what these facts reveal about the rules and procedures involved. 

This is a contextual approach that can maintain a broad focus without loosing its bite. The crux of arguments used to conclude violation is the observation that the system currently in place can offend against the role that owners {\it should} occupy in order to be able to meet those obligations and exercise those freedoms that are attached to the properties they posses.

On this narrative, interference becomes illegitimate when it demonstrates a failure of governance. In the case of \textcite{hutten06}, this boiled down to the observation that it was illegitimate to address problems in the Polish housing sector by placing the burden ``on one particular social group'', namely the owners.\footcite[225]{hutten06} This conclusion was backed up by the concrete observation that the rules and procedures in place meant that owners who were obliged to maintain their properties in good condition for their tenants were in fact prevented from doing so because they were not permitted to charge rents that would cover the costs.

In the case of {\it Kelo}, Justice O'Connor argued in a similar fashion when she concluded that the system which had led to the decision to condemn Suzanne Kelo's house was likely to function so as to systematically ``transfer property from those with fewer resources to those with more''. To Justice O'Connor, there was little doubt that this could become a general pattern, if safeguards were not put in place. %Indeed, it would have to be assumed that a multi-million dollar company would always be in a better position than a homeowner when arguing that  ``economic development'' would result from their ownership. \noo{More subtly, her opinion also hinted at the inconsistency involved in asserting abstractly that economic development would benefit the community indirectly, all the while the development would \isr{in} fact require razing it.}

\section{Institutional Proposals for Increased Legitimacy}\label{sec:ir}

In this subsection, I first present the Special Purpose Development Companies (SPDCs) proposed by Lehavi and Licht.\footcite{lehavi07} I relate this proposals to theoretical approaches to the issue of compensation, before I go on to note some shortcomings and open questions that I will later address in my case study. I then go on to consider the Land Assembly Districts (LADs) proposed by Heller and Hills.\footcite{heller08} I consider this proposal in light of the stated motivation, which is to design an effective mechanism of self-governance that can replace eminent domain in economic development cases. I present some unresolved questions and argue that there is a tension in the proposal between its narrow scope, imposed to prevent majority tyranny and other forms of abuse, and its broad goal of empowering local communities. 

\subsection{Special Purpose Development Companies}

An important distinguishing feature of economic development takings is that they give the taker an opportunity to profit commercially from the development. This may even be the primary aim of the project, with the public benefiting only indirectly through potential economic and social ripple effects. Property owners facing condemnation in such circumstances might expect to take a share in the profit resulting from the use of their land. However, in many jurisdictions, including the US, the rules used to calculate compensation prevents owners from getting any share in the commercial surplus resulting from development.\footnote{See, e.g., \cite[965-966]{fennell04}.} In particular, various {\it elimination rules} are typically in place to ensure that compensation is based entirely on the pre-project value of the land that is being taken.\footcite[See][81]{freilich06} That is, the value of the development potential itself is not to influence the compensation payment (at least not to a greater extent than it was already reflected in the value of the property prior to the development plans). The policy reasons for such rules is that they ensure that the public does not have to pay extra due to their own special want of the property. After all, this is one of the main purposes of using eminent domain in the first place: to ensure that the public does not have to pay extortionate prices for land needed for important projects. However, when the purpose of the project is itself commercial in nature, there appears to be a shortage of good policy reasons for excluding this value from consideration when compensation is calculated. This is especially true when, as in the US, compensation tends to be based on the market value of the land taken. Why should a commercial condemner's prospect of carrying out economic development with a profit be disregarded when assessing the market value? In any fair and friendly transaction among rational agents, one would expect benefit sharing in a case like this. Yet for economic development backed up by eminent domain, the application of elimination rules ensures that all the profit goes to the developer. 

Some authors have argued that failures of compensation is at the heart of the economic takings issue and that worry over the public use restriction is in large part only a response to concerns about the ``uncompensated increment'' of such takings.\footcite[See][962]{fennell04} In addition to the lack of benefit sharing, previous work has identified two further problems of compensation that also tend to become exasperated in economic development cases. First, the problem of ``subjective premium'' has been raised, pointing to the fact that property owners often value their own land higher than the market value, for personal reasons.\footcite[963]{fennell04} For instance, if a home is condemned, the homeowner will typically suffer costs not covered by market value, such as the cost of moving, including both the immediate ``objective'' logistic costs as well as more subtle costs, such as having to familiarize oneself with a new local community. Second, the problem of ``autonomy'' has been discussed, arising from the fact that an exercise of eminent domain deprives the landowner of \isr{their right to decide how to manage their} property.\footnote{Discussed in \cite[966-967]{fennell04}. For a general personhood building theory of property law, see \cite{radin93}. For a general economic theory of the subjective value of independence, see \cite{benz08}.}

In \footcite{lehavi07}, the authors propose a novel approach for addressing the ``uncompensated increment'' in economic takings cases. Their proposal is based on a new kind of structure that they dub a {\it Special Purpose Development Corporation} (SPDC). The idea is that owners affected by eminent domain will be given a choice between standard pre-project market value and shares in a special company. This company will exist only to implement a specific step in the implementation of the development project: the transaction of the land-rights. The SPDC may choose either to offer their rights on an auction or else negotiate a deal with a designated developer.\footcite[1735]{lehavi07} Hence, the idea is to ensure that the owners are paid a value that reflects the post-project value of the land, but in such a way that the holdout problem is avoided. In particular, the SPDC will have a single task: to sell the land for the highest possible price within a given time frame.\footcite[1741]{lehavi07} After the sale is completed, the SPDC will divide the proceeds as dividends and be wound up.\footcite[1741]{lehavi07}

Other suggestions have taken a more static approach to compensation reform, such as proposing to give owners a fixed premium in cases of economic development, or developing mechanisms of self-assessment to ensure that compensation is based on the true value the owner attributes to his own land.\footnote{A range of static proposals have been proposed in the literature: Merrill proposes 150 \% of market value for takings that are deemed to be ``suspect'', including takings for which the nature of the public use is unclear, see \cite[90-93]{merrill86}. Krier and Serkin propose a system that provide compensation for a property's special suitability to its owner, or a system where compensation is based on the court's assessment of post-project value, see \cite[865-873]{krier04}. Fennell proposes a system of self-evaluation of property for taking purposes with tax-breaks given to those who value their property close to market value (to avoid overestimation), see \cite[995-996]{fennell04}. Bell and Parchomovsky also propose self-evaluation, but rely on a different mechanism to prevent overestimation; tax liability is based on the self-reported value and no property can be sold by its owner for less than his reported value, see \cite[890-900]{bell07}.} Compared to such proposals, the idea of SPDCs is more sophisticated and should be looked at in more depth. 

The conceptual premise for the proposal is that takings for economic development can be seen as compulsory incorporation, a pooling of resources useful in overcoming market failures.\footcite[1732-1733]{lehavi07} Just as the corporation is formed to consolidate assets in order to facilitate effective management, so is eminent domain used to assemble property rights in order to facilitate efficient organization of development. According to Lehavi and Licht, this also provides a viable approach to problems of ``opportunistic behavior''; hierarchical governance after assembly ensures that order and unity can be regained even if interests in the land are distributed among a large and heterogeneous group of potentially mischievous shareholders.\footcite[1733]{lehavi07} In the words of Lehavi and Licht:

\begin{quote}
The exercise of eminent domain powers thus resembles an incorporation by the government of all landowners with a view to \isr{bringing} all the critical assets under hierarchical governance. Establishing a corporation for this purpose and transferring land parcels to it thus would be merely a procedural manifestation of the substantive economic reality that already takes place in eminent domain cases.
\end{quote}

As soon as we look at the rationale behind economic development takings in this way, any remnant of good policy reasons for ensuring that the developer gets all the profit seems to disappear. Rather, we are led to consider compensation as an issue entirely separate from the exercise of the takings power. After the land has been \isr{reorganised} by eminent domain and an SPDC has been formed, the land rights might as well be sold {\it freely} to a developer. In this way, the land will be sold for a price that is closer to an actual market value, on the market where the land is destined for development.\footcite[1735-1736]{lehavi07} More generally, the SPDC becomes an aid that the government can use to create more \isr{favourable} market conditions for transferring land that has commercial potential in its public use. Due to the compulsory pooling of resources, no owner can exercise monopoly power by holding out, but due to decoupling of compensation from assembly, the owners can now negotiate with potential developers for a share of the resulting profit. Moreover, the fact that the SPDC offers its rights on an actual market can also help ensure that more information \isr{becomes} available regarding the true economic value of the development, something that may in turn help ensure that only the good projects will be successful in acquiring land. Hence, according to Lehavi and Licht, an additional positive effect of SPDCs is that developers and governments will \isr{shy }away from using the eminent domain power to benefit projects that are not truly welfare-enhancing.\footcite[1735-1736]{lehavi07}

In addition to these substantive consequences, the SPDC-proposal also stands out because it has a significant institutional component, pointing to its potential for restoring procedural legitimacy as well as substantive fairness. Lehavi and Licht discuss corporate governance issues at some length, but without committing themselves to definite answers about how the operations of the SPDC should be \isr{organised}.\footcite[1040-1048]{lehavi07} Indeed, while their proposal is perhaps most interesting because of its procedural aspects, it also appears to be rather preliminary in this regard. The main idea is to let the SPDC structure piggyback on existing corporative structures, particularly those developed for \isr{securitisation} of assets.\footnote{See generally \cite{schwarcz94}. For an up-to-date overview, targeting special challenges that became apparent during the 2008 financial crisis, see \cite{schwarcz13}.} The basic idea is that the corporate structure should be insulated from the original landowners to the greatest possible extent; it should have a narrow scope, it should be managed by neutral administrators, and it should entrust a third party with its voting rights.\footcite[1742]{lehavi07} This is meant to prevent failures of governance within the SPDC itself, making it harder for majority shareholders and self-interested managers to co-opt the process. For instance, if a possible developer already holds a majority of the shares in an SPDC, this structure would prevent him from using this position to acquire the remaining land on \isr{favourable} terms. 

Lehavi and Licht observe that under US law, the government would often be required to make shares in an SPDC available to the landowners as a public offering.\footcite[1745]{lehavi07} Lehavi and Licht deem this to be desirable, arguing that full disclosure will provide owners with a better basis on which to decide whether or not to accept SPDC shares in place of pre-project market value. It will also facilitate trading in such shares, so that they will become more liquid and therefore, presumably, more valuable.\footcite[1746]{lehavi07} 

Lehavi and Licht's proposal is interesting, but I think a fundamental objection can be raised against it. In particular, it seems that their governance model more or less completely alienate property owners from the decision-making process after SPDC formation. Limiting the participation of owners is to a large extent an explicit aim, since governance by experts is held to increase the chances of ensuring good governance. But is expert rule really the answer?

It seems that from the owners' point of view, Lehavi and Licht's proposals for governance reduces the SPDC to a mechanism whereby they can acquire certain financial entitlements. These may exceed those that would follow from standard compensation rules, but they do not directly empower owners vis-{\'a}-vis developers and the government. Instead, a largely independent structure will be introduced. It is this new \isr{organisational} structure, rather than the owners, that will now become an important actor in the eminent domain process. In principle, it is meant to represent owners, but to what extent can it do so effectively? After all, it is specifically intended to operate as neutral player, charged with \isr{maximising} the price, nothing more. Hence, it appears that the SPDC will not be able to give owners an arena to negotiate on the basis of property's social functions. Indeed, the institutional component of the SPDC proposal specifically targets a narrow, entitlements-oriented, perspective on what it means to be an owner and why property should be protected. 

\noo{
the personal and social importance they attribute to their land rights. How the problem of ``autonomy'' is addressed by the proposal is therefore hard to see and the ``subjective premium'' also appears to be in danger, unless it can be objectively quantified and covered by the surplus from a voluntary sale. But if such quantification is possible, then why not simply tell the appraiser to award some premium under standard compensation rules?

More generally, it seems to me that while all three categories of ``uncompensated increments'' are interesting to study from a financial viewpoint, severe doubts can be raised regarding the feasibility of addressing the subjective aspects of this as a question of compensation. It may be that issues related to ``subjective premium'' and ``autonomy'' are seen as public use issues for good reason; they are hard to quantify otherwise. Moreover, attempting to do so might do more harm than good. On the one hand, it might skew the political process, since owners that have been ``bought off'' don't object to ill-advised development projects, as long as they generate financial revenue. But what about projects that are undesirable for other reasons, for instance because they completely change the character of a \isr{neighbourhood}, or because they are harmful to the environment? On the other hand, }

With regards to individual aspects of property's social function, such as the personal attachments it engenders, or the sense of autonomy it provides, the idea that money can compensate for the owners' loss is not entirely implausible. Some standard examples include compensation for relocation costs and compensation for the cost of juridical assistance. And, indeed, in many jurisdictions, the law already provides for such compensation.\footnote{See, e.g., \cite[121-126]{garnett06}.}

In general, however, may would no doubt object that financial compensation is beside the point with regards to subjective values pertaining to the individual's own unique relationship with their property. The aftermath of {\it Kelo} itself can serve as an illustration of this.

After the case, Suzanne Kelo remained defiant at first, but eventually decided to settle in 2006, for an offer of USD 442 155, more than USD 319 000 above the appraised value.\footcite[1709]{lehavi07} Despite this, there is no indication that Suzanne Kelo changed her view of the taking. Indeed, after the long struggle she had taken part in, it is easy to imagine that financial compensation, if it was to be an effective remedy at all, would have to be very high. Even after she had settled, Kelo apparently toured the country speaking out against economic takings.

Hence, the significant overcompensation she received, compared to standard market value, did not restore legitimacy, not even to her personally. Moreover, to the community as a whole, it apparently did more harm than good. Indeed, the other owners affected by the same development plan were not pleased, arguing that recalcitrant owners had been unjustly rewarded for holding out.\footcite[1709]{lehavi07} 

This is indicative of the fact that when we move to consider the role of the community, including property dependants that have no formal claims as owners, the compensatory perspective falls short of providing a meaningful approach. Indeed, as laid down by Lehavi and Licht, it does not seem like SPDC will do much good for the community. It will not better enable its members to fulfil their obligations and responsibilities with regards to each other, their land, and society's desire for economic development. Rather, it will render them just as passive as under traditional eminent domain proceedings, the only difference being that they might benefit financially if the SPDC administrators do their job well.

\noo{ subjective importance of property and autonomy can itself prove offensive. At least it seems likely that it would often come to be seen as inadequate and inefficient.\footnote{For more detailed criticism of the compensation approach to the public use issue, see \cite{garnett06}.} Moreover, an owner that is compelled to give up his home after an inclusive process where the public interest has been debated and clearly communicated is likely to feel like he incurs less costs related both to his subjective premium and his autonomy. Hence, the lack of participation in the decision-making process can in itself increase the uncompensated loss. Clearly, no externally managed ``bargain-oriented'' SPDC will be able to resolve this problem.

I conclude that SPDCs have serious shortcoming with regards to the subjective aspects of undercompensation, aspects that can only be addressed if the focus turns towards participation. However, SPDCs do seem promising when it comes to profit-sharing. This, after all, is what the structure is specifically aiming to achieve. In addition, I agree that SPDCs will likely have a positive effect on the other actors in the eminent domain process. In particular, I agree with Lehavi and Licht that greater openness is likely to result, revealing the true merits of development projects, at least in so far as these are translatable into financial terms. The fact that developers must negotiate with an SPDC who can threaten to make the land available an an open auction will likely deter developers and government from pursuing fiscally inefficient projects. Hence, the risk that governments will \isr{subsidised} such projects by giving them cheap access to land will also be reduced. In addition, the presence of a third voice, speaking on behalf of owners, is likely to help achieve a better balance of power in development takings. 

This is a particular concern in cases when competition fails to arise after SPDC formation. To ensure that there are other interested parties, in particular, sems like an important precondition for the proposal to work in practice. In this regard, it is important to \isr{realise} that a lack of interest from other developers may not be due to the superiority of the original developer's plans. It might rather be due to the fact that the scope of the assembly giving rise to the SPDC is so defined as to make alternatives unfeasible. The danger of abuse in this regard seems significant, particularly when developers themselves participate in coming up with the plans that give rise to SPDC formation. 


Moreover, as long as owners remain marginalized in the planning phase, it is easy to imagine situations where the plan itself will be formulated in such a way that only one developer is in a position to successfully implement it. A simple example would be if a prospective developer already owns some of the land that is critical to the plan, and is able to ensure that this land is kept out of the scope of the SPDC. Clearly, if SPDCs are to operate effectively, such instances of manipulation need to be avoided, suggesting that the proposal as it stands needs to be fleshed out in greater detail.
}

The problems addressed here both seem to point to the fact that the SPDCs, while more flexible than other suggestions, are still too static to achieve many of their objectives. In particular, to arrive at genuine market conditions for assessing post-project value, there is still a need for changes in the dynamics of the planning process underlying the taking. Moreover,

This only adds to the number of external authorities on whose judgement and discretion the faith of the community will depend. If the SPDC proposal is thought of as representing as a layer {\it between} the owners, the government and interested developers, the result can also be a further marginalisation of individual owners, for whom it is now even harder to gain access to primary decision-makers.

To better fulfil the goal of ensuring increased legitimacy, there is a need for a mechanism that goes beyond expert bargaining and provides owners with better access to the decision-making process. In the next subsection, I will consider a proposal that aims to address this, by proposing a framework for self-governance. 
}

\section{The Gray Test}\label{sec:gt}

Pointing to early US case law on public use as a ``laboratory of elementary proprietary ideas'', Kevin Gray builds on the evidence found there to provide a set of conditions for recognising what he calls ``predatory takings''. \footnote{See \cite[28-30]{gray11}.} His conditions capture key aspects of eminent domain abuse that I believe should be recognised by a theory of economic development takings inspired by the notion of human flourishing and an institutional perspective on legitimacy. Below, I briefly present the criteria proposed by Gray, as well as three riders that I believe suggest themselves on the basis of the discussions presented earlier in this and the previous chapter. I will refer to the resulting set of conditions as the {\it Grey test}, to be understood as a proposed general heuristic for assessing the legitimacy of takings, especially in situations when there are strong commercial interests present on the taker side. %With such an approach in hand, it becomes possible to embark on a structured evaluation of Justice O'Connor's predictions about the fallout of the deferential stance adopted in {\it Kelo}.

Several combinations of conditions might be sufficient to justify designating a taking as eminent domain abuse. The purpose of the Gray test is not to produce a definite set of such conditions that provide a final answer in any case. Rather, the aim is to provide a heuristic to facilitate concrete assessment against the social, economic and political circumstances surrounding the taking in question. If an economic development taking represents an abuse of power, one would expect it to run afoul with regard to some, and probably several, of the criteria set out in the following points.

\subsubsection*{Balance of Power among the Parties}

In a typical case of eminent domain abuse, the parties that stand to benefit will be more economically and politically powerful than those from whom property is taken.\footnote{See \cite[30-31]{gray11}. Gray himself omits any explicit mention of political power, but it is present in Justice O'Connor's dissent in {\it Kelo}, and in my view clearly belongs here.} This can be reflected in the takers' ability to solicit legal assistance and other services to defend the taking, as well as in the  owners' inability to launch a coordinated defence.\footnote{See \cite[30-31]{gray11}.} If there is an imbalance of power, this is particularly likely to be noticeable early on, during the planning stages, before the decision to condemn has actually been made. 

After the decision has been made, the procedural position of the owners might improve. However, this might not serve to restore any meaningful balance between the parties; when special procedural protections kick in, it will often be too late for the owners to launch an effective defence against the taking. For instance, strict rules concerning cost reimbursement for costs incurred {\it after} the decision to take has already been made, is not a sufficient response to an imbalance of power, especially not in legal systems that do not offer extensive judicial review of takings purposes.

More generally, a possible imbalance of power should be assessed against the decision-making process as a whole, going back to the first initiative made for taking the property in question. A critical  assessment of what role the owners have played in the decision-making process is a good way to uncover more information about imbalances of power, and whether or not such imbalances could have unduly influenced the outcome.

\subsubsection*{The Net Effect on the Parties}

As Gray notes, a hallmark of eminent domain abuse is that the net effect of the taking is a ``significant transfer of valued resource from one set of owners to another''\footnote{See \cite[31]{gray11}.} In itself, this is not a conclusive sign of abuse, but it directs us to ask two important questions. First, we should inquire critically into the main purpose of the taking. Is the transfer of resources between the parties an acknowledged motive or an unacknowledged side-effect of some ostensibly distinct public purpose?

In the latter case, it might be clear that the public purpose is only a pretext for benefiting the taker, in which case it counts as clear evidence of abuse. In less obvious cases, if the transfer of resources arising from the fulfilment of the public purpose was not properly discussed and critically examined by the decision-maker, this too can point towards predation.

The assessment will be different if redistribution of (control over) resources is openly acknowledged as part of the rationale justifying eminent domain. In such cases, it is pertinent to ask further  questions about the economic and social status of the parties, and the structure of the decision-making process, to shed light on whether the redistributive motive itself appears democratically legitimate. If there is eminent domain abuse, one would except the taking to fail to stand up to scrutiny in this regard.

In some cases, it might be debatable whether a taking passes the net effect test. However, the importance of scrutiny is still significant, since it helps bring the crucial questions into the open, thereby ensuring higher quality of the decision-making regarding the taking. Indeed, if the Gray test is applied at an early stage of the proceedings, this in itself might help increase acceptance of the decisions reached. Making room for more extensive legitimacy tests in takings law might well end up bolstering the government's power to take property, as long as the power is used faithfully.

\subsubsection*{Initiative}

In many suspicious economic development takings, the party benefiting commercially from the taking is the party that initially made the suggestion for using eminent domain.\footnote{See \cite[32]{gray11}.} In uncontroversial cases, on the other hand, the initiative tends to come from some government body that seeks to pursue a specific policy goal, e.g., to provide a public service or bestow a benefit on a particular group that is found to be in need of support. The contrast between this and cases when the initiative lies with the commercial beneficiaries themselves point to a disturbance of the decision-making underlying the decision to use eminent domain. As such, it is an important hallmark of abuse.

To investigate further under this point, one should take into account the wider social and political context of the taking, particularly the position of the parties involved. If the beneficiary is both more powerful and privileged than the owners {\it and} takes the initiative for the taking, this is clearly a sign pointing towards predation. On the other hand, if the beneficiaries are marginalised groups who could only expect any consideration if they were to take the initiate themselves, the situation might have to be viewed differently. %In these cases, the fact that the system leaves room for marginalised non-owners to acquire property interests might have to be considered a strength rather than a weakness. Still, as discussed in later points, the appropriateness of using eminent domain for redistributive purposes can be questioned, even if the redistributive goal itself appears democratically legitimate. In such cases, however, the question of legitimacy is unlikely to turn on the initiative test.

\subsubsection*{Location}

The location, in a broad sense of the word, of the property that is taken, can be a strong indicator that eminent domain is inappropriate.\footnote{See \cite[33-34]{gray11}.} For instance, cases involving the taking of dwellings are naturally more suspect than cases involving the taking of barren or unused plots of land. Similarly, the taking of property that is important to the subsistence of the current owner should raise the bar for when a taking may be considered legitimate. Moreover, if the taker's choice of location appears to be one of convenience rather than necessity, this points towards predation. It is particularly telling if alternative locations would be less intrusive, or obviate the need for using eminent domain altogether.

%The location of the property can also attain relevance independently of the current owner. For instance,

On the other hand, the location of the property can sometimes point towards {\it increased} legitimacy of a taking that would otherwise appear suspect. This might be the case, for instance, if the property that is taken has special value to the taker or the community specifically because of its strategic importance with respect to the taker's own property or the rights of non-owners.\footnote{For instance, if riparian owners cannot make rational use of the water flowing over their land without intruding on the land of their neighbours, using eminent domain to resolve this might be considerably less suspect than other kinds of economic development takings. This particular scenario was much discussed in the US during the 19th century, in relation to mill acts which authorised neighbour-to-neighbour takings of limited property rights needed for development. See the discussion in Chapter 2, Section \ref{sec:us}. For an example involving non-owners, consider the increased legitimacy of interference in cases when property rights frustrate efforts to secure rights of non-owners, such as rights to drinking water in cases when riparian owners prevent non-owners access to water for their basic needs.} The proper balance of burdens and benefits might still be upset, but a taking that fits smoothly into a `special value' narrative will be less suspect than one that does not.

\subsubsection*{Social Merit}

As Gray notes, a taking that is hard to justify on the basis of its social merits is more likely to be predatory.\footnote{See \cite[34]{gray11}. Gray writes of lap-dancing clubs and cigarette factories as examples of purposes that are suspect. Importantly, such purposes might well fulfil a public interest requirement via the economic development narrative, yet still fail a social merit test that focuses rather on the social dimensions of the use to which the property will be put.} This asks for closer scrutiny of the kinds of public interests that can be used to justify a taking. If the justification narrative surrounding a taking revolves solely around `trickle-down' effects and the successful business ventures that the taking will facilitate, there is reason to be suspicious. Specifically, if the taking cannot sustain a social merit narrative, whereby attention is shifted away from purely economic considerations, this is a strong independent indication that the taking might count as predation.

The point here is not that the language of social merit should replace the language of public use or public interest as some kind of conclusive test of legitimacy. Rather, the point is that one should always be encouraged to analyse takings specifically in terms of non-economic, social, effects. This is particularly important in difficult cases, because it can help us arrive at a better understanding of where exactly the taking sits on the gray scale between admissible governance and predatory exploitation. 

%If a taking appears to stand up to scrutiny only when embedded in a purely economic narrative, this in itself suggests a lack of legitimacy. Indeed, even if one concedes that incidental economic effects are relevant, it seems clear that however one circumscribes a notion such as public use, this notion certainly encompass {\it more} than merely those incidental economic effects that tend to occupy center stage in legitimacy disputes.\footnote{Indeed, it bears emphasising that those arguing against economic development takings might achieve more by emphasising non-economic aspects, compared to arguing that incidental economic benefits should not at all be considered relevant as a justification for eminent domain.}

\subsubsection*{Environmental Impact}

According to Gray, a typical feature of eminent domain abuse is that it has an adverse environmental impact. Moreover, a typical feature of eminent domain abusers is that they show disregard for such adverse affects.\footnote{See \cite[34]{gray11} (``predatory takers tend to be relatively unperturbed if they lay waste to the earth'').} This is an additional element that pertains specifically to the status of the taker, asking us to consider whether it is appropriate to grant their activities public interest status. It is not primarily a question of how the development stands with regard to environmental regulation. Rather, what is at stake is whether or not the characteristics of the taker and the development plans make it appropriate to use the power of eminent domain. 

It might be appropriate to use environmental law as a starting point, but the relevant environmental standard with regard to the legitimacy question should be drawn up more strictly than the standards generally applied to the type of development in question. Indeed, one should be entitled to expect {\it more} in terms of environmental awareness and concern from a developer and a development plan that benefit from the power of eminent domain. 

Arguably, the mere fact that takers engage in active lobbying for leniency in relation to environmental standards can be enough to shed doubt on the proposition that they act in the public interest. What might otherwise be considered natural and admissible behaviour for a common commercial company can be improper or inadmissible behaviour for one that benefits from eminent domain powers. I note that this particular observation has general import, pertaining to a potentially wider set of obligations that takers may be expected to take on, not only environmental ones. This brings me to the first rider that I propose to add to Gray's original evaluation points.

\subsubsection*{Rider 1: Regulatory Effects}

As discussed in Chapter 2, property has an important regulatory effect, also outside the realm of positive law. This effect typically changes following a taking, sometimes quite dramatically.  For instance, if locally owned property is taken by external commercial actors for high-intensity commercial use, the post-taking regulatory status of the property will most likely be completely different to its status prior to the interference. Moreover, the changed status might have as much to do with informal social functions as it has to do with positive regulation.

It might be, for instance, that the property in question is found in a jurisdiction that emphasises  the freedom of owners to do as they please without state interference. In this case, the fallout of allowing external commercial actors to take locally owned property can be particularly severe, as the new owner is likely to be unconstrained by locally grounded systems for sustainable resource management. In these cases, there is a risk that there will be a `tragedy of the taken', arising from how the taking undermines an important building block of sustainability. 

Indeed, a society based on egalitarianism and strict limits on state interference might find it especially difficult to appropriately restrain the actions of actors who use the eminent domain power to accumulate property for high-intensity use.\footnote{This problem can of course arise independently of the use of eminent domain, e.g., in the context of land grabs arising from voluntary or semi-voluntary transactions. However, the situation appears particularly problematic if the state itself is complicit in bringing about the problem, by undermining property's social function through the use of the takings power.} If this is resolved by increasing the state's power to interfere with private property through regulation, the effect can be a further undermining of local management frameworks, increased subsequent use of eminent domain, and a general spreading and amplification of the democratic deficit already inherent in the original act of taking.

A different regulatory concern is that the legal status of the property can change, for instance because the development in question brings it under the scope of different rules. If so, it should be examined whether the new rules offer weaker protection for the local community, the environment, or the general public interest, in which case it reflects badly on the initial decision to use eminent domain.%\footnote{The case study of Norwegian waterfall expropriation will offer an example of this mechanism, c.f., Chapter 5, Section \ref{sec:x}.}

\subsubsection*{Rider 2: Impact on Non-Owners}

Following up on the theoretical arguments made in Chapter 1, it is appropriate to direct special attention at the status of non-owners directly affected by economic development takings. It is of particular interest to ascertain whether or not the interests of such non-owners were given due consideration prior to the decision to use eminent domain. If their interests appear to have been neglected, or have not been considered at all, there is additional reason to be sceptical of the purported public interest of the taking. Indeed, just as disregard for the environment is a typical sign of predation, a general disregard for local non-owners is also an indicator of abuse.

To shed further light on this, one might first ask what role non-owners played in the decision-making process. If the non-owners directly affected by the taking were allowed to express their opinion, and enjoyed some measure of influence, this can enhance legitimacy. If, on the other hand, the most immediately affected members of the public were not consulted, or not given a proper voice in the proceedings, it indicates abuse.

There is also an important substantive aspect to consider: how is the taking going to affect property dependants without recognised ownership rights? If it is clear that they will suffer severe adverse effects, for instance by being displaced from their homes or by loosing their livelihoods, this must be counted as an indication of predation irrespective of any mitigating procedural arrangements used to create the impression of ``consultation'' or the like.

Importantly, it also follows from the social function perspective that awarding compensation can not by itself excuse shortcomings in this regard. If people are displaced, for instance, the fact that new dwellings are provided somewhere else does not detract from the fact that a community has been destroyed. It is possible that the needs of the public necessitate such a drastic interference with property's proper function, but this should then at once give rise to a more in-depth scrutiny of legitimacy. Moreover, the bar to pass the legitimacy test should be raised considerably in such cases.

\subsubsection*{Rider 3: Democratic Merit}

Perhaps the most important characteristic to consider when assessing the legitimacy of a taking is its democratic merit. In an important sense, putting a taking to the test against this measure serves to encapsulate all the other points raised above. Specifically, it asks us to consider the totality of these factors in order to judge whether good governance standards have been observed within a system based on democratic decision-making. The inquiry made in this regard should not be focused on second-guessing government policies, but should compel us to take seriously the idea that a commitment to democracy places real constraints on the exercise of government power. In this way, an overarching focus on democratic merit can hopefully render the principles of scrutiny expressed by the Gray test as a possible template for courts across different jurisdictions, with respect to both constitutional and human rights provisions.

%By emphasising how deference depends on substantive standards of democratic decision-making, on might succeed in justifying more extensive judicial review of takings on the basis of both human rights and constitutional property law.

The overarching question that arises with respect to democratic merit is whether the taking in question 
can be said to arise from a legitimate process of decision-making, in the pursuit of a fair and equitable outcome. It bears emphasising that in line with a more modern appreciation of the meaning of democracy and human rights, the relevant assessment under this point involves both procedural and substantive elements. Fairness in itself is a constraint on the democratic process, particularly when fundamental economic and social rights are involved. At the same time, the notion of democratic merit rightly brings procedural questions to the foreground. Indeed, it might be a weakness of Gray's original proposal that it does not single out procedural issues for special consideration.

On the one hand, it is inappropriate to reduce the takings question to a matter of administrative law. But on the other hand, the way in which the taking decision was made can often tell us much about its legitimacy, including how it stands with regard to broader notions of fairness. It is particularly important, in this regard, to inquire into the position of local owners and communities during the planning process leading up to the decision to use eminent domain. In the context of property as a human right, moreover, a stricter standard might be appropriate here, compared to that which would otherwise follow from administrative law.

Importantly, our commitment to property as a social institution requires us to take into account that the owners generally make up the group of people who will be most directly affected by any decision involving the future of their property. As such, they should normally be granted a decisive voice in decision-making processes leading up to economic development. At the same time, the social function account leaves room for recognising that this presumption in favour of emphasising the rights of owners can be defeated by the context. It is clear, for instance, that the substantive interests of absentee landlords might be limited compared to the substantive interests of local non-owners who depend more directly on the property for their livelihoods. In these cases, the social function approach allows us to recognise that a taking might have significant democratic merit, even if it is based on a form of decision-making that prioritises the interests and participation rights of non-owners.

Nevertheless, within a system based on private property rights it will always be appropriate to show caution in this regard. The presumption should always be, within such a system, that the owners are the primary stakeholders in decision-making processes involving their property. Moreover, if this presumption is defeated, it would usually point to a structural weakness of property's function within society, a weakness that should arguably be addressed by more general reforms of property, not by inflating the state's power to undermine it. If caution is not observed here, property can soon become a less secure basis on which to support local communities, including those marginalised groups that are most in need of protection from predators.%\footnote{A concrete discussion of the point at stake here, in the context of Indian law, is given in \cite{...} (``.....'').}

\noo{ Some commentators emphasise how takings should not have a redistribute purpose, describing this goal as being behind both public interest restrictions and rules that are meant to ensure full compensation. On the basis of the social function theory, we can agree that takings should not normally be used to redistribute property, because it weakens the institution of property as such. But we may also add that if redistribution is indeed required to restore property's social function, it should be implemented through property reforms that target structural imbalances, not state action that targets individual owners.

Indeed, if imbalances in the distribution of property is used to justify giving the state increased powers to interfere on a case-by-case basis, it does {\it more} than implement redistribution (which might well have considerable democratic merit). Importantly, granting such power to the government, particularly if the power is vested in the executive branch, risks undermining property as an institution. Hence, property might become [...] }

\noo{
The difficult takings cases will be those which seem suspect even though marginalised groups directly benefit, and even though none of the other points of the Gray test appear to indicate predation. Normally, such takings should be regarded as legitimate (assuming, of course, that they do not offend against the relevant positive expropriation law). However, I believe the democratic merit test rightly points to a possible exception, if such takings unduly contribute to undermining property as an institution, thereby setting the stage for future abuses. Arguably, the {\it Agri} case from South Africa, discussed in Section \ref{sec:x}, is an example of such a case.\footnote{Indeed, an assessment against democratic merit along the lines I propose here appears to provide a plausible basis on which to argue against the decision made by the Constitutional Court, c.f., the discussion in Section \ref{sec:x}. Moreover, by focusing on the systemic effect of the decision in that case, rather than the liberal rights of the affected owners, we would be in an excellent position to draw a crucial distinction between the minority and the majority in the Constitutional Court. Specifically, the democratic merit condition might well support the conclusion that while the majority opinion reflects a predatory attitude, a decision made in keeping with the minority argument (for the same concrete outcome) would have been legitimate.}

In this thesis, the cases considered will most often not involve subtleties like that of {\it Agri}. Indeed, when Norwegian waterfalls are taken for hydropower development, no one argues that this is necessary to correct past injustices or to bestow benefits marginalised groups. Quite the contrary, as I will show, these takings typically involve the opposite scenario, where property is taken from marginalised groups and handed over to more resourceful owners. In such cases, the social function theory of property suggests that there is a {\it prima facie} good reason for suspecting that the takings lack democratic legitimacy. There might be extenuating circumstances, but at least special scrutiny is in order. Plainly, the apparent lack of democratic merit is why takings like these, the quintessential economic development takings, deserve to be considered as a separate category in takings law.
}
\noo{ 

arly with the


In this thesis, however, such takings will not be the focus of attention. Rather, attention will be directed to takings where the marginalised groups own the property to start with,

One might still  misgivings about allowing economic development takings to go ahead, because they unduly inflate the power of government. 

Property has merit because it offers such protection, and for this reason it should be protected, as an institution. Property reform should aim to strengthen property, not weaken it, and exercising the power of eminent domain on a case-by-case basis is {\it not} the same as implementing a property reform.

As with the other points, assessing the democratic merit of a taking requires taking into consideration both procedural and substantive aspects. First, we should consider how the decision to use eminent domain was made, and if the owners in particular were granted a say proportional to their stake in the matter. Second, we should consider the effect of the taking on future decision-making processes involve the property rights in question. The mere fact that eminent domain serves to take property from the many and give it to the few, thereby making the property regime less egalitarian, is an argument for increased scrutiny of legitimacy.

}

By itself, however, a legitimacy test cannot make property a more secure basis for promoting good outcomes in cases when the public desires economic development. What the Gray test provides is a list of possible symptoms to look our for when attempting to diagnose a suspected case of eminent domain abuse. Hopefully, this can help flag problems and limit damages, but it can not be regarded as a solution, especially not in cases when the public's apparent desire for economic development is a genuine reflection of a democratic commitment. 

In short, after diagnosing a lack of legitimacy, the question becomes how to find a cure, preferably without harming the patient, i.e., the democratic system. In the next section, I consider this challenge in more depth, premised on the idea that there is a need for alternatives to eminent domain in cases when the collective wishes to take decisive steps to promote economic development on privately owned land.

\section{Alternatives to Takings for Economic Development}\label{sec:lad}

As mentioned briefly in Chapter 2, the work of Ostrom and others on common pool resources suggests that sustainable resource management can often be better achieved through local self-governance than through markets or states.\footnote{See generally \cite{ostrom90}. For a recent exposition of the main ideas, placing the work in a broader academic context, see the revised version of Ostrom's Nobel Lecture, \cite{ostrom10}.} The connection between this work and property theory is highly interesting, and has been explored in some recent work, particularly by US legal scholars.\footnote{See generally \cite{rose11,fennel11}.} As these scholars have observed, the connection can be made at a very high level of generality. Indeed, in a democracy, property as such has a kind of (partial) commons structure, since property as an institution depends on the collective choices we make regarding the legal order.\footnote{For similar observations, see \cite[51]{rose90}; \cite[577]{heller01}.} In cases when property is made subject to eminent domain, this perspective becomes particularly salient, since then the collective explicitly withdraws its backing for the rights of the owner, in favour of collective decision-making about the future of the property in question.\footnote{A common pool resource is typically identified by the fact that exclusion is difficult or costly, while use can cause depletion (and hence should be limited), see, e.g., \cite[57]{ostrom10b}. Hence, the mere fact that some property is apt to be regulated, or taken, by the collective, demonstrates that property has common pool characteristics (although these might be imposed by the polity, rather than arising from the nature of the underlying good).} Moreover, in case of an economic development taking, the property in question typically pertains to land or some other natural resource, which invariably form part of a larger resource system with some common pool characteristics.\footnote{See, e.g., \cite[16]{fennel11} (``we are {\it always} operating at least partially within a commons of some sort''). I also mention Smith's notion of a ``semicommons'', used to describe settings where common pool arrangements for resource management interact with individual property rights, see generally \cite{smith00,smith02}.}

%The social function theory in particular, compels us to consider the place of private property in such contexts. To complement these observations, it is crucial to note that 
Importantly, to designate something as a common pool resource does not in any way imply that the resource in question is open-access or that it is held as a form of common property, a public trust, or under some other legal construction moving away from the sphere of private property.\footnote{See, e.g., \cite[58]{ostrom10b}.} Perhaps more controversially, designating something as a common pool resource does not in any way imply that the resource {\it should} be removed from this sphere.\footnote{See \cite[58]{ostrom10b} (``there is no automatic association of common-pool resources with common-property regimes -- or, with any other particular type of property regime'').} According to Ostrom and Hess, the appropriate property regime for a given common pool resource is a pragmatic question that depends on the circumstances.\footnote{See \cite[58]{ostrom10b}.}
%The distinction between common pool resources and other kinds of potentials inhering in land and other typical objects of property is not sharp. The characteristic features of such resources is that (1) excluding others from access to them is costly and (2) their beneficial potential may be depleted by use.\footnote{See, e.g., \cite{ostrom07}.}
%As observed by X and others, this means that almost {\it any} property can conceivably be understood as pertaining to a common pool resource. After all, exclusion is a key aspect of property, which is almost always potential costly.\footnote{If exclusion can be ensured at no cost, the concept of ownership is also typically not taken to apply, as it appears redundant. A typical example would be the feelings and thought of an individual, or the emotional attachment that exists between two people.} Moreover, exclusion as such is hardly required unless the resource itself is susceptible to some kind of depletion resulting from use. Indeed, as noted by Carol Rose and others, it is possible to view virtually any resource that is a candidate for property as a common pool resource.
%The theories of resource management developed by political economists working in the tradition of Ostrom usually do not have much to say about the role of private property. This might be due to the continued influence of a narrow entitlement-based property narrative. On this narrative, property as a legal category can come to appear rather insulated from broader social and political concerns. By contrast, a social function theory of property makes it very natural to consider the role that property can play in the development of local frameworks for sustainable resource governance. %Moreover, while private ownership is sometimes thought to stand in opposition to the commons, it has been noted by several scholars that a property regime is in itself a way to manage common resources, by recognising a special link between individuals and their possessions.
%In the context of taking of property for economic development, the link between property and resource management is particularly clear. Economic development takings reflect a special approach to resource management whereby the state regards it as desirable to remove existing property rights from the equation entirely. When seen in this light, the question of legitimacy that arises also pertains to the institutional consequences of an economic development taking. Is the taking conducive to sustainable resource management? 
It should be noted, however, that this neutral position on the relationship between property and common pool resources is premised on a bundle of rights understanding.\footnote{See \cite[59]{ostrom10b}.} Potentially, a more ambitious theory of property could suggest a different perspective. Specifically, the question arises as to how theories of common pool resource management relates to the social function account. This is a particularly interesting avenue for future work, as it could shed light on the normative stance that private property can be a good basis for sustainable self-governance, at least when backed up by a human flourishing account of what private property should be.

In this thesis, I will limit myself to noting how the link between property and theories of commons governance provides a possible route towards an institutional perspective on how to solve legitimacy problems associated with economic development takings.  %Intuitively, such a perspective already suggests itself by the work done in this and the previous chapter, which has emphasised the position of the local community and the function of property as an anchor for democracy. It remains to work out in further detail what exactly such a perspective has to offer in relation to the issue of economic development takings.

To make progress in this regard, it will be useful to first briefly consider one of the most important theoretical legacies of Ostrom's work, namely a list of eight design principles that she formulated on the basis of empirical studies.\footnote{See \cite[90]{ostrom90}.} These principles were formulated because they seemed to be particularly crucial in ensuring good governance at the local level, and have since been supported by a growing body of empirical evidence.\footnote{See \cite{cox10} (the authors also suggest splitting some of the original principles in two parts, resulting in a slightly more fine-grained list, not needed in this thesis). } In brief, the so-called CPR principles are the following:

\begin{enumerate}
\item {\bf Well-defined boundaries:} There should be a clearly defined boundary around the resource in question, and a clear distinction should exist between members of the user community, who are entitled to access the resource, and non-members, who may be excluded. This will internalise the costs of resource exploitation and other externalities, ensuring that proper incentives for sustainable management arise within the community of resource users.\footnote{Importantly, the possibility of excluding non-members marks a distinction between open-access resources and common pool resources, where the latter appears much less susceptible to a commons tragedy than the former, because externalities are internalised to a clearly defined community. See \cite[91-92]{ostrom99}.}
\item {\bf Congruence between appropriation and provision rules and local conditions:} Management principles should be flexible and responsive to changing local conditions. Moreover, management practices should be anchored in the economic, social, and cultural practices prevalent at the local level. In addition, the individual benefits should generally exceed the individual costs associated with membership in the community of users, and collectively managed benefits should be distributed fairly among community members.\footnote{See \cite[92]{ostrom90}.}
\item {\bf Collective-choice arrangements:} The individual members of the user community should have an opportunity to participate in decision-making processes regarding the rules that govern the user community and the resource management. In addition to securing fairness and legitimacy, this will enhance the quality of the decision-making, as the users themselves have first-hand knowledge and low-cost access to information about their situation and the state of the resource in question.\footnote{See \cite[93]{ostrom90}.}
\item {\bf Monitoring:} There should be mechanisms in place to ensure that the behaviour of users is monitored for violations of management rules. To increase efficiency, monitoring should be locally organised. Moreover, to ensure local responsiveness and legitimacy, individuals acting as monitors should themselves be members of the user community or in some way answerable to this community.\footnote{See \cite[94-100]{ostrom90}.}
\item {\bf Graduated sanctions:} There should be an effective system in place for penalising violations of user community rules. These penalties should be graduated so that more severe or repeated violations are sanctioned more severely than minor or one-time transgressions.\footnote{See \cite[94-100]{ostrom90}.}
\item {\bf Conflict-resolution mechanisms:} The user community should be endowed with low-cost procedures for conflict resolution. These procedures should be sensitive to local conditions, to ensure local legitimacy.\footnote{See \cite[100-101]{ostrom90}.}
\item {\bf Minimum recognition of rights:} The user community should be protected from interference by external actors, including government agencies. As a minimum, the existence of local institutions and the right to self-governance should be recognised and respected by external government authorities.\footnote{See \cite[101]{ostrom90}.}
\item {\bf Nested enterprises:} There should be vertical integration between local, small-scale, management institutions and larger institutions aimed at protecting and furthering non-local interests. This integration should be based on the minimum recognition of rights mentioned in the previous point. Furthermore, it should provide a template for integrated decision-making about larger scale issues, where local competences are employed incrementally in more general settings, involving also institutions working on behalf of municipalities, regions, states and the international community. Local institutions for resource management should not only be respected by such larger scale structures, they should also feed into larger scale decision-making and be called to respond to greater community needs.\footnote{See \cite[101-102]{ostrom90}.}
\end{enumerate}
\noo{
There is an interesting link between the principles described here and the property norms discussed in Chapter 1. There the focus was on norms arising in relation to the social function of property, particularly those related to the idea of human flourishing through membership in a community. The main argument was that property plays an important constitutive role in this regard, as the rights and obligations inhering in private property has an important function as a promoter of collective action and equity at the local level, as well as a protector of local communities against exploitation and undue interference from external actors.

The role that property can play in this regard is linked to the local institutions for resource management that property rights can help sustain. Under individualistic, entitlements-based, theories of property, one might worry that institutions for collective management, even at the local level, will have to be rendered very thin in order not to conflict with the rights of owners. However, on the social function theory, this should no longer be a concern. On this theory, there is no opposition between strong property rights and strong community institutions for collective management of local resources, even if these resources are held as private property. The inclusion of non-owners as full members of the user community is also consistent with the social function theory of property and its emphasis on obligations as well as rights.\footnote{More subtly, limitations on the owners' rights imposed on the basis of property's social function do not necessarily deprive owners of their protected status as property holders. Hence, property does not lose its power to protect the community, merely because the rights of individual owners is limited by community interests. For instance, even if (local) institutional arrangements limits the owners' right to alienate their property, this does not mean that they are no longer owners. Private property with restrictions on alienation can provide the legal basis for local resource management, without removing us from the realm of private property. This is seen, for instance, in Norway, as discussed in Section x of Chapter 2. By contrast, Ostrom and Hess adopt a more narrow view on ownership, where the right to alienation is considered so basic that resource users who lack this right are not considered owners at all, but merely proprietors. See \cite[...]{ostrom07}.}

The design principles listed above further underscores how well the human flourishing theory of property coheres with the theory of common pool resources. Specifically, the property concept endorsed by human flourishing theorists promises to deliver a straightforward and robust implementation of several key design features. Recognising something as property will tend to deliver clearly defined boundaries and a minimal recognition of rights. Moreover, on a human flourishing account, property is also required to deliver congruence between burdens and benefits among property users, particularly at the community level. Finally, property provides a template for a form of organisational nesting where larger scale institutions may be awarded regulatory competences, but must still be prevented from usurping the power to manage local resources as {\it de facto} proprietors. In short, it appears that property alone can take us halfway towards a well-functioning CPR institution.

The final step, pertaining specifically to the institutional framework for collective management, might well place significant limits on the ``despotic dominion'' of individual owners. However, on the social function theory, such limitations might be understood as means to fulfil property's purpose, rather than as an attack on the rights of owners. More generally, the social function understanding of property makes it possible to sustain a property narrative by regarding resource users as owners even if local institutional arrangements mean that they are not in possession of some of the sticks typically found in liberal property bundles. Arguably, local people should still be entitled to recognition as owners of their resources, even if the form of ownership in question does not conform to the standard liberal expectation of what private property looks like. If this is true, it would also appear that abrogation of private property is neither required nor desirable, in order to secure sustainable management of resources at the local level.

%This is likely to offer local people a significantly greater level of protection than they could otherwise expect, also under human rights provisions targeting their social and economic condition.

This observation connects Ostrom's theory of common pool resources to the human flourishing theory of property at a high level of abstraction, suggesting that the two are in fact mutually conducive to one another. There are two different ways in which this also connects concretely to the issue of economic development takings.} 
There are at least two interesting connections between self-governance principles such as these and the issue of economic development takings, especially as that issue is approached in this thesis, on the basis of the social function theory of property. First, one may observe that when economic development takings appear to lack legitimacy with respect to social functions, this is typically also an indication that the surrounding framework for resource management is not well-designed. In particular, it appears that the Gray test closely tracks many of the design principles proposed by Ostrom.

%It is clear that organising economic development is a crucial aspect of resource management. Moreover, the use of eminent domain always takes place within a management context that can in turn be assessed against the design principles proposed for common pool resources. Importantly, this links up with the discussion in Chapter 1, where it was argued that the assessment of legitimacy of takings should be broad and contextual. This discussion culminated in the presentation of the Gray test, which focuses on social and economic aspects when guiding our judgement of when a taking is legitimate. Interestingly, it seems that whenever this test leads us to consider a taking as an act of predation, there is good reason to think that the accompanying framework for resource management violates the design principles for common pool resources.

For instance, consider the balance of power between the owners and beneficiaries of a taking, the first point to consider according to the Gray test. When a taking fails on this point, doubts naturally arise also with regard to the underlying framework for resource management, particularly aspects pertaining to the recognition of local rights, the adequacy of collective-choice arrangements, and the congruence between appropriation, provision and local conditions. If property is taken by powerful actors, chances are that these actors are not representative of local community interests. Moreover, takings characterised by an imbalance of power typically indicate that the government is in fact quite unwilling to recognise the rights of local people, even when these rights are formally recognised as property rights.

%This is illustrated very clearly by the US cases, where economic development takings tend to result in controversy precisely when the taker is an external commercial actor who poses a threat to a local community. In these cases, taking property serves to weaken the link between resource management and local conditions.\footnote{In some situations, such as in {\it Poletown}, the taking can consists in removing whole communities, replacing the properties found within it by commercial property, owned and managed by a single large enterprise.} 
By contrast, the situation might be different if it involves a taking that is not suspect according to the Gray test. For instance, if property is taken from absentee landlords and given to local land users in order to facilitate development, this might be an honest attempt at setting up a management framework that complies with CPR principles. In such a case, one would also not expect the balance of power between owners and takers to point towards abuse.

%Indeed, it seems that many, possibly all, the conditions of the Gray test tend to track corresponding design principles for CPRs. For instance, the environmental impact test, whereby an economic development is suspect if it demonstrates disregard for nature, appears closely related to both the issue of monitoring of resource abuse and graduated sanctions for bad behaviour. More generally, the regulatory effect of a suspect economic development taking also closely tracks these points.

The second link between CPR design and economic development takings is arguably even more interesting. This link becomes apparent as soon as we shift attention away from diagnosing a lack of legitimacy towards coming up with alternative management principles that can restore it. Specifically, work done on local governance of common pool resources point to an {\it alternative} way of approaching the goal of economic development in cases that might otherwise result in the use of eminent domain. 

This has not received much attention in the literature so far. One notable exception, discussed in depth in the following subsection, is the work of Heller, Dagan and Hills.\footnote{The work of Lehavi and Licht also deserves a brief mention, even though it focuses on compensation rather than alternatives to eminent domain. The reason is that this work relies on proposing a novel institution that also touches on issues related to self-governance. In particular, Lehavi and Licht propose that post-taking, collective, price bargaining should be carried out on behalf of owners by a Special Purpose Development Company, in an effort to give them a chance to get their share of the commercial benefit arising from development. See \cite{lehavi07}. For a more in-depth discussion of this proposal, and the compensatory approach to economic development takings more generally, see \cite{dyrkolbotn15}.} Looking at their work will serve to make the abstract discussion above more concrete, and will set the stage for a comparison between their proposal and solutions that can be facilitated by the system of land consolidation presented in Chapter 5.

\subsection{Land Assembly Districts}\label{sec:lad}

In an article from 2001, Heller and Dagan considered the connection between CPR design and overarching (liberal) property values.\footnote{See \cite{heller01}.} From this, they arrived at a proposal for what they call a ``liberal commons'', which adds some design constraints rooted in a desire to protect individual autonomy and minority rights. In particular, they emphasise the value of exit, the opportunity for members of the governance structure to alienate their share in the commons resource (conceived of as a property right).\footnote{See \cite[567-572]{heller01}.} The right of owners to leave the collective is thought of as a safety mechanism, to prevent failing institutions from trapping its members in a state of oppression. This, it is argued, is an important overarching design constraint, described as a ``liberal'' idea, that should complement the other design principles for local management of common pool resources.\footnote{Despite their commitment to protect the right of exit, Heller and Dagan are also aware of the destabilising effect exit can have on an otherwise well-functioning institution. To address this, they discuss additional mechanisms, such as rights of first refusal, that can ensure that exit does not prove too disruptive to the local collective, as long as a sufficient number of members chose to remain. See \cite[596-702]{heller01}.}

In a later article, responding to the {\it Kelo} controversy, Heller and Hills build on the idea of the liberal commons by proposing a novel approach to the takings issue, consisting of a proposal for a new institutional framework that can facilitate land assembly for economic development. The core idea is to introduce {\it Land Assembly Districts} (LADs), institutions that will enable property owners in a specific area to make a collective decision about whether or not to sell the land to a developer or a municipality.\footcite[1469-1470]{heller08} The idea is that while anyone will be able to propose and promote the formation of a LAD, the official planning authorities and the owners themselves must consent before it is formed.\footcite[1488-1489]{heller08} Clearly, some kind of collective action mechanism is required to allow the owners to make such a decision. 

%Importantly, it is meant to replace eminent domain altogether, in certain kinds of cases. The goal is to ensure democratic legitimacy while also setting up a template for collective decision-making that will prevent inefficient gridlock and holdouts. 

Hiller and Hill suggest that voting under the majority rule will be adequate in this regard, at least in most cases.\footnote{See \cite[1496]{heller08}. However, when many of the owners are non-residents who only see their land as an investment, Heller and Hills note that it might be necessary to consider more complicated voting procedures, for instance by requiring separate majorities from different groups of owners. See \cite[1523-1524]{heller08}.} How to allocate voting rights in the LAD is given careful consideration, with Heller and Hills opting for the proposal that they should in principle be given to owners in proportion to their share in the land belonging to the LAD.\footnote{See \cite[1492]{heller08}. For a discussion of the constitutional one-person-one-vote principle and a more detailed argument in \isr{favour} of the property-based proposal, see \cite[1503-1507]{heller08}.} Owners can opt out of the LAD, but in this case, eminent domain can be used to transfer the land to the LAD using a conventional eminent domain procedure.\footcite[1496]{heller08}

Heller and Hills envision an important role for governmental planning agencies in approving, overseeing and facilitating the LAD process. Their role will be most important early on, in approving and spelling out the parameters within which the LAD is called to function.\footcite[1489-1491]{heller08} While it is not discussed at any length, the assumption appears to be that the planning authorities will define the scope of the LAD by specifying the nature of the development it can pursue in quite some depth. Hence, despite the overarching goal of self-governance, the power of the planning authority still appears significant under the LAD proposal as it currently stands.
\noo{
A possible challenge that arises, not discussed by Heller and Hills at any length, is that the scope of the LAD needs to be broad enough to allow for meaningful competition and negotiation after LAD formation. To achieve this might be difficult, particularly in light of incentives to make the outcome of the LAD process more predictable. Indeed, both governments, initiating developers, and landowners eager for development might want to ensure that the scope of the LAD is defined narrowly enough to give confidence that zoning permissions will not later be denied. In addition, there is the obvious nefarious incentive that some actors might have to ensure that a specific development is chosen. In light of this, LAD regulation is needed to ensure a balanced approach to the issue of how the initial development possibility should be defined, and to what extent this definition should limit the authority of the LAD to choose alternative projects.
}

If the owners do not agree to forming a LAD, or if they refuse to sell to any developer, Heller and Hills suggest that the government should be precluded from using eminent domain to assemble the land.\footcite[1491]{heller08} This is a crucial aspect of their proposal that sets the suggestion apart from other proposals for institutional reform that have appeared after {\it Kelo}. A LAD will not only ensure that the owners get to bargain with the developers over compensation, it will also give them an opportunity to refuse any development to go ahead. Hence, the proposal shifts the balance of power in economic development cases, giving owners a greater role also in preparing the decision whether or not to develop, and on what terms. Hence, the LAD proposal promises to address the democratic deficit of economic development takings, without failing to \isr{recognise} that the danger of holdouts is real and that institutions are needed to avoid it.

There are some problems with the model, however. First, it seems that planning authorities might have an incentive to refuse granting approval for LAD formation. After all, doing so entails that they give up the power of eminent domain for the land in question. For this reason, Heller and Hills propose that a procedure of judicial review should exist whereby a decision to deny approval for LAD formation can be scrutinized.\footcite[1490]{heller08} However, the question then arises as to how deferential  courts should be in this regard, echoing the conundrum that engulfs the safeguard intended by the public use restriction. Presumably, one would want the courts to strictly scrutinise LAD rejections, to instil that LADs should normally be promoted. However, would the courts be comfortable providing such scrutiny, also against a government body claiming that the ``public interest'' speaks against LAD formation? This would likely depend on the exact formulation and spirit of the LAD-enabling legislation. To work as intended, some sort of presumption in favour of LAD approval appears to be in order, but this in turn can have the effect of making it easier for powerful landowners to abuse the LAD system, e.g., by pushing through LADs that enable them to impose their will on other community members.

This worry is related to a second possible objection against the LAD proposal, concerning the practicalities of the process leading up to the LAD's decision on whether or not to accept a given offer. Is it possible to organise such a process in a manner that is at once efficient, inclusive and informative, without making it too costly and time consuming? Here Heller and Hills envision a system of public hearings, possibly \isr{organised} by the planning authorities, where potential developers meet with owners and other interested parties to discuss plans for development.\footnote{See \cite[1490-1491]{heller08}. It might also be necessary for the planning authorities or other government agencies to take on some responsibilities with respect to providing guidance and assistance to less resourceful members among the owners.} The process envisioned here would resemble existing planning procedures to such an extent that additional costs could hopefully be kept at a minimum. 

The significant difference would concern the relative influence of the different actors, with the   owners as a group receiving a considerable boost as a result of the LAD. Rather than being sidelined by a narrative that sees the use of eminent domain as the culmination of planning, the owners are now likely to occupy center stage throughout, as they now will have the final say on whether or not the development will go ahead.

From this, however, arises the question of how the interests of other locals, without property rights, will be protected. Heller and Hills assumes that local non-owners will also be represented during the stages leading up to the LAD's final decision, but their role in the process is not clarified in any detail.\footcite[1490-1491]{heller08} This raises the worry that LADs might undermine local democracy by giving property owners a privileged position with respect to policy questions that should be decided jointly by all members of the community. The severity of this risk depends heavily on the circumstances. In a context of egalitarian property ownership and sensible government regulation of land uses and LAD operations, the risk should be minimal. In principle, the local anchoring that LADs provide should also benefit non-owners, by bringing the decision-making process closer and making it more easily accessible for the people most directly affected, including non-owners. Moreover, if some members of the local community remain marginalised, this should possibly be regarded as a regulatory failure or a reflection of underlying inequality in society, not a shortcoming of the LAD proposal as such. In these cases, a reasonable approach might even be to {\it expand} the function of LADs, by granting voting rights to a larger class of local property dependants, not only formally titled owners.\footnote{The important invariant to maintain, I believe, is that the locally anchored institution should be the active, invested, agent, while more centralised and/or expert-dominated government bodies should act as passive, impartial, regulators. In the processes leading to economic development takings, this equation is typically reversed, with government bodies and commercial companies being the active agents, while the owners and the local community are the passive agents whose property rights and dependencies place some nominal limits on the authority of other parties (limits which, due to the weakness of owners as a group, tend to be easily disregarded).}

However, the LAD proposal raises some highly problematic issues pertaining to the proposed mechanism of collective decision-making. As Kelly points out in a commentary, the basic mechanism of majority voting is deeply flawed.\footcite{kelly09} For instance, if different owners value their property differently, majority voting will tend to \isr{disfavour} those with the most extreme viewpoints, either in \isr{favour} of, or against, assembly. If these viewpoints are assumed to be non-strategic and genuine reflections of the welfare associated with the land, the result can be inefficiency. In short, the problem is that a majority can often be found that does not take due account of minority interests. 

For instance, if a minority of owners are planning development on their own land, and this conflicts with some LAD proposal targeting a larger area, the minority might find it difficult to defend themselves against the force of the LAD. Indeed, such a minority might effectively loose the battle for their property as soon as a LAD is formed, if the development description underlying LAD formation is incompatible with the kind of development they wish to pursue. For such owners, a presumption in favour of LAD formation might prove highly disadvantageous.\footnote{Of course, one might imagine these landowners opting out of the LAD, or pursuing their own interests independently of it. However, they are then unlikely to be better off than they would be in a no-LAD regime. In fact, it is easy to imagine that they could come to be further \isr{marginalised}, since the existence of the LAD, acting `on behalf of the owners', might detract from any dissenting voices on the owner-side.} %This could become particularly inefficient in cases when the alternative development is more socially desirable than the development sketched during LAD formation. In such cases, LAD formation will not improve the quality of the decision to develop, since it pushes the decision-making process into a track where those interests that {\it should} prevail are voiced only by a marginalised minority inside the new institution.

%More generally, the lack of clarity regarding the role of LADs in the planning process is a problem. If LAD formation as such can be used to privilege a specific development scheme over possible alternatives, developers would have a great incentive to form a LAD as early as possible. 
Indeed, developers might come to rely on LADs to push through {\it de facto} condemnations of property, through a procedure that leaves minorities less protected than the traditional takings process. Indeed, it would be theoretically possible for any landowner to use a LAD to condemn any neighbouring property smaller than their own. Eventually, a whole community might be taken over by one or a few powerful landowners, through a sequence of appropriately designed LADs and development projects. %At the same time, it is easy to acknowledge that problematic situations may arise from wide LAD powers, for instance if a majority forms in \isr{favour} of a scheme that involves razing only the homes of the minority, maybe on the rationale that these are the most blighted properties.

Despite these worries, the ideal of the LAD proposal is clearly stated and highly attractive. LADs should help to establish self-governance for land assembly and economic development. In particular, Heller and Hills argue that LADs should have ``broad discretion to choose any proposal to redevelop the \isr{neighbourhood} -- or reject all such proposals''.\footcite[See][1496]{heller08} As they put it, two of the main goals of LAD formation is to ensure ``preservation of the sense of individual autonomy implicit in the right of private property and preservation of the larger community's right to self-government''.\footcite[See][1498]{heller08} The problem is that these ideals turn out to be at odds with some of the concrete rules that Heller and Hills propose, particularly those aiming to ensure good governance of the LAD itself.

%echo many of the ``corporate governance''-ideas that also feature heavily in Lehavi and Licht's proposal. Indeed,

In relation to the governance issue, Heller and Hills emphasise, in direct contrast to their comments about ``broad discretion'' and ``self-governance'', that ``LADs exist for a single narrow purpose -- to consider whether to sell a neighborhood''.\footcite[See][1500]{heller08} This is a good thing, according to Heller and Hills, since it provides a safeguard against mismanagement, serving to prevent LADs from becoming battle grounds where different groups attempt to co-opt the community voice to further their own interests. As Heller and Hills puts it, the narrow scope of LADs will ensure that ``all differences of interest based on the constituents' different activities and investments, therefore, merge into the single question: is the price offered by the assembler sufficient to induce the constituents to sell?''.\footcite[1500]{heller08}

This means that there is a significant internal tension in the LAD proposal, between the broad goal of self-governance on the one hand and the fear of \isr{neighbourhood} bickering and majority tyranny on the other. Indeed, it is hard to see how LADs can at once have both a ``narrow purpose'' as well as enjoy ``broad discretion'' to choose between competing proposals for development. If such discretion is granted to LADs, what prevents special interest groups among the landowners from promoting development projects that will be particularly \isr{favourable} to them, rather than to the landowners as a group? What is to prevent landowners from making behind-the-scene deals with \isr{favoured} developers at the expense of their \isr{neighbours}? It might be difficult to come up with rules that prevent mechanisms of this kind, without also making substantive self-governance an impossibility.

If a LAD is obliged to only look at the price, this might prevent abuse. But it will not give owners broad discretion to consider the social functions of property when choosing among development \isr{proposals}. %Effectively, it will render LADs as little more than a variant of SPDCs, where the owners are awarded an extra bargaining-chip, namely the option to refuse all offers. 
In my view, it is undesirable to restrict the operations of LADs in this way. It is easy to imagine cases where competing proposals, perhaps emerging from within the community of owners themselves, will be made in response to the formation of a LAD. Such proposals may involve novel solutions that are superior to the original development plans, in which case it is hard to see any good reason why they should not be taken into account, even if they are proposed by a minority. Moreover, it is hard to see why they should be disregarded simply because they are less commercially attractive, or because the  developer interested in pursuing such a proposal cannot offer the highest payment to the owners. In the end, the decision that the LAD makes concerns the future of the community as a whole. This is not an exercise in profit-maximization, and there are good reasons to believe that LAD regulation should encourage a broad perspective, not enforce a narrow one.

However, when it comes to the details, Heller and Hills seem quire adamant that the degree of self-governance needs to be limited in favour of strict regulation to reduce the risk of LAD abuse. In particular, they argue that ``LAD-enabling legislation should require especially stringent disclosure requirements and bar any landowner from voting in a LAD if that landowner has any affiliation with the assembler''.\footcite{heller08} Here, the notion of self-governance is made even thinner, as owners will effectively be barred from using LADs as a template for gaining the right to participate in development projects themselves.

Moreover, new questions arise. For one, what is meant by ``affiliation''? Say that a landowner happens to own shares in some of the companies proposing development. Should they then be barred from voting? If so, should they be barred from voting on all proposals, or just those involving companies in which they are a shareholder? If the answer is yes, how can this be justified? Would it not be easy to construe such a rule as discrimination against landowners who happen to own shares in development companies? On the other hand, if the landowner in question is allowed to vote on all other proposals, would it not be natural to suspect that their vote is biased against assembly that would benefit a competing company? Or what about the case when some of the landowners are employed by some of the development companies? Should such owners be barred from voting on proposals that could benefit their employers? This seems quite unfair as a general rule. But in some cases, employment relations could play a decisive factor in determining the outcome of a vote. This might happen, for instance, if an important local employer proposes development in a \isr{neighbourhood} where it has a large number of employees. Heller and Hills give no clear answer to the questions arising in this regard, and at this point, the circle has in some sense closed in on their proposal. Indeed, just as courts today struggle with the ``public use'' requirement, it seems that the proposed ``affiliation'' criterion for depriving someone of their voting rights would provide a very shaky basis for judicial review. 

More generally, it seems that how to best organise a LAD remains an open problem. The challenge is to ensure that LADs deliver a real possibility of self-determination, while also ensuring good governance and protection against abuse. That it remains unclear how to do this is acknowledged by Hiller and Hills themselves, who point out that further work is needed and that only a limited assessment of their proposal can be made in the absence of empirical data.\footnote{See \cite[]{heller08}.} Later in the thesis, I will shed light on this challenge when I consider the Norwegian framework for land consolidation. This framework can be looked at as a sophisticated institutional embedding of many of the central ideas of LADs. In particular, I will discuss how Norwegian land consolidation can be employed in cases of economic development, and how it is increasingly used as an alternative to expropriation in cases of hydropower development. This will allow me to shed further light on the issues that are left open by Heller and Hills' important article.

\section{Conclusion}\label{sec:conc2}

In this chapter, I have given a more in-depth presentation of the legitimacy question that arises for economic development takings. I considered existing approaches and argued in favour of an approach that combines procedural and substantive standards, with the intention being that the latter should be used also to assess the fairness and democratic merit of the decision-making as such, not merely the outcome. I argued that such an institutional approach to fairness has started developing at the ECtHR, as a consequence of its development of the pilot judgement framework for assessing cases that might indicate systemic problems at the state institutional level.

I went on to provide a possible heuristic for assessing the legitimacy of economic development takings,  in a manner that is compatible with an institutional fairness perspective. I started from a list of legitimacy indicators provided by Gray, and added three new ones, based the work done in this and the previous chapter. The resulting heuristic, the Gray test, should be able to identify cases of eminent domain abuse, particularly those that offend against social functions of property at the institutional level, rather than merely the financial entitlements of owners.

Following up on this, I considered the question of how to increase legitimacy without giving up on the idea that the collective should be entitled to push adamantly for economic development on private land. I argued that the work done by Elinor Ostrom and others on common pool resources provide a suitable starting point for making institutional proposals in this regard. Specifically, I briefly presented her design principles for local self-governance institutions, which I believe can be used as a starting point also for designing procedures to replace eminent domain for economic development.

Finally, I considered a proposal that has already been made along these line, namely the idea of Land Assembly Districts due to Heller and Hills. I analysed this proposals in some detail, pointing out some shortcomings, including a deep tension between the overarching goal of self-governance and the danger of majority and/or elite tyranny. Moreover, I noted that proposals such as these are hard to make at the theoretical level; the appropriate institution for self-governance needs to be attuned to local conditions. Indeed, this is one of the key design principles proposed by Ostrom. 

This observation marks the end of the first part of the thesis. In the next part, I will consider the case of Norwegian hydropower specifically. This will lead to an analysis of legitimacy of takings for this purpose along the lines of the Gray test, as well as a case study of an already operational alternative for self-governance in place of eminent domain. In this way, the second part will aim to shed light on key aspects of the theory developed in the first part, while exploring further the idea that social functions run as a common thread through individual property rights.

%Importantly, the Court in Strasbourg cannot be a constitutional court for all its nation states -- plurality must be accepted and embraced, along with deference to national bodies. But what this also means is that the Court will not be as easily guided by the invisible hand of national tradition, precedent, and dogma, not even when assessing core systemic features of great importance to the elites of the signatory states. Because of its special position, the Court must look directly at substantive ideas inherent in the ECHR. If, as such, it becomes a political body, it would still arguably be a useful counterweight to more sinister ones, such as national courts. On a more optimistic note, the new developments at the ECtHR set the stage for a potentially powerful mix of deference and truly independent scrutiny, targeting the very foundation of the decision-making within the signatory states, without thereby undermining the value of democracy.

\noo{

when conducting this kind of assessment.
I began by noting that the issue is particularly pressing for land users that are not regarded as bringing about economic growth. Hence, I argued that the issue is closely related to that of land grabbing, which is currently receiving much attention, both academic and political. Under the social function understanding of property there is in principle no difference between protecting property rights arising from formal title and property rights arising from use. That said, special issues arise in the latter case, not least because it is unclear how the law should deal with rights resulting from cultural practices that western property regimes are not designed to handle. In addition, I noted that special issues related to poverty and basic necessities such as food and water arise with particular urgency in relation to land grabbing.

The nature of my case study makes it natural for me to focus on traditional western systems of property law. Hence, I went on to discuss how economic development takings are dealt with in such legal systems, focusing on Europe and the US respectively. For the case of Europe, this assessment was made more difficult by the fact that the category is not an established part of legal discourse. However, by looking to England for concrete examples, I noted that such cases do arise and that they are increasingly seen as controversial.

I then went on to consider the property protection offered by P1(1) of the ECHR, and how it is applied by the Court in Strasbourg. I zoomed in on those \isr{aspects} that I believe to be the most relevant for economic development takings. While I noted that this category has yet to be discussed by the ECtHR, I argued that a recent shift in the Court's property adjudication is suggestive of the fact that it would likely approach such cases similarly to how Justice O'Connor approached {\it Kelo}. In particular, I noted how the Court has recently adopted a stricter standard of assessment. This standard, I argued, is \isr{characterised} primarily by increased sensitivity to systemic imbalances causing alleged P1(1) violations. Hence, to regard economic development takings as a special category appears to fit well with recent jurisprudential developments at the Court in Strasbourg.

I went on to consider US sources on economic development takings, noting that the issue has receive an extraordinary amount of attention in recent years. I adopted an historical approach to the material, by tracing the case law surrounding the public use restriction in the fifth amendment to the US constitution, which was much debated even before the specific issue of economic development takings rose to prominence. I focused particularly on case law developed by state courts, and I argued that it shows great sensitivity to the need for contextual assessment. Indeed, it seems that many state courts originally adopted an implicit social function view of property when assessing such cases.

I then looked at the history of Supreme Court adjudication of public use cases. I noted that the doctrine of deference was developed early on, but that it was initially directed mainly at state courts. In fact, I showed that the Supreme Court itself explicitly approved the contextual and in-depth approach these courts relied on when dealing with the legitimacy issue.

The shift, I argued, came with {\it Berman}, in which the Supreme Court adopted a deferential doctrine that was directed specifically at the state legislature.\footcite{berman54} This was quite a dramatic departure from the Court's previous attitude towards state takings. Moreover, it was almost entirely backed up by precedent set in cases when {\it federal} takings had been ordered by Congress.

I went on to consider the fallout of {\it Berman} at state level, which culminated with the infamous {\it Poletown} case. This case prompted wide-spread accusations of eminent domain abuse and thus set the stage for {\it Kelo}.

After completing the historical overview, I went on to consider the literature after {\it Kelo}. I expressed particular support for those responses that focus on the need for {\it institutional} reform, to address  dangers that Justice O'Connor pointed to in her minority opinion. As a shorthand, I proposed referring to the mechanisms she identified as the {\it democratic deficit} of economic development takings. 

I then gave a thorough presentation of a recent reform suggestion that might help address this deficit, namely the proposal for Land Assembly Districts (LADs) made by Heller and Hills. According to the motivation behind this proposal, local communities should be entitled to greater self-governance in economic development scenarios, organised through LADs. Importantly, this proposal \isr{recognises} the need for a mechanism to avoid inefficient and socially harmful gridlock due to holdouts among unwilling owners. Instead of eminent domain, however, a different mechanism is proposed, namely that of a majority decision made by a land assembly district.

I pointed out some problems and seeming inconsistencies in the proposal, particularly regarding the lack of clarity regarding the exact role LADs are supposed to play during the planning process. I argued that while the risk of abuse and failure increases with the level of participation, so does the overall potential for achieving a positive effect on legitimacy. I concluded that to reduce the democratic deficit in economic development cases, a wide power of participation must be granted to the land owners and their communities. This is needed, in particular, to restore balance in the relationship between owners and others directly connected with the land, the planning authorities, and the commercial actors interested in development for profit. The question that is as of yet unresolved is how to \isr{organise} such participation in a way that avoids obvious pitfalls, such as administrative inefficiency and tyranny by majorities or elites that assume control of the local agenda.

In Chapter \ref{chap:6}, I will shed light on this question by considering the Norwegian institution of land consolidation, which has a very long tradition behind it. It is a flexible \isr{framework} which includes, among other things, a template for establishing institutions that can function as a LAD. I will focus on how land consolidation functions in cases of economic development that would otherwise likely be pursued by eminent domain. The case study is based on considering hydropower development, but I will also discuss planning law and development more generally, as the Norwegian government is now considering making consolidation, traditionally a rural institution, a primary mechanism for land development even in urban areas.

Before I delve into this, I will present an overview of Norwegian hydropower and the role of waterfalls  as private property. This will serve as an introduction to the second part of this thesis, to which I now turn. }