\chapter{Compulsory Participation in Hydropower Development}\label{chap:6}

\section{Introduction}\label{sec:intro6}

In this Chapter, I will consider an {\it alternative} to expropriation in the context of economic development. This is also where I ended my theoretical discussion in Part I, by presenting and analysing the land assembly proposals put forth by Heller and Hills in US. Here, I return to this point in the context of my case study, by exploring the Norwegian institution of {\it land consolidation}. 

In recent years, this institution has been used extensively to facilitate hydropower projects. So far, however, it is used almost exclusively in situations when owners themselves organize such projects. In these situations, expropriation orders are rarely sought and rarely authorized, even if some owners object to the plans. Instead, various consolidation measures are used, including the practically important ``use directives'', which can be used to set up organizational frameworks for compulsory implementation of a development plan, possibly against the owners' own wishes, but in a way that presupposes their continued involvement. Essentially, a use directive can be used to compel owners to cooperate and participate in a decision-making process that has economic development as an overarching, binding, aim.

In addition to the growing importance of consolidation in the context of hydropower development, there have been recent legislative developments in Norway that have caused the consolidation alternative to gain importance also in relation to other forms of development. This includes urban and non-agrarian development projects, something that represents a departure from the tradition of land consolidation as a purely agricultural institution. Some argue that novel uses of land consolidation leave the owners in a precarious position, and that it weakens private property rights.\footnote{See, e.g., \cite{stenseth07}.} In this Chapter, I explore the opposite hypothesis, namely that the use of consolidation for economic development can strengthen property as an institution, when it replaces traditional expropriation proceedings. In particular, I argue that the Norwegian system of land consolidation can be used to address the democratic deficit of economic development takings in an elegant way.  

However, I note how this vision presupposes that the land consolidation process continues to function as a service to owners and local communities, as a means for helping them, possibly by nudging them, to implement development projects in accordance with public interests. Land consolidation must not be turned into a tool for powerful commercial interests that seek cheap and effective access to property rights held by weaker parties. Here, it seems that the current system in Norway is likely to be put under strain, if the consolidation alternative to expropriation gains further practical relevance. In this regard, I believe it is important to remain vigilant and to consider what safeguards should be put in place to protect the integrity of the traditional process, as new and powerful actors enter the consolidation scene.

The structure of the Chapter is as follows. I begin in Section \ref{sec:lce}, by presenting the basic idea of using land consolidation as an alternative to expropriation. I discuss broad notions of land consolidation in general, relating them also to the discussing found in Heller and Hills' article. Then I point out some special features of the Norwegian system, which I believe make this system particularly natural to consider in the context of economic development cases. 

Then, in Section \ref{sec:lcc}, I present the Norwegian system of land consolidation in more depth, focusing on the procedural rules and the rules in place to protect property rights against disproportionate interference. I focus particularly on the so-called ``no-loss'' guarantee, which states that a consolidation measure can only be implemented when the benefits will make up for the harms, for all affected properties individually. Hence, in general, land consolidation is quite distinct from expropriation. However, in the context of commercial development, the no-loss criterion can be satisfied, through appropriate forms of benefit sharing. In this way, land consolidation becomes a legitimate alternative to expropriation in such cases, in a manner that also underscores the current owners' right to participate in, and profit from, the undertaking.

In Section \ref{sec:lch}, I discuss land consolidation specifically in the context of hydropower development. I consider several cases in detail, based on court documents and recent work done in a master thesis on consolidation, for which the author carried out interviews with affected owners. Then, in Section \ref{sec:lca}, I offer a assessment and discuss some future challenges. I also note some shortcomings of the current system, while arguing that these do not detract from the great potential of using consolidation to facilitate economic development, particularly in commercial settings. In Section  \ref{sec:conc}, I offer a short summary and a conclusion.

\noo{Essentially,}

\section{Land Consolidation as an Alternative to Expropriation}\label{sec:lce}

The notion of land consolidation is somewhat ambiguous. At its core, it refers to  mechanisms whereby boundaries in real property are redrawn to reduce fragmentation, without affecting the relative value of the different owners' holdings.\footnote{See, e.g., the entry on {\it land consolidation} in \cite{mayhew09}.} However, it is also common to use consolidation to refer to mechanisms for pooling together small parcels of land to create larger units.\footnote{See, e.g., \cite{lerman06}.} There is a tension between these two notions of consolidation, with some claiming that consolidation in the latter sense is sometimes used to surreptitiously bestow benefits on powerful property owners, at the expense of weaker groups.\footcite[237-239]{lipton09}

In light of this, I should stress at the outset that I will use the term land consolidation in a very broad sense in this chapter, much wider than {\it both} of the interpretations mentioned above. Land consolidation, as I use the term, refers to any mechanism by which the state intervenes, at the request of some interested party, to (re)organize property rights in a given local area. Hence, a consolidation measure might as well involve {\it increased} fragmentation of property, if this is deemed a rational form of consolidation of the property {\it values} of the affected area. Importantly, I also use land consolidation to refer to efforts directed at organizing the {\it use} of property, not just redrawing boundaries.

Some might argue that this terminology is strained, but I adopt it for a reason. It is motivated by the fact that in Norway, the institution known as ``jordskifte'', officially translated as land consolidation, has a very broad meaning.\footnote{See, e.g.,\cite{reiten09,rognes03}.} Norway is not unique in this regard. Land consolidation has a similarly broad scope in many jurisdictions of continental Europe, as well as in Japan and in parts of the developing world.\footnote{See \cite{sky07,vitikainen04}.} Moreover, in Heller and Hills' work on land assembly districts, a comparison with land consolidation is presented, based on the same broad notion that I use here.\footcite{heller08} One of my main aims in this chapter is to pick up on this, by offering a more detailed comparison and assessment, specifically anchored in the Norwegian system and its application in the context of hydropower development.

As land consolidation tends to involve interference in property rights, one may ask about the legitimacy of various consolidation measures, held against rules that protect private property owners.\footnote{For an analysis of the Norwegian land consolidation process held against the provisions of the ECHR, I refer to \cite{utgard09}.} Such legitimacy issues have  been raised before the ECtHR on a few occasions, resulting in the Court finding fault with the Austrian system of land consolidation in particular.\footnote{These decisions must be understood in light of the potentially excessive duration of consolidation proceedings under Austrian law, during which restrictions are also imposed on the owners, limiting their opportunities to enjoy their properties while awaiting a final outcome. See, e.g, \cite{erkner87,poiss87}.} Moreover, one may sometimes argue that a land consolidation measure {\it is} a form of expropriation, even if it is not recognized as such by the legislature. In the US, for instance, a land consolidation provision ordering escheat (to Indian tribes) of fractional property interests in Indian reservations was struck down as an uncompensated taking by the Supreme Court.\footnote{See \cite{hodel87}.}

\noo{ Move: On the other hand, if land consolidation is used to facilitate or impose specific uses of property, it can also be used as an {\it alternative} to expropriation, a compulsory measure that can obviate the need for depriving owners of their property rights. I think this latter perspective on land consolidation is particularly interesting, and it is the perspective I adopt in this chapter.}

In relation to the legitimacy issue, the Norwegian system stands out in two important regards. First, the consolidation procedure is managed by judicial bodies, namely the {\it land consolidation courts}.\footnote{See generally \cite{langbach09}. The fact that the land consolidation process is administered by a judicial body appears to be unique to Norway, see \cite[45]{sky01}.} Second, land consolidation is largely seen as a service to owners, not a tool for increased state control and top-down management.\footnote{See generally \cite{sky09}.} In particular, a case before the land consolidation courts is almost always initiated by (some of) the affected owners themselves and the court often acts as a ``problem-solver'', aiming to facilitate dialogue and cooperation among owners.\footnote{See generally \cite{rognes98,rognes03,rognes07}.} Moreover, it is a core principle of law that no land consolidation measure may be undertaken unless the benefits make up for the harms, for all the properties involved.\footnote{See the \cite[3 a)]{lca79}.} This is known as the ``no-loss'' criterion and it is one of the key principles of land consolidation in Norway.\footnote{See generally \cite{rygg98}.} The combination of a judicial procedure that emphasises owner-participation and a no-loss criterion means that, arguably, land consolidation in Norway {\it strengthens} property as an institution.

In particular, I believe that land consolidation can serve as an effective countermeasure against two of the most widely discussed challenges to any property regime. First, consolidation can serve to protect an egalitarian distribution of property rights against the deleterious effect of inefficiency and underdevelopment that might otherwise arise from fragmentation. Importantly, it can do so without disturbing the underlying property structure and without necessarily resulting in disproportionate benefits or harms to owners and other private parties. In particular, land consolidation can ensure commercial development without pooling together property rights and without handing property over to powerful market actors. Second, land consolidation can serve to ensure sustainable and rational management of jointly owned land, without necessarily forcing an enclosure process (enclosure {\it can} be the result of land consolidation, but it is only one of many measures in the consolidation toolbox).

In short, land consolidation can be used to address both commons\footcite{hardin68} and anti-commons\footcite{heller98} problems, in a way that protects, and possibly enhances, desirable social functions of property, through a judicial system of participatory/adversarial decision-making. Hence, land consolidation in Norway is based on a conceptual premise that -- potentially -- offers protection to owners and their properties, by recognizing them as members of a community that are mutually dependent on each other. In this way, the form of property protection offered in the context of land consolidation is distinct from the protection offered in the context of expropriation. This in itself is interesting, particularly from the perspective of property's social functions, as discussed in Part I.

The vision of land consolidation at work here is one that sees it as a means for setting up a mini-democracy on demand, to organize decision-making processes in a way that grants those most intimately affected -- the owners and other rights holders -- a say that is proportional to their stake in the matter at hand. As discussed in Part I, this is something that it often seems hard or impossible to achieve through the standard administrative/political route, particularly when powerful commercial actors engage in extensive lobbying and are allowed to assume the position of primary stakeholders in projects involving the property of others.

Importantly, since land consolidation can be used to facilitate or impose specific uses of property, it can also be an {\it effective} alternative to expropriation, a compulsory measure that can obviate the need for depriving owners of their property rights. Clearly, consolidation as an alternative to expropriation is particularly natural for economic development projects. Importantly, the no-loss criterion will generally be possible to fulfil in these cases, through benefit sharing. Moreover, it then falls on the land consolidation court to {\it ensure} that benefit sharing results, so that consolidation measures may be applied in accordance with the law.

Interestingly, it is usually also required that the benefits resulting from consolidation must be distributed  among the affected properties in accordance with their relative value prior to the consolidation measure\footnote{This principle is not as strictly encoded as the no-loss criterion, but is formulated as an ``ought''-rule. See the \cite[31|41]{lca79}. For the special case of consolidation to implement a zoning plan, the rule is absolute, see \cite[3 b)]{lca79}.} So the principle of benefit sharing at work is not compensatory at all, but rather one that sees the owners as active participants in the development, possibly against their own will. I find this highly interesting, particularly from the point of view of the human flourishing conceptions of property that I discussed in Part I.

More pragmatically, the emphasis on benefit sharing in land consolidation reveals a concrete potential advantage of land consolidation over expropriation, from the owners' point of view. In particular, while benefit sharing is required by consolidation law, it is much less commonly achieved through compensation in the context of expropriation.\footnote{This is largely due to the so-called {\it no scheme} principle, which states that compensation to the owner following expropriation should not reflect changes in value that are due to the expropriation scheme. I am not aware of a single jurisdiction that does not include a variant of this principle. For a detailed investigation into the question of whether or not it stands in the way of benefit sharing in economic development cases, I point to \cite{dyrkolbotn15}.} This means that the use of land consolidation in place of expropriation has considerable potential also in relation to the problem of the ``uncompensated increment'' in economic development takings, as discussed in Part I.

On the other hand, it also means that commercially motivated developers may have an {\it incentive} to favour expropriation over consolidation. This raises the question of whether or not owners can use land consolidation as a {\it defence} against expropriation. Can an expropriation threat be countered by bringing a case before the consolidation court, whereby the court is asked to organize the development in question on behalf of the owners themselves?

As long as an expropriation order has not been granted, such a move would be possible in theory. However, whether the land consolidation court would take on such a case is unclear. It would probably depend on the nature of the development in question. In addition, it seems unclear how the expropriation authorities would react; an ongoing consolidation case might not detract them from issuing an expropriation order to an external developer. After such an order has been granted, the law as it stands leaves no room for a consolidation defence. Quite the contrary, in these situations, the land consolidation courts may themselves be called on to implement the expropriation order (by awarding compensation) as part of a consolidation procedure.\footnote{See \cite[6]{lca79}.}

In the future, if the consolidation alternative to expropriation is to develop successfully, it seems that the owners' right to request consolidation in place of expropriation needs to be strengthened. So far, the Norwegian system is moving along a trajectory where land consolidation as an alternative to expropriation is primarily seen as a service to developers and the public, not as a means for empowering owners. It is noteworthy, in particular, that following a change in the law that takes effect in 2016, the {\it developers} will be granted a new right, namely that of bringing a case before the land consolidation courts, to seek help in implementing their project. Developers might well be motivated to do so, because of the potential for reduced administrative costs, a more effective and flexible procedure, and a chance of limiting compensation claims by imposing (cheaper) compensatory consolidation measures (e.g., by providing owners with replacement property).

One question that arises in relation to this is the following: will land consolidation remain a service to owners, or will it become a service to developers who seek cheap access to property owned by others? This question is about to become pressing in Norway, as the scope of land consolidation continually broadens, making it intersect with expropriation law to a greater extent than before. In addition to the new rules granting developers a formal standing in certain consolidation disputes, this development is also strongly felt in the move to apply land consolidation also in the context of urban development, outside the traditional scope of agricultural pursuits.\footnote{See generally \cite{stenseth07}.}

\noo{However, the issue of benefit sharing is bound to come up, particularly in the context of commercial development. In this regard, the risk for developers is that they will be compelled to share the benefits with the owners. However, in order for this to happen, the Land Consolidation Court must actively take steps to make it happen, by recognizing the owners' right to benefit sharing. Moreover, while benefit sharing is a fundamental principle for land consolidation among owners, it remains to be seen if this way of thinking will be preserved when new and powerful external actors enter the scene.}

The idea that consolidation can serve as an alternative to expropriation also raises practical questions concerning how exactly it would work. Here there is already some interesting empirical data available, arising mainly from situations when owners themselves undertake economic development, but make use of consolidation measures instead of seeking permission to expropriate property from uncooperative neighbours.

Interestingly, in the context of hydropower development, this use of land consolidation has become very important in recent years. In 2009, the Court Administration estimated that land consolidation had helped realise small-scale hydropower projects with a total annual energy output of about 2 TWh/year. Moreover, in a recent Supreme Court case, the importance of land consolidation was stressed specifically, as a justification for requiring a commercial taker to pay additional compensation to the owners of waterfalls that were to be used for hydropower generation.\footnote{See \cite{klovtveit11}.}

In the next section, I give further details on the Norwegian system, focusing on the ``use directive'' that have proven particularly important in relation to hydropower. Then, in Section \ref{sec:lch}, I consider land consolidation to facilitate small-scale hydropower specifically. I approach this as a test case for the more general proposition that land consolidation can be a legitimacy-enhancing alternative to expropriation for  economic development.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\noo{
In particular, a case can now be brought before the land consolidation court by an external developer who would otherwise need to expropriate land to implement a project. However, this change in the law also contributes to a shift away from seeing land consolidation as a service to owners, towards seeing it also a service to developers who seek control of property they do not own. This shift could in turn change the dynamics of land consolidation in a way that makes it less distinct from expropriation.

Even though this definition is broad, I note that a clear distinction can be drawn between land consolidation and national or regional land {\it policies}, which do not target specific properties. The distinction between land consolidation and land-use planning can be harder to draw, but looking to the theoretical starting points of these two kinds of interventions, suffice to establish sketch.

While state planning is an expression of the state's right to regulate the use of land, a land consolidation measure is a {\it service} provided by the state, to facilitate property uses and structures that are deemed desirable from the point of view of the properties as productive units under private ownership.

The distinction between consolidation and measures of land reform may sometimes also be difficult to draw, particularly with my wide notion of consolidation. However, while land reforms tend to arise from centrally directed measures that apply generally within a jurisdiction and come about as the result of a special political initiative, consolidation usually denotes a more flexible framework where local communities are restructured in a way that aims to bring benefits to all owners and rights holders within that community. As such, consolidation rules may alleviate the need for new land reforms, and they may come to represent a ``bottom up" approach to the restructuring of real property.\footnote{The potential for this has been noted even for the traditional understanding of consolidation, as a reduction in the level of property fragmentation. See, e.g., \cite{oldenburg90}. For a different perspective, arguing that land consolidation is generally not sufficient to achieve the noble ends of land reform, see \cite{lipton74}. For a more recent, comprehensive, assessment of the relationship between land reform and consolidation (in the narrow sense), I refer to \cite[237-244]{lipton09}.}

In the following, I adopt this normative stance on the {\it purpose} of consolidation. Hence, I use land consolidation to refer to a regulated process of land reorganization that come about as a result of a concrete, often local initiative, has a limited geographical scope, relies on the involvement of the local population, and seeks to promote the best interests of all the affected land users. I remark that while Norway has a particularly broad approach, land consolidation more or less in line with my understanding here serves an important function in many jurisdictions.\footnote{For a survey of contemporary land consolidation rules in Europe, reflecting also the need for a wide understanding of the term, I point to \cite{vitikainen2004}.}

One attractive feature of consolidation is that it provides a flexible, dynamic, framework that allows for gradual adaptation of ownership structures, so that they better suit prevailing economic and social conditions. Moreover, land consolidation can become significant in relation to concrete development projects, particularly when such projects necessitate cooperation among several owners. This, in particular, is the use of consolidation that I aim to shed particular light on in this chapter, by giving a case-study of Norwegian law.

Mechanisms for facilitating and organizing concrete development schemes are now integrated into the law relating to consolidation in Norway. These rules do not form part of the historical core of consolidation rules in Norway, however, the focus on land consolidation for development is of a more recent date. It is reflected in a number of new provisions, most recently in the \cite{lca13} which will take effect on 1 January 2016.\footnote{Act no 97 of 10 June 2013 relating to the determination and change of structures of ownership- and rights to real property etc.}

When land consolidation is used as a means to organize development projects it also becomes natural to view it as an alternative to expropriation, especially in cases when development has commercial potential and is meant to be carried out by companies operating for profit. Moreover, the controversy that often surrounds such cases further suggests that it should be explored to what extent processes of consolidation can replace expropriation as an implementation mechanism for development of this kind. As we will see, the principle of local participation and benefit sharing is more firmly entrenched in the rules and procedures that govern the consolidation process than in the processes that govern the use of expropriation. 

The contrast between expropriation and consolidation is particularly clear in Norwegian law, where a ``no-loss" principle is enforced with regards to the latter, protecting all affected owners and rights holders. It states that the consolidation process should not leave any owner or rights holder worse off after consolidation. The aim of consolidation is to bestow a benefit on \emph{all} interested parties.\footnote{See \cite[3 a)]{lca79} and \cite[3-18]{lca13} (takes effect in 2016). For a paper discussing the rule in more detail we point to \cite{rygg98}. Rygg is also critical of what he sees as a development away from a strict interpretation of the no-loss rule.} For instance, if ownership is highly fragmented, consolidation mechanisms may be used to exchange property between owners or to introduce joint ownership, but due to the no loss rule it will not be possible to use consolidation in order to deprive some owners of their property to the benefit of others.

In the following, I map the differences between consolidation and expropriation in Norwegian law, starting with an overview of the land consolidation rules, focusing on the development towards giving these rules greater application in connection with concrete development schemes. I then study some cases of locally controlled hydro-power development where land consolidation was used as a means to organize projects involving many different owners and rights holders. We argue that these cases illustrate how the consolidation rules currently in place are well suited to meet local demands for participation and benefit sharing, more so than the existing framework regulating expropriation.

The structure of the remaining part of the chapter is as follows. In Section \ref{sec:lcc} I briefly present the basic rules regarding land consolidation in Norwegian law, including a presentation of the special consolidation courts used to administer the process. Then I go on to consider in more depth the rules relating to so-called \emph{use directives}, permitting the court to actively pursue development projects on behalf of, and in cooperation, with local owners. Use directives represent a form of compulsory cooperation which I believe deserves further attention in the context of land development, especially as an alternative to expropriation. I follow up with a case-study of hydropower in Section \ref{sec:lch} and provide an assessment in Section \ref{seclca}. \noo{nd in Section \ref{sec:5} I contrast the use of directives with more commonly seen approaches to pooling of resources and commercial land development. In Section \ref{sec:6} I offer a conclusion.
}}

\section{Land Consolidation in Norway}\label{sec:lcc}

Rules regarding land consolidation have a long history in Norwegian law. The first consolidation rules were included already in King Magnus Lagabøte's \emph{landslov} (law of the land) from 1274, the first piece of written legislation known to have been introduced at the national level in Norway.\footnote{See Chapter 4, Section 2 in \cite{nou02}.} The earliest rules targeted jointly held rights in farming land, giving owners and rights holders on that land an opportunity to demand apportionment that would give them exclusive rights on a parcel of land corresponding to their share of the joint rights.\footnote{The share in joint rights belonging to each individual farm was historically determined based on the amount of rent (``skyld") that each farmer paid to the land owner (a figure that was also used to determine the level of taxation). However, following the union with Denmark and especially after the advent of enlightened absolutism, tenant farmers in Norway increasingly bought their land from their land owners.  Indeed, tenant farming became relatively uncommon in Norway after the 18th Century, but the notion of ``skyld'' was kept as a measure of the share each farm had in the now jointly owned larger estates. The notion is still important, for instance in apportionment proceedings, as discussed in \cite{ravna09a}.}

Many of the rules still in place were developed during the 19th century. At this time, the main use of land consolidation was to divide jointly owned and fragmented property rights into parcels in order to facilitate higher intensity farming and better resource management. \noo{, but the relative importance of such measures increased greatly since it was seen as a necessary adjustment in an age when industrialization introduced a range of new and more efficient farming techniques. In particular, as changes in farming methods resulted in an increased need for capital in agriculture, full ownership came to be regarded as more favourable since it meant that better security could be offered to financial institutions.\footnote{References needed.} } However, it was noted that individuation of property rights was not necessarily required. Instead, collective-action mechanisms could be used, to facilitate a more dynamic approach that would not fundamentally disturb the established property structure.\footnote{This idea was behind a range of provisions introduced during the 19th century, not all pertaining to land consolidation. For instance, a special management structure was set up to govern forestry on common land, to avoid overexploitation and ensure rational management, without necessitating enclosure. See generally \cite{stenseth10a}.}

The rules regarding use directives must be understood in this context. They were introduced to facilitate a legal framework for implementing or altering specific use patterns without altering the underlying structure of ownership. Nevertheless, the rules would facilitate a considerable pooling of resources and decision-making power, to set the stage for more intensive and coordinated forms of use.

The initial objective was to enable rural communities to adapt to changing economic conditions without fundamentally altering them or leading to displacement or depopulation. Moreover, the scope of use directives were typically limited to the regulation and reorganization of already established forms of joint use. Hence, use directives were not commonly used to facilitate completely new kinds of development.

Over the last few decades, this has changed. Today, use directives are increasingly applied also to organize development projects that are not agricultural in the traditional sense, even for properties that have no prior connection. Moreover, many additional mechanisms of land consolidation have been introduced, all aiming in various ways to ensure better organization of land use and ownership. It is helpful to recognize three main categories of consolidation tools, as summarized in the following table:

\begin{itemize}
\item \emph{Apportionment of land}: Rules that empower the court to dissolve systems of joint ownership by apportioning to each estate a parcel corresponding to its share, or by reallocating property through exchange of land.\footnote{This is the traditional form of land consolidation in Norway and the main legislative basis for it is provided in the \cite[2 a)-b)]{lca79}.}
\item \emph{Delimitation of boundaries:} Rules that empower the court to determine, mark and describe boundaries between properties and the content and extent of different rights of use attached to the land.\footnote{The main legislative basis for this form of consolidation is found in the \cite[88]{lca79}.}
\item \emph{Directives for use}: Rules that empower the court to prescribe rules for the use of jointly held land, and to organize such use, including setting up organizational units for carrying out specific development projects.\footnote{These rules are found in the \cite[2 c)|34-35]{lca79}.}
\end{itemize}

In all cases, the consolidation court can only employ these tools when they are called on to do so by someone with a recognized legal interest.\footnote{See \cite[5]{lca79}.} This includes the owners, and while other actors have been allowed to enter the consolidation scene in recent years, most cases are still brought before the courts by owners themselves.

An additional requirement on the consolidation court is that it can only implement consolidation measures in so far as they are needed to alleviate problems and difficulties.\footnote{See the \cite[1]{lca79}.} To determine whether or not this requirement has been met, the court will look to the prevailing economic and social situation, as well as predictions for the future.\footnote{See generally \cite{reiten09}.} In this regard, the court is also influenced by what it regards as the prevailing public interests in property use. Today, the role of the perceived public interest is growing in importance, and recent reforms have sought to strengthen it.\footnote{See for instance \cite{prop12} (proposal from the Ministry of Agriculture to the parliament regarding the Consolidation Act 2013).}

But the contextual nature of land consolidation has always been clear. Indeed, the basic building blocks of the current system can be traced back to the influence of technological advances in farming and the modernization processes that Norwegian society underwent in the 19th century. The law responded to these changes and consolidation became an increasingly important instrument for change and development in this period. It was also at this time that it was decided to establish a tribunal system for administering the process, first in the Land Consolidation Act from 1857, which was revised and developed further in 1882 and 1950.\footnote{An overview of the history of consolidation law is given in Chapter 3 of \cite{prop12}.} The procedural rules closely mimics those that pertain to the regular civil courts. This ensures that consolidation tools are only put to use if a court orders it, and only after a public hearing where all involved parties are given an opportunity to present their case, give supporting evidence, and to contradict each others' testimony. For a more detailed description of the consolidation court, I refer the reader to Section \ref{subsec:lcp} below.

The current system for land consolidation is based on the \cite{lca79}, which will be replaced in 2016 when the new \cite{lca13} takes effect.\footnote{Act no 97 of 10 June 2013 relating to the determination and change of structures of ownership- and rights to real property etc.} The new act does not introduce any dramatic changes to the law, but it further widens the scope of consolidation, particularly with regards to directives for joint use. The new act also contains the following explicit description of the purpose of land consolidation, which provides a interesting bird's eye view on the legislature's perspective:

\begin{quote}
Section 1-1 The purpose of the Act

The purpose of the Act is to facilitate efficient and rational use of real property in the best interests of the owners, the rights holders and society. This objective will be pursued by the land consolidation courts which will implement remedies for impractical structures concerning ownership and use of property, ascertain and determine property boundaries, as well as decide appraisal disputes and other cases pursuant to this and other acts.

The Act also seeks to facilitate fair, responsible, quick and effective processing of cases in independent and impartial public courts that will operate in such a way as to enhance confidence in the consolidation process.\footcite[1]{lca13}
\end{quote}

This statement of purpose highlights how the new act incorporates and extends the trend towards giving the consolidation process wider scope. I also note how it reiterates and emphasises that the process is to be tribunal in nature. More generally, the quote illustrates that land consolidation is likely to become more important in the future, increasingly also outside the traditional agricultural setting within which this body of law has hitherto developed.\footnote{For instance, following a change in the \cite{lca79} in 2006, land consolidation may now also be called on in order to manage restructuring of ownership in urban areas, in connection with specific development schemes. This rule has been extended further in the new Act, and it will be interesting to see how the division of labour will be in the future, between planning authorities, regular courts and the land consolidation courts.}

It is now explicitly stated that the purpose of land consolidation is to make conditions of property use more favourable for all the affected owners and rights holders. Hence, the new act underscores how consolidation represents a form of interference that is fundamentally different from expropriation. As before, the consolidation court is not empowered to take action unless it is called on to do so by one of the stakeholders in the property.\footnote{See \cite[1-5]{lca13}.} However, according to the new Act, a developer who has obtained permission to expropriate is to be counted as a stakeholder in that land for the purposes of consolidation.\footnote{Previously, a developer was only regarded as a stakeholder in consolidation in some cases of public projects, c.f., \cite[5|88|88 a)]{lca79}.} This further reflects how it is becoming increasingly natural to see land consolidation as an alternative to expropriation. It also flags how the relationship between expropriation and consolidation is now becoming an important topic in Norwegian land law.

In 2005, the Ministry of Agriculture made some comments in this regard, in connection with a revision of the \cite{lca79} that gave consolidation greater applicability in urban areas and with respect to implementing public plans.\footnote{See, in particular, \cite[2 h-i)]{lca79}.} Some members of the preparatory committee had raised the concern that giving consolidation extended scope in this way would be problematic since it would encroach on expropriation law and effectively render consolidation a form of expropriation. The Ministry disagreed, commenting as follows.

\begin{quote}
The Ministry would like to point out that one of the main preconditions for consolidation is that a net profit is created for the land in question. This profit is then divided among the parties in an orderly fashion. Individually, the law also guarantees that no one suffers a loss, see s 3 a). [...] In the Ministry's opinion, expropriation takes place on a different factual and legal basis. In cases of expropriation the public makes decisions that deprives the parties of economic value. The purpose then becomes to compensated them in accordance with s 105 of the Constitution, not to increase the value of their land or the annual income they may derive from it.\footnote{See Chapter 3.3 of \cite{otprp78} (report to parliament from the Ministry regarding changes in the \cite{lca79}.}
\end{quote}

When preparing the new Act, the Ministry of Agriculture reiterated this position, but they did not reflect further on the question of the exact relationship between consolidation and expropriation. They observed, however, that changing the law so that expropriating parties could appear in consolidation cases was \emph{reasonable}, since it would then be left up to the developer whether to make use of their permission to expropriate or to rely on consolidation.\footnote{See \cite[84]{prop12}.}

The choice made by the expropriating party in this regard will be of great importance to the affected owners and rights holders. In particular, as the Ministry themselves makes clear, it is an absolute precondition for the implementation of a rights-altering consolidation measure that it serves to make the structure of ownership and use more favourable. This requirement, moreover, refers explicitly to the \emph{area within which consolidation takes place}.\footnote{See the \cite[3-3]{lca13}.} No similar rule is in place to protect the affected local community following expropriation. Moreover, the practices that have developed for dealing with consolidation cases are centred on the interests of the local owners and their communities to an extent that is quite different from any procedure that is currently in place to facilitate development by use of expropriation.

For instance, the rule regarding expropriation that corresponds most closely to the no-loss rule  
requires merely that the benefit to private and public interest exceeds the disadvantages \emph{overall}, not locally and certainly not individually.\footnote{See the \cite[2]{ea59}.} At the same time, consolidation rules do not place any restrictions on the kinds of development that can be carried out. The consolidation rules pertain instead to \emph{how} it should be organized. 

The ``what''-question may be left entirely to the planning processes administered by the planning authorities (the municipalities in Norway). Moreover, the consolidation courts must always base their decisions on existing public regulations of the property use.\footnote{In the \cite[3-17]{lca13} it is explicitly stated that the consolidation court cannot prescribe solutions that are not in keeping with such regulation. However, it is also made clear that the consolidation court itself can apply for necessary planning permissions on behalf of the owners and the land in question.}

Hence, if the public interest suggests a particular form of land use, the fact that a planning decision detailing development of such use is implemented through consolidation does not entitle the court to review the plans themselves, going against the public interest. But it does introduce an obligation, emerging at the time of implementation, to turn specifically to the interests of original owners and rights holders and to look for solutions that minimize the burden and maximizes the benefit for all the properties involved.

The rules that give the consolidation court authority to give directives of use are particularly relevant in this regard. I return to these rules below, after I have presented some further details on the consolidation process itself. The procedural guarantees resulting from the fact that this process is tribunal in nature are in themselves important to consider when looking at consolidation as an alternative to expropriation.

\subsection{The Consolidation Process}\label{subsec:lcp}

A consolidation case is usually initiated by an owner or a permanent rights holder.\footnote{See s 5, para 1 of the \cite{lca79}.} The request for consolidation measures is to be directed at the relevant district consolidation court, one of the 34 district courts for land consolidation that have been set up by the King in accordance with s 7 of the \cite{lca79}. The request is meant to include further details about the affected properties, the owners and rights holder involved, as well as the specific issues that consolidation should address. But the requirements in this regard are not usually interpreted very strictly and the consolidation court will often take it upon itself to further clarify what the case should encompass, more so than in regular civil disputes.\footcite[39]{langbach09} However, the court may still reject the consolidation request if it finds that it suffers from formal shortcomings, pursuant to the same rules as those that apply to civil disputes.\footnote{See s 12, para 2 of the \cite{lca79}, which refers to s 16-5 of the \cite{cda05}.}

If the court decides that the request is well-formed and that it includes sufficient detail to permit material consideration, they go on to prepare public hearings, following the rules set out in Chapter 3 of the \cite{lca79}. These rules mirror those that are in place for civil hearings in general, including the duty to inform affected parties, the parties' right to present their claims, as well as their duty and right to give testimony and provide evidence supporting it.\footnote{See the \cite[13|15|17 a|18]{lca79}.} As in civil cases, the decision is usually only reached after at least one oral hearing where the parties may present and comment on the evidence and the issues raised by the case. 

But unlike in civil cases, the main hearing typically takes place on the disputed land itself and often revolves around practical rather than legal issues. Moreover, a consolidation case will usually not take the form of a two-party adversarial process, but rather as a multi-party discussion where the court interacts with a large number of interested parties who may have a range of common as well as conflicting interests. The typical case involves 5-10 people, but in some cases there can be hundreds of parties involved.\footcite[39]{langbach09} In addition, it is quite common that the parties are not represented by legal council.\footcite[109-111]{rognes00} And even if they are, they are typically expected to take an active part in the proceedings themselves.\footnote{See generally \cite{rognes00}.}

The request for consolidation will be the court's point of departure when assessing the case, but the court is not bound by the claims put forth in it, or by the claims put forth by the other parties. This again marks a differences with most cases of civil dispute. With a few exceptions explicitly listed in statute, the consolidation court may decided to use any measure that it deems suitable to ensure a favourable structure of rights and ownership for the future. However, there is some restriction placed on the court in that the measures taken must be regarded as \emph{necessary} in light of considerations based on the original request.\footnote{See s 26 and 29 of the \cite{lca79}.} So the court should remain focused on the issues raised by the parties, but is free to address these issues using the tools they deem most suited for the job. The consolidation court, in particular, is meant to be a general ``problem solver", much more so than an  ordinary civil court.\footnote{See generally, \cite{rognes07}.}

When a decision is reached, the parties are notified and the decision is presented and argued for in keeping with the rules of the \cite{cda05}.\footnote{See the 
\cite[7|22]{lca79}.} The appropriate format for the decision depends on its content. A regular civil ruling is the form used for decisions that only involve ascertaining the boundaries between properties, while a special ``consolidation decision" is used to implement apportionment and directives of use. The difference in form affects the appeals procedure; while civil rulings are dealt with by the regular courts of appeal, the consolidation decisions can only be appealed to one of 4 designated consolidation courts of appeal.\footnote{See  the \cite[61]{lca79}.}

The procedural rules remain largely the same in the consolidation court of appeal and there is a new assessment of all aspects of the case.\footnote{See s 69 of the \cite{lca79}.} The decision of the consolidation court of appeal can only be appealed on the grounds that it is based on an incorrect understanding of the law, or that procedural mistakes were made. In this case, the ordinary appeal courts have authority, with the Supreme Court being the last instance of possible appeal.\footnote{See s 71 of the \cite{lca79}.}

In general, consolidation cases are different from other civil cases mainly in that they have fundamentally different scope. A consolidation case is not primarily concerned with deciding the merits of individual claims, but focuses on introducing structures of ownership and rights that will prove favourable in general. In this respect, the process has an administrative character. However, the fact that it is organized similarly to a civil dispute means that the affected parties play a more prominent role in the decision-making process than they do when decisions are made by regular administrative bodies.

Given the special context of arbitration, it is not surprising that the judges appointed to the consolidation courts are required to have a special skill set, different from that of regular civil law judges. In fact, consolidation judges are required to have successfully completed a special master level degree in consolidation, which is not a law degree, but a separate form of education.\footnote{See s 7, para 5 of the \cite{lca79}. The degree in question is currently offered only at the Norwegian College of Life Sciences and Agriculture.} 

The consolidation court also relies on the participation of lay people who sit alongside the specialist judge.\footnote{See s 8 of the \cite{lca79}.} These lay judges are appointed by the specialist judge from a committee of laymen that are elected by the local municipalities.\footnote{See s 8 of the \cite{lca79} (the appointment itself is regulated in the \cite[64]{ca15}).} To the extent possible, the appointed laymen should have special knowledge of the issues raised by the case. But they are drawn from the general population.\footnote{See s 9, para 5 of the \cite{lca79}.}

Summing up, the consolidation process has both administrative and adversarial characteristics. While the content and scope of the court's decision will often have an administrative flavour and is not primarily directed at settling any specific dispute, the process is judicial. Hence everyone is entitled, and to some extent even \emph{obliged}, to have his voice heard and to partake in the process. Moreover, while the process is guided and overseen by the court, it is fundamentally based on considerations arising from the interests of the parties. 

However, this interest is always interpreted in light of prevailing notions of what counts as favourable and rational property use. Importantly, in relation to this latter assessment, the court will look beyond the interests of the individual owners. The court will pose the question with regards to the use of the land as such, drawing on its understanding of the relevant economic, social and political conditions.\footnote{See generally \cite{reiten09,sky09}.} At the same time, the decisions are made on the basis of information that presented and discussed in public hearings. Moreover, the affected parties will take part in discussions that may also address more overarching concerns about the form of land use that should be regarded as favourable for the area in question.\footnote{See also \cite{rognes07}.}

To flag the dual nature of the consolidation process it is tempting to designate it as a process of judicially structured \emph{deliberation}. The final decision-making authority is granted to the court, but the court is required to act on behalf of the rights holders, on the basis of their wishes, bu always in the best interest of their properties and their community. This form of decision-making based on multi-party deliberation is interesting in its own right, as it provides a template for management of land that caters to the idea of public oversight and control as well as to the idea of local participation and self-governance. At any rate, it is a form of land management that seems especially suitable as a means to implement concrete projects undertaken in the public interest, particularly when these would otherwise appear to adversely affect individual land owners and local communities.

Moreover, the notion of land consolidation can serve as an additional conceptual layer between the planning stage and the implementation step of a development plan, a layer of management devoted to translating public interests and private plans into concrete action on private property. This, I believe, is a layer of administration that deserves more attention and more fine-grained tools than those currently offered in systems relying on expropriation. Clearly identifying a consolidation layer in property management for economic development can also make for a cleaner delineation between commercial implementation on the one hand, governed by the market, and public planning on the other, governed by administrative law and political bodies, avoiding unhealthy coalescence between the two.

In the next section, I argue that Norwegian consolidation law already include tools that make it possible to rely on the consolidation courts to provide such a service. The rules that I believe warrant this conclusion are the rules relating to joint use directives, which I now present in more detail.

\subsection{Joint use, joint action and joint investment}\label{sec:3}

Traditionally, use directives would target property rights that were owned jointly or for which some form of shared use had already been established.\footnote{In accordance with s 2 c) of the \cite{lca79} .} However, in the 1979 Act, the power of the courts to issue use directives was extended, so that directives could also be issued when there was no prior connection between the rights and properties in question, provided \emph{special reasons} made this desirable.\footnote{See s 2 c), para 2 of the \cite{lca79}.} Traditional examples include directives for the shared use of a private road which crosses several different properties, or regulation of hunting that takes place across property boundaries.

The joint use rules emerged as an alternative to apportionment of jointly owned property, a more subtle and less invasive measure that could often give rise to the same positive effect as a full division of ownership, but without leading to unwanted fragmentation of control and use of property. Hence, in the now repealed Land Consolidation Act 1950 it was stated that use directives should be the \emph{primary} mechanism of consolidation, such that apportionment could only take place if such directives were deemed insufficient to reach the goal of creating more favourable conditions for the use of the land.\footnote{See s 3 no 3 and 4 of the Land Consolidation Act 1950 and the discussion in \cite[30-37]{nou76}.} In the 1979 Act, the two mechanisms were formally put side by side, but the intention behind this was to ensure greater flexibility of the system, not to reduce the scope of use directives. Quite the contrary, the 1979 Act explicitly intended to promote the increased use of such directives, also in conjunction with other measures.\footnote{See the discussion in \cite[35-37]{nou76} and \cite[47-48]{otprp56}.}

Since the Act was introduced, there has been a gradual increase in the willingness of the court to rely on use directives to facilitate \emph{new development} on the land, not just as a means to regulate an existing activity. In parallel with this development, the consolidation court has gradually come to take on cases that pertain to organization of land use that was previously thought to lie outside its area of competence. 

The \cite{lca79} lists a range of different concrete circumstances in which such directives can be applied.\footnote{See s 35 of the \cite{lca79}.} The list is not understood to be exhaustive however, and as the notion of agriculture has broadened, so has the scope of use directives in land consolidation.\footcite[103]{otprp57}

In the \cite{lca13}, the list is replaced by an explicit general rule which makes it clear that the  consolidation courts have the authority to give directives whenever they regard this to be favourable to the properties involved.\footnote{See s 3-8 of the \cite{lca13}.} In addition to this, the new Act also introduces a general rule which gives the court authority to \emph{set up} systems of joint ownership when a joint use directive is deemed insufficient.\footnote{See s 3-5 of the \cite{lca13}.} Hence, apportionment and pooling of property is now on equal footing, although a priority rule is introduced for the latter; pooling will only be considered if directives of joint use are regarded as an insufficient means to ensure more favourable conditions. Moreover, the new Act maintains the principle that directives regarding the joint use of land for which there are no existing joint rights can only be given if there are special reasons.

This requirement is not intended to be very strict and the Ministry of Agriculture was initially inclined to remove it.\footnote{For a discussion on this see \cite[140-141]{prop12}.} However, it was eventually decided that it should be kept in order to flag that two distinct questions arise in such cases. The court must always consider the question of whether or not joint use is in fact desirable, in addition to the question of how it should be organized.

In addition to giving directives prescribing how joint use is to be organized, the consolidation court may also give rules compelling owners to take joint action to realise potentials inherent in their land. Rules to this effect were novel to the \cite{lca79}, and made contingent on special circumstances.\footnote{The rules are given in the \cite[2 e)|42-44]{lca79}.} Currently, joint action can only be prescribed to property owners, not other parties.\footnote{See the \cite[34 a)]{lca79}.} Following the new \cite{lca13}, however, the consolidation courts will be authorised to prescribe joint action also to right holders. In addition, the list of circumstances that warrant joint action prescriptions will be replaced by a general joint action rule.\footnote{See s 3-9 of the \cite{lca13}.}

When commenting on this change in the law, the Ministry noted that the joint action rules currently in place have been widely used, and that applying them is now one of the core responsibilities of the consolidation courts.\footnote{See \cite[146]{prop12}.} Joint action directives can include prescriptions for joint investments.\footnote{See s 3-9 of the \cite{lca13}.} On the one hand, this means that such directives can be used to facilitate capital-intensive new development, making consolidation a more effective tool to implement economic development. On the other hand, questions arise regarding the extent to which it is legitimate to rely on compulsion in this regard.

The magnitude of the joint actions and investments required to undertake projects can easily become quite burdensome for individual owners, a worry that is particularly likely to arise in case of large-scale commercial development. The \cite{lca79} attempts to resolve this by a rule stating that if joint actions or investments may come to involve``great risk", the court must set up two \emph{distinct} organizational units to undertake it.\footnote{See the \cite[34 b)|42]{lca79}.} First, the rights needed to undertake the scheme will be pooled together and managed by an owners' association, and then, to undertake the scheme itself, a cooperative company structure will be set up on behalf of the owners. 

Hence, the risk is diverted away from the individual owners onto a company controlled by them. This company will be entitled to any potential profit from the scheme, but it will also be required to pay rent to the owners' association on terms established by the parties themselves, with the help of the court.\footnote{See s 34 b) of the \cite{lca79}.} Moreover, the owners are entitled to shares in this company proportional to their share of the relevant rights in the land, as determined by the consolidation court. The owners are not obliged to take part in the undertaking by acquiring such shares, and will benefit from membership in the owners' association regardless of whether or not they  choose to do so. 

This two-tier system seems like it provides just the mechanism that is required to allow owners to engage actively also with large-scale projects that involve external commercial actors. In particular, the owners' association may strike a deal with such a developer, if the owners themselves do not desire to undertake the actual development. However, new conflicts may arise between the owners in this regard, if some desire to undertake development themselves, while others wish to strike a deal with an external developer. Should the owners that wish to undertake their own development be prioritized by the owners' association, and should the land consolidation be empowered to order such a prioritization? This question has arisen with quite some force in several cases concerning hydropower, discussed in the next Section.

\noo{Despite the potential for disagreement among owners, I believe it is a strength of the system that  owners retain decision-making power and a right to benefit, even when complex development schemes are to be implemented. Indeed, the rules currently found in Norwegian consolidation law adds weight to the claim that and consolidation might point to an alternative and possibly fruitful way of implementing development projects in a system which presupposes that development takes place through commercial initiatives on the basis of public  planning and control.}

Leaving aside the issue of how to deal with disagreement among owners, I conclude that the system currently in place already provide tools that allow the consolidation courts to organize large-scale development on behalf of owners, even when it this requires considerable property (re)organization and diversification of risk. Importantly, I note that the consolidation rules also point to a form of implementation that is likely to allow the public to exercise more extensive oversight and control. This, I believe, is a natural consequence of the fact that the system \emph{curbs} the power and influence of commercial forces. From this fact alone, it must be expected that the consolidation procedure is well suited for pursuing also the social, economic and political aims which motivates the underlying planning decisions. Moreover, commercial development through consolidation gives the public a greater say during the implementation stage. After all, the organizational structure and the implementation plans are formulated by courts, and the courts themselves are explicitly obliged to consider also the public and societal interests in the development.

In addition to this, both planning authorities and commercial developers may take up a role as formally recognized parties in the consolidation process. This seems particularly useful in connection with large-scale industrial development, as it might otherwise be hard to implement such projects successfully. In these cases, then, the consolidation system sets up an arena for interaction and deliberation between the three main groups of stakeholders: the public, the local owners and the commercially motivated developers. Such an arena is so far missing at the implementation stage of big development projects, where owners and their communities in particular tend to become completely marginalized, particularly when expropriation is used.

It remains unclear to what extent the Norwegian consolidation rules will actually be used to give property owners a leading voice in large-scale development. The tension between expropriation and consolidation has yet to arise in case law from this angle. However, consolidation is beginning to receive much attention as a ``practical'' alternative to expropriation. Hence, I believe it is only a matter of time before deeper questions of participation rights and benefit sharing will arise.

\noo{Moreover, as I have already mentioned, consolidation has become an important means for organizing owner-led hydropower development. In the following section, I describe this in more depth, and I argue that it provides a blueprint for how the consolidation system should approach large-scale commercial development in general.

On the theoretical side, it is also unclear to what extent original owners may \emph{demand} that the rules are put to use. For instance, may an owner request consolidation to prevent a permission to expropriate from being implemented? It will be interesting to see how the Norwegian legal systems will deal with this and related questions, after the new Act takes effect in 2016.

I conclude this section by addressing a new special rule that has been included in the new act, and which is specifically targeted at the planning authorities, encouraging them to make use of consolidation to achieve  greater fairness in public planning. These rules are contained in Chapter 5 of the new act and they target benefit that arises from planning in cases when the benefit appears to fall disproportionately on some owners. Such cases of ``windfall" benefit due to public plans are often flagged as problematic, and they arise with particular frequency in systems based on commercial implementation. For example, if one parcel of land is designated for housing and some neighbouring land is designated as a playground, it might easily come to be seen as unfair that a considerable financial benefit falls to the owner of the land designated for housing, while the playground owner is left with virtually nothing.

Following a change of the Consolidation Act in 2006 which has been further extended in the new act, the law now makes it possible for the planning authorities to decide that apportionment of the \emph{benefit} arising from the plan may be carried out by the consolidation courts.\footnote{See s 3-30 of the \cite{lca13} and s 12-7 no 13 of the \cite{pb08}.} When doing so, the consolidation court will follow the same procedure as in other cases, and it will allocate the benefit arising from the plan based on an assessment of the development potential of the different parcels. Importantly, the court will consider this question independently, and the decision will not be based on the particular manner in which the plan dictates that development is to be carried out. For instance, if the land used for the playground could just as well have been used for housing, the court may decide that the rights to housing development is to be shared equally between the two properties. On the other hand, if there is some independent reason why the playground property is not suited for housing, the court will reduce this property's share in the housing development correspondingly.

The court can implement solutions such as this more effectively and rationally by applying the other tools that it has available. For instance, if the the owner of he playground is entitled to an equal share in housing, then apportionment can be used to actually provide him with such a share, trading it for a corresponding share in the playground. However, if such material reallocation of development rights does not prove feasible, the new act also opens up for a solution where the benefit sharing is implemented using financial compensation.\footnote{See s 3-32 of the \cite{lca13}.}
}

To sum up, directives of use rules are highly versatile and may be used to organize extensive projects of land development on behalf of original owners. This form of development makes it possible for original owners to maintain their interest in the land, it can prevent the need for expropriation, and it may give the public a greater opportunity to exert influence and control over how their planning decisions are implemented in practice. In the next section, I consider in depth the particular case of hydropower, where the consolidation courts have recently started to make use of a wide arsenal of its tools to ensure that development can be carried out in this way.

\section{Compulsory participation in hydropower development}\label{sec:lch}

In this section, I look at four recent cases in detail, all of which involved directives of use for hydropower development by original owners. The waterfalls dealt with in these cases are all located in the county of \emph{Hordaland}, in south-western Norway. Three of the cases involved small-scale hydro-power which some of the owners wanted to develop themselves, while the fourth was a case when the owners were also considering a development plan which would involve cooperation with an external energy company. The cases are particularly interesting because we have access to data on how the process of consolidation, and the outcome, was perceived by the owners themselves. This material is due to Sæmund Stokstad, who conducted interviews for his master thesis on land consolidation, which was devoted to the study of how consolidation measures is now increasingly being used to facilitate hydropower development.\footcite{stokstad11}.

In the following, I first present each case separately, focusing on those issues that were raised regarding how to organize development, the solutions prescribed by the court, and the subsequent reception among the parties. I then assess this from the point of view of developing a better understanding of compulsory cooperation as an alternative to expropriation. I conclude with some unresolved questions, particularly regarding those situations when the court is called on to resolve concrete questions regarding how the development should be organized. These are the cases when the relationship between consolidation law and other legal frameworks, such as company law, planning law, and water law, becomes pressing, and there are many unresolved questions.

\subsection{\emph{Vika}}

The case was brought before the consolidation court in 2005, by owners who all agreed that hydropower development should be pursued.\footcite{vika05} The owners disagreed on how to organize the owners' association, and on how the shares in this association were to be divided among the different properties involved, 15 in total. The main principle was agreed upon from the start, however, namely that the owners would rent out their waterfall to a separate development company which every owner would have a right (but not a duty) to take part in. 

The parties in \emph{Vika} were highly involved in the consolidation process, and the statutes for the owners' association were based on suggestions made by the owners themselves. The main point of disagreement concerned how the shares in this association should be allotted, a question that was made more difficult by the fact that some owners benefited from old water-mill rights in the river. In the end, the consolidation court held that these rights were tied to the form of use relevant at the time they were established. Hence, the rights were not regarded as having any financial value and could therefore be extinguished without compensation.\footnote{As provided for in the \cite[2|36|38]{lca79}.}

There was also some disagreement about whether the voting rights in the owners' association should be tied to the number of shares belonging to each owner, or if the owners should simply be allotted one vote each, irrespectively of their share in the waterfall. The consolidation court went for the first option. However, the way in which they allotted shares in the owners' association deserves special mention. In particular, the court decided to take into account that some additional water entered the waterfall from smaller rivers where only a sub-group of the owners had waterfall rights. These owners' share in the association was increased accordingly, and this is surprising in light of Norwegian water law, as water rights are otherwise not tied to where the water comes from, but arises solely from the rights one has in the waterfall itself.\footnote{See the \cite[13]{wra00}.}

The statutes of the owners' association also contain a second interesting provision, based on a suggestion made by the owners. This provision states that all rights in the association are to be tied to the underlying agricultural properties and that they can not be partitioned off from these properties to be sold separately. In Norway, such division of agricultural land would in any event require permission from the local municipality.\footnote{See s 12 of the \cite{la95}.} In recent years, however, this protection of farming communities has grown weaker in practice, and it was the view of the owners in \emph{Vika} that a dissociation of water rights from the underlying agricultural properties should be forbidden altogether.

According to Stokstad, a general consensus had developed among the parties whereby the land consolidation procedure was seen as a success. It allowed for an orderly and fair decision-making process regarding the conflicts that had arisen, and it was based on continuous interaction between the owners and the court, where everyone felt they had been given an opportunity to have their voice heard. Initially, tensions among the owners had been high, but the consolidation process had served to alleviate them. Some owners also pointed to the fact that the main hearing had been physically conducted in the local community, in a meeting hall that was familiar to the owners. This also gave them a feeling that they were meant to actively partake in the decision-making process. 

When the interviews were conducted, some 5 years after the case was concluded, the owners also appeared to agree that the association was working as intended and that the climate of cooperation among the owners was good. The hydropower scheme itself had been completed in 2008, yielding an annual production of around 15 GWh/year, providing enough energy for around 700 households. Moreover, following the experience of land consolidation, a culture of deliberation towards consensus had developed among the owners, and great emphasis had subsequently been placed on attempting to find common ground and to reach agreement on important issues. This was reflected, for instance, in the fact that the owner who contributed the land for the power station was given a generous annual fee, in addition to his compensation as a waterfall owner. According to Stokstad, this fee exceeds what he would likely get if this decision had been left to the discretion of the consolidation court. Hence, it reflects a premium that the owners were now willing to pay to ensure agreement and a continued good climate for cooperation.

I agree with Stokstad that the case of \emph{Vika} serves as an example of how land consolidation can empower local communities and enable them to embark on substantial development projects.

\subsection{\emph{Oma}}

The second case I consider is \emph{Oma}, which was brought before the courts in 2006.\footcite{oma06} In this case there were four involved properties. The owners of three of them, $A,B$ and $C$, wanted to develop hydropower, while the fourth, owner $D$, was opposed to the development. Rather than attempting to expropriate the necessary rights from owner $D$, owners $A,B$ and $C$ took the case to the consolidation court. They argued that development would benefit all the properties involved, and also pointed out that an alternative project, which would not make use of owner $D$'s rights, would be less economical. Hence, in their view, the consolidation court should compel $D$ to cooperate in a joint scheme. Owner $D$ protested, arguing that the project would not economically benefit him, and that it would also be to the detriment of his plans to build cottages for holiday dwellers in the same area.

The case of \emph{Oma} differs from that of \emph{Vika} in that the question of whether it was appropriate to use compulsion was more prominent. In particular, this aspect came up already in relation to the question of whether or not hydropower development should be pursued at all. As I discussed in Section \ref{sec:lcc}, the fact that some owners do not desire development does not prevent the consolidation court from putting directives in place to facilitate it. However, the courts often exercise restraint in such cases. In \emph{Oma}, however, the court agreed with the majority  that an owners' association with compulsory membership should be set up.\footnote{In doing so, the court relied on s 2 c) of the \cite{lca79}.}

To justify the use of compulsion against $D$, the court first observed that joint development of hydropower would benefit all the properties in question, including $D$. Then they commented specifically on owner $D$'s plans for building cottage homes, noting first that he was unlikely to be given planning permission, and secondly that hydro-power would not in any event adversely affect such plans in any significant way. Moreover, the court noted that while owner $D$'s rights were relatively minor, they were quite crucial for the profitability of the project, particularly because owner $D$ controlled the best location for the construction of a dam to collect the water used in the scheme. Overall, the court's conclusion was that a joint hydropower scheme would be a better option for everyone than a project that did not include owner $D$'s property.

The question then arose as to how the shares in the owners' association, and the right to rent that would go with it, should be divided among the owners and their land. In regard to this question, the court departed significantly from one of the basic principles of Norwegian hydropower law. The principle states that no right to hydropower can be derived from being in possession of land suitable for the construction of dams or other facilities necessary to exploit the waterfalls.\footnote{The principle is well-established in expropriation law, going back to the Supreme Court decision in \cite{herlandsfossen22}.} The land consolidation court broke with this principle in the case of \emph{Oma}, deciding instead to set the value of the land designated for construction of a dam and a power station at $6 \%$ of the total value of the rights that went into the owners' association. 

The proportion of financial benefit and decision-making power awarded to the unwilling owner $D$ thus increased accordingly, since these rights were all held by him. In fact, his share went from $1.75 \%$ to $7.75 \%$, so the consolidation process itself led to a situation where he would have a far greater incentive for supporting the development. In many ways, the decision in \emph{Oma} was more to the benefit of owner $D$ than any other among the involved parties. If the rights in question had been expropriated, for instance, owner $D$ would be given next to nothing in compensation and would lose his rights forever. Instead, the solution prescribed by the consolidation court gave him a lasting and substantial interest in local hydro-power.

According to Stokstad, interviews with the parties shows how the process and outcome of consolidation in \emph{Oma} served to create a much better climate for further cooperation among the parties. Indeed, when the interviews where conducted, 4 years after the courts' decision, owner $D$ had changed his mind and was now in favour of the development. Moreover, he had also decided that he wanted to take part in the development company. He was not obliged to do so, but his right to take part was ensured by the agreement with the development company, regulated by the statutes of the owners' association.\footnote{The owners' right to take part in the development company is obligatory in some situations, pursuant to \cite[34 b) no 3]{lca79}.}

The owners all reported that the consolidation process had been very successful and that the court had listen to them, allowing everyone to have their voices heard. Moreover, some owners reported that the court had cleverly maintained a ``birds eye view" on the best way to develop the land in question, ensuring both long terms benefit to all involved properties as well as creating an improved climate for cooperation and mutual understanding. The consensus was that making concessions to owner $D$ was appropriate and had been in the interest of all the involved parties. In 2011, the hydropower project was completed and today its output is roughly 3 GWh/year.

I believe \emph{Oma} serves as a good illustration of how consolidation can be an effective instrument for facilitating locally controlled development, also in cases when this requires the use of compulsion against some owners. Interestingly, in this case the successful outcome appears to be partly due to the fact that the consolidation court actively used its discretionary powers when deciding how to organize joint use. This power allowed them to deviate from established rights-based legal doctrine and adopt a more context-dependent approach, pursuing solutions that suited the situation better. Interesting legal questions arise in this regard, particularly regarding the competence that the consolidation court has in such cases, and the extent to which decisions can be made subject to review by the normal courts on the basis that they do not follow established principles and practices of hydropower law.

For instance, one may ask what would have happened if the majority owners in \emph{Oma} had appealed the decision to the regular courts on the basis that $D$ was awarded too many shares in the owners' association. Would this be regarded as a question of the court's interpretation of the law regarding the owners' \emph{rights}, or would it be regarded as a discretionary decision regarding the best way to organize development? If a rights-based perspective was adopted, the decision would almost certainly be overturned. If not, it would seem beyond reproach, as an exercise of the consolidation courts' discretionary power.\footnote{Recall that consolidation decisions can only be appealed to the regular courts on procedural grounds or on the ground that the law has been applied incorrectly.}

A second interesting question that arises is whether or not consolidation can work as well as it did in \emph{Oma} in cases where conflicts run more deeply, or where the parties favoring development are a minority among the owners. The next two cases I consider shed some light on this issue.

\subsection{\emph{Djønno}}

This case was brought before the courts in 2006, by a local owner $A$ who wanted to develop hydropower in a small river crossing his land, the so called \emph{Kvernhusbekken}.\footcite{djonno06} This owner wanted the court to help him implement a hydropower project, by compelling the other owners, $B, C$ and $D$, to rent out their share of the waterfall on terms dictated by the court. The starting point for the other owners was that they did not want any hydropower development at all, and they were not willing to rent out their rights to owner $A$ or any other developer. There was also a dispute regarding the ownership of the waterfall rights, with $A$ believing initially that he controlled a large majority. It soon became clear that this was not the case, and before the main hearing, the parties agreed that the waterfall in question was owned jointly with shares divided according to each owners' share of the joint rights still pertaining to the individual properties (rights that had been exempted from previous consolidation measures). Owner $A$'s share in these rights did not amount to more than $5 \%$, so his own financial interest in hydropower was in fact limited compared to the owners who opposed development.

On the other hand, the rights needed for the necessary physical constructions were predominantly held by owner $A$ alone, and $A$ maintained that the court should use compulsion to allow him to go on with his plans. The court agreed that hydropower would be rational use of the waterfall, and they initially assessed the case against the rules relating to compulsory joint undertakings.\footnote{See the  \cite[2 e)]{lca79}.} This promised to facilitate very concrete directives regarding how the hydropower development should be carried out, down to the level of the needed investments and building steps. 

However, the court eventually held that this approach would place too much of a burden on the owners opposing hydropower. Hence, they chose to resolve the case using directives for joint use. By doing so, the court also restricted the scope of their decision to the establishment of an owners' association that would be responsible for renting out the rights. The court would not consider the question of deciding on a concrete scheme.

The model used for the owners' association was similar to the one the court adopted in \emph{Oma}. This included allocating shares in the owners' association in a way that took into account the special importance of land needed for physical constructions. In total, this land was held to correspond to $6 \%$ of the shares in the association. Since these rights were held by owner $A$ alone, his share in the association doubled. In addition to this, owner $A$ purchased the shares from owner $B$, so that his total share ended up amounting to $22 \%$. Still, for the majority of stakeholders, membership in the association was imposed by the consolidation court against their will.

The wording of the statutes for the association apparently attempts to take into account that it would be run by a majority of unwilling shareholders. The wording used is different from that used in the other statutes, and it is stated in very clear terms that the association is going to rent out the rights in the waterfall such that hydropower can be developed. In \emph{Oma} and \emph{Vika}, on the other hand, the statutes merely state that this is the \emph{purpose} of the association, leaving the shareholders with the freedom to determine whether or not to go through with development.

In interviews, those who were compelled to take part in the association against their will expressed dissatisfaction and surprise at the result. Moreover, while the association had apparently tried to be loyal to the wording of the statutes, and had looked for partners who might be interested in developing hydropower, there had been no willingness among the majority to engage actively with this work. No deals had been made, no separate development company had been set up, and the conflict among the owners was ongoing. As of 2011, owner $A$ was still pushing for development on terms that were unacceptable to the other owners. Hence, while the case of \emph{Djønno} is an example that consolidation can be used even when it involves compulsion against the majority of owners, it also serves to illustrate that the chance of a successful outcome may then be more limited.

The question arises as to how such cases should be dealt with by courts in the future. According to owner $A$, the problem was that the directives of use were not specific enough. In his opinion, the directives should not have been restricted to merely setting up an owners' association for renting out the rights. In addition, the court should have actively engaged also with the question of how the development company should be organized, and should at least have given guidance as to \emph{who} should carry it out. Among the majority owners, on the other hand, the prevailing feeling was that the development in question, which they would be required to partake in against their will, was more or less doomed to fail already from the start.

I believe this reflects an important insight. It seems reasonable to assume that unless one is prepared to see an increase in the use of compulsion, compulsory cooperation will only work when at least a basic agreement that development should take place can be established among the majority of the involved owners.

\subsection{\emph{Tokheim}}

This case was brought before the consolidation court in 2008, by the owners of \emph{Tokheimselva}.\footcite{tokheim08} The five involved owners all agreed that development should take place, but they disagreed about how it should be done, and about the proportion of each owners' share in waterfall. Some owners argue that development should be organized by the owners themselves, but other owners thought it would be best to rent out the rights to an external developer. The case was further complicated by the fact that the waterfall in question was so big that it would be possible to develop hydro-power that would require transferral concession pursuant to the \cite{ica17}, a concession that can only be given when the purchaser is a company where the State controls at least $\frac{2}{3}$ of the shares. 

Like the precious cases we have considered the consolidation court eventually based its decision on s 2 c) of the \cite{lca79}, setting up an owners' association such that each owner was allotted a share in accordance to the rights that the court found he had in the waterfall. Unlike the previous cases we have considered, there was no adjustment made for land that would be needed for physical constructions. However, the statutes state that owners will be entitled to a lump sum estimated on the basis of the damages and disadvantages that a concrete hydropower project will bring. This also marks a departure from established practice in expropriation law, where it has been a long established principle that owners can be compensated on the basis of \emph{either} the value of their waterfalls \emph{or} the damages and disadvantages caused by the project, not both.\footnote{See for instance the case of \emph{Vikfalli}, \cite{vikfalli71}.} 

In other respects, the statutes for the owners' association follow the same model adopted in the previously considered cases. They do not, however, resolve any of the controversial questions regarding how development should be carried out, and the question of the extent to which interested owners should be given the opportunity to develop the resource themselves. This was the main issue raised by the case, but the consolidation court explicitly decided not to address it. In particular, the statutes of the owners' association explicitly provides separate rules that cover both the case that a group of owners undertake development themselves, and the case that development is carried out by an external company. 

In interviews, the owners expressed that they were happy with how the case was dealt with by the court. Everyone was heard, and the owners' association was set up in consultation with the parties. However, the main issues were still unresolved as of 2011, and this was felt by the owners as a major shortcoming of the outcome of consolidation. Some of the owners expressed criticism against the court for not engaging more actively with the most pressing issues.

The case of \emph{Tokheim} serves to illustrate that established practices of consolidation, while being well received and understood by local owners, face some new challenges in relation to hydropower, challenges that consolidation courts might be reluctant to take on. It seems that the court in \emph{Tokheim} felt that they were not in a position to assess the question of what kind of development would be best, and it also seems that they were wary of expressing any opinion about the legal status of a project led by local owners, in relation to concession law. They did not, in particular, form an opinion about whether it would be possible for local owners to carry out their own large scale development, in a waterfall that might otherwise be subject to the provisions set out in the \cite{ica17}.

It remains to be seen whether such an agnostic attitude can be maintained by the consolidation courts as local owners increasingly turn to them for help in resolving disputes regarding hydropower. Moreover, it will be interesting to see how the new \cite{lca13} will influence case law in this area. It seems that a case like \emph{Tokheim} could benefit from the court taking a broader view, possible even including public bodies as parties in the case, as will become possible when the new Act takes effect.

\section{Assessment and Future Challenges}\label{sec:lca}

The concrete cases that I discussed in the previous section shows, in my opinion, that the system of land consolidation is well suited as an alternative to expropriation in the context of hydropower development. At the same time, the cases suggest that the land consolidation courts may find it hard to deliver effective directives of use in situations when the different stakeholders disagree fundamentally about how the water resources should be managed. In addition, one may question the effectiveness of land consolidation courts in contexts when rules and regulations from other areas of law come into play. It seems, in particular, that the land consolidation courts may be cautious about implementing solutions that they fear will contradict sector-specific provisions that regulate the form of economic development that their directives aim to facilitate.

In so far as sector-specific rules disadvantage owners and benefit external commercial interests, as is the case for hydropower development, one may fear that the land consolidation courts will become impotent as soon as powerful market players enter the scene. It may be considerably easier to strike a fair balance between the interests of local farmers of comparable economic and political standing, then to do the same when one of the stakeholders is a partly state-owned power company that is accustomed to expropriating property rights.

Paradoxically, the potential weaknesses of the land consolidation courts in this regard may be enhanced by the fact that they are not authorized to make use of appropriate forms of compulsion against owners, on pain of interfering too much in property as an individual right. A lack of power to compel threatens to undermine the effectiveness of the land consolidation court, thereby making it possible to argue that the public interest in development can not be sufficiently accommodated through the use of consolidation measures. Hence, one may decide to fall back on expropriation, to the detriment of owners.

In fact, there is some evidence to suggest that land consolidation law is currently quite volnurable  to this mechanism. One indication of this is the Supreme Court case of 
{\it Holen v Holen}, concerning a quarry owned and operated by a local landowner.\footcite{holen95} In order to continue extracting his minerals, the owner of the quarry would have to interfere with the property of a neighbouring owner, who was using his land for more traditional forms of agriculture. The farmer was unwilling to reach an agreement with the quarry owner, so the latter brought a case before the land consolidation court. The court noted that it would be possible to reach an accommodation that would benefit both parties, and issued directives of use that would allow the quarry to continue its operations.

The directives gave the quarry owner access to the farmer's land, who was in turn granted replacement property from the quarry owner. The consolidation court also noted that the quarry would, in the future, be likely to extract minerals that belonged to the farmer. For the minerals as such, awarding replacement property made little sense, so the court decided that the minerals still had to be regarded as the property of the farmer (who had no interest in extracting them). However, a directive of use was issued that gave the quarry owner a right to extract these minerals, provided he paid market value for them. 

Hence, not only was the farmer awarded replacement property for agricultural purposes, he was also granted a share of the benefits that would result from the continued operation of his neighbour's quarry. This was clearly beneficial to his property, economically speaking. The owner himself, however, objected to the arrangement. The Supreme Court found in his favour. This was not because they sanctioned his right to block the continued operations of the quarry, or because they thought the replacement property or the payment model was inappropriate. Instead, the Court held that the right to extract the farmer's minerals could not be transferred to someone else by a consolidation measure, even if the farmer was ensured payment. This, the Court held, was a form of compulsion that fell outside the scope of use directives in land consolidation.

The perspective underlying this decision is interesting, because it underscores a reluctance to use land consolidation in what would otherwise be a fairly typical economic development scenario. As such, it also becomes a possible objection to the proposal that land consolidation is a practical alternative to expropriation in such scenarios. However, {\it Holen v Holen} was decided in 1995, and as I have already discussed at length, the law has developed in recent years, so that the use of land consolidation as an alternative to expropriation is now explicitly endorsed by the legislature in certain circumstances. However, I think {\it Holen v Holen} reminds us that we should take critics seriously, also because they may be able to raise convincing formal objections against compulsion in land consolidation, on the basis of earlier case law. More generally, the exact relationship between land consolidation and expropriation law, including the constitutional property clause, appears to be an increasingly relevant open question that awaits further clarification.

However, I would like to reiterate that I do not agree with those who argue that land consolidation offers less protection to owners than administrative expropriation. In my opinion, the property protection offered in the context of land consolidation has a different flavour, but is not necessarily weaker. I acknowledge, however, that how one judges this depends on one's vision of property. In particular, it depends on which one of property's functions one deems most worthy of protection.

Granted, an administrative expropriation procedure can offer more extensive {\it formal} safeguards. A range of procedural rules must be observed, pertaining to notification to the owners, impact assessments, a duty to provide guidance and reasons for the decision, and a possibility (sometimes several) for administrative appeal. Then, after an expropriation order has been granted, the owner can still challenge it before the appraisal courts, in principle at the expropriating party's expense. 

In practice, however, the administrative expropriation procedure often leaves the owners marginalized, as they are overshadowed by other more powerful stakeholders. This is particularly clear in situations when expropriation arises as a result of more comprehensive planning or licensing procedures. As discussed in Chapter 4, this as is the case, for instance, in the context of hydropower development. In addition to this, the possibility of raising validity objections before the courts is mostly a theoretical one. It is very unusual for such objections to be made successfully, as the courts typically defer to the discretion of the administrative decision-maker in expropriation cases.

More generally, the narrative of expropriation is one where the owners may have to endure a loss in the public interest, for which they must be compensated as individuals. By contrast, the narrative of consolidation is one where the owners themselves are tasked with making a {\it contribution} to the development project, in the best interest of both the local community and society in general. In particular, the owner's role is no longer than of a passive {\it obstacle} to development, but is transformed to that of an active {\it participant}, one who might have to be nudged to fulfil their potential. In addition, the properties as such receives recognition as important units of assessment, independently of the interests of their current owners. Moreover, the owners as a {\it group} come into focus, as the process is meant to facilitate rational {\it collective} action.

This is achieved by placing owners in a partly deliberative, partly adversarial, context, which presupposes their active input to the decision-making process. In addition, the {\it grounds} for imposing compulsory measures that interfere with property rights need to be anchored explicitly in the social functions of the affected properties. A measure is warranted only when it enhances property values, in the sense of improving conditions for the communities that takes their livelihoods from the affected properties. Clearly, this principle offers a kind of {\it material} protection that is largely absent in administrative expropriation law. In particular, compulsion is only warranted when it is in the interest of a concrete local community, not merely an abstract ``public''.

In some cases, such a narrative can be hard to maintain, since the local community must in fact yield to the interests of greater society. However, I believe this perspective can often be replaced by something more constructive. This is particularly clear in the context of takings for profit, since these do as a matter of fact intend to increase the (commercial) value of properties that are taken. That is not to say that any commercial development is in the interest of the local community. Far from it, the interest of the local community might be more strongly committed to preserving established use patterns or protecting the environment. But this too is something that will come under a much clearer light when the local community is thought of as an active participant in development rather than an impediment to progress.

For these reasons, I conclude that land consolidation is highly attractive, even if it must involve a significant degree of compulsion in order to effectively replace expropriation measures. In a system based on private property rights it seems only reasonable that the property owners and their local communities remain as the primary stakeholders in development, even if this development is large-scale, takes places in the public interest, and possibly needs to be imposed by compulsory means. This, in itself, does not seem to provide any moral justification for transferring decision-making power and commercial benefits away from the original group of owners, by bestowing it on powerful commercial interests. The imposition of compulsion, in particular, is only tolerable to the extent that it is actually needed to fulfil the objectives that are used to justify it. The system of land consolidation, I believe, demonstrates that current practices surrounding takings for profit fail in this regard. A better option is not only desirable, but also possible, provided there is political will to prescribe it. 

\noo{ principled objections against land consolidation in expropriation contexts appear largely misplaced when for the sub-group of takings that realise commercial potentials. However, a second question arises, of a more practical nature. Will the land consolidation process work in practice, if it is applied to organize commercial development. Increased powers of compulsion might be required, and in keeping with my argument above, I believe such powers may well be granted, as long as land consolidation remains directed at improving the situation for existing properties and their owners, rather than bestowing benefits on someone else.}

However, I would like to finish by raising a concern pertaining to the future development of the land consolidation procedure itself, it is is relied on to provide a better approach to takings for profit. If land consolidation is used to this end, will it remain a service to owners, or will it gradually be transformed into a service to developers, who seek cheap access to property owned by others? This is an empirical question concerning the robustness of specific land consolidation procedures, and it is about to come into focus in Norway. It is too early to provide answers, however, since there has not yet been any cases where the issue has come to the forefront. 

However, as any legal person with a right to expropriate may now act as a party to a consolidation dispute, the question is bound to arise. In particular, what will the role of the new parties be? Will they be embedded in the legal framework in such a way that they become potential partners that owners can rely on to implement projects in the public interest, or will they be regarded as the main stakeholders, whom the land consolidation courts should assist so that they may successfully impose their will on recalcitrant owners? It will be very interesting to follow this development further, to see if the promise of using land consolidation to regain legitimacy for economic development under compulsion can be fulfilled, or if it will lead to a degeneration of the land consolidation process itself.

\section{Conclusion}\label{sec:conc}

In this Chapter, I have addressed land consolidation as an alternative to expropriation for economic development, anchored in a case study of hydropwer. I started by presenting the basic idea of using land consolidation to organize commercial projects and I emphasised that the notion of consolidation at work here is a broad notion that includes measures seeking to compel owners to use their property in the public interest. I briefly presented an overarching vision of this kind of land consolidation, noting also that a broad notion is already at work in many jurisdictions, including in Norway. 

I then presented the Norwegian system, which is based on a judicial decision-making framework that sets the land consolidation procedure apart from that found in many other jurisdictions. I also noted how the procedure is conceptualized as a service to owners, with a no-loss guarantee in place to ensure that consolidation measures are only implemented when the benefits make up for the harms for all the involved properties.

I then went on to present the rules pertaining to compulsory use of property, that empower the consolidation courts to organize specific development projects on behalf of owners, possibly against their will. I noted how recent changes in the law envisions an extended scope for these rules, including in the context of non-agrarian and urban development. I then went on to consider some concrete examples, from the context of hydropower development, where owner-led projects already tend to rely heavily on land consolidation rather than expropriation. 

From this case study, I concluded that the land consolidation alternative works well when there is basic agreement among the owners that development is desirable, but can be less effective when there is deep disagreement about whether or not development should proceed at all, or if some owners object to the idea of owner-led development because they prefer to cooperate with an external developer. In these contexts, I argued, it might be necessary to enhance the power of the land consolidation court, also in the direction of extending its authority to compel owners to take part in, or allow the owner-led implementation of, development projects that they fundamentally disagree with. This is already possible, but only to some extent, and the exact boundaries here are now an open question, as recent legislative changes might suggest that the case law on this issue is ripe for review.

I then argued against critics that see this as a threat to property rights. In particular, I suggested that the procedural protection offered by administrative law in the context of traditional expropriation is not as relevant to the {\it actual} position of owners as the fact that they remain in focus throughout the consolidation process, as the main stakeholders in development. Hence, when land consolidation is used, external commercial interests are not able to dominate the process. 

I noted, however, that this could now be set to change, as a recent change in the law grants developers the right to act as parties in consolidation cases and to bring cases before the courts themselves, if they favour it over expropriation. On the one hand, such a change will enhance the power of the land consolidation court, making it more effective in dealing with cases that involve external parties. On the other hand, there is a possibility that the presence of new and powerful stakeholders will change the nature of the land consolidation process itself, so that it transforms from a property-enhancing institution for self-governance into a planning and implementation instrument for developers.

I general, I believe this Chapter has shown that the land consolidation regime in Norway functions in a way that sheds interesting light on collective-action alternatives to expropriation. It is also a dynamic and versatile framework, more so than other suggestions, such as the land assembly districts proposed by Heller and Hills.\footcite{heller08} In my opinion, the institution of land consolidation can provide a useful kind of democracy-on-demand for economic development takings, facilitating a better balance between the interests of the owners, the local community, and society as a whole. In this vision, external commercial actors are at worst going to be partners in crime, at best partners in progress. They will not, however, be allowed to run the show.